---
title: "t-tests"
format: revealjs
slide-number: true
editor: visual
execute:
  echo: true
html:
  code-fold: true
  code-summary: Show the code
  scrollable: true
---


## last time

- How do we get our regression estimates
- How do we evaluate our regression model


## this time

- Example of regression with 2 means aka a t-test



## t-test

- The data generating process is still assumed to be a normal

$$Y_i \sim \mathcal{N}(\mu, \sigma)$$


$$\mu_i \ = \beta_0 + \beta_1 X_i$$

----------

```{r}
#| code-fold: true
# For reproducibility
set.seed(42)

# --- 1. Create Sample Data ---
# Create data for two groups
group0 <- rnorm(50, mean = 100, sd = 15)
group1 <- rnorm(50, mean = 110, sd = 15)

# Combine into a data frame
data <- data.frame(
  y = c(group0, group1),
  group = factor(rep(c("control", "tx"), each = 50))
)

# --- 2. Run the Regression Model ---
regression_model <- lm(y ~ group, data = data)

summary(regression_model)
```



-------------

$$t = \frac{\bar{X}_1-\bar{X}_2}{{\sigma}_p * \sqrt{\frac{1}{N_1} + \frac{1}{N_2}}}$$


```{r}
#| code-fold: true
# --- 3. Run the t-test ---
# Note: var.equal = TRUE makes it equivalent to standard OLS
t_test_result <- t.test(y ~ group, data = data, var.equal = TRUE)

t_test_result
```



---------

What similarities exist between t.test and lm function? 


## pooled SD

- Important in t test is to get an assessment of the variance in each group. Often these variances can differ, so we want to get a pooled estimate, as we are assuming the groups come from the same population with the same sigma. 

- As such, pooled sd is our estimate of sigma

$$ \hat{\sigma}_p = \sqrt{\frac{(N_1-1)\hat{\sigma}^2_1 + (N_2-1)\hat{\sigma}^2_2}{N_1+N_2 - 2}} $$




## pooled SD

```{r}
#| code-fold: true
rse_from_regression <- sigma(regression_model)
cat("Residual Standard Error (from Regression):", rse_from_regression, "\n")

# --- 3. Manually Calculate the Pooled Standard Deviation ---
n1 <- length(group0)
n2 <- length(group1)
s1_sq <- var(group0) # Variance of group A
s2_sq <- var(group1) # Variance of group B

pooled_sd <- sqrt(((n1 - 1) * s1_sq + (n2 - 1) * s2_sq) / (n1 + n2 - 2))
cat("Manually Calculated Pooled SD (for t-test):", pooled_sd, "\n")
```



------------

- Standard error of the difference ($SE_{\hat{\beta}}$) indexes our precision of our estimate. Like any regression SE estimate, it uses $\hat{\sigma}$ 

$$SE_{\hat{\beta}} = \sqrt{\frac{(N_1-1)\hat{\sigma}^2_1 + (N_2-1)\hat{\sigma}^2_2}{N_1+N_2-2}} \sqrt{\frac{1}{N_1} + \frac{1}{N_2}}$$

$$SE_{\hat{\beta}} =\hat{\sigma}_p * \sqrt{\frac{1}{N_1} + \frac{1}{N_2}}$$

- Unlike the pooled SD ($\hat{\sigma}_p$), the standard error of the difference $SE_{\hat{\beta}}$ decreases the larger your sample size is. 


---------------

```{r}
#| code-fold: true
library(easystats)
model_parameters(regression_model)
```

```{r}
#| code-fold: true

# --- 3. Manually Calculate the Standard error of the difference ---
n1 <- length(group0)
n2 <- length(group1)
s1_sq <- var(group0) # Variance of group A
s2_sq <- var(group1) # Variance of group B

b_se <- sqrt(((n1 - 1) * s1_sq + (n2 - 1) * s2_sq) / (n1 + n2 - 2))*sqrt((1/n1) + ((1/n2)))
cat("Manually Calculated SE difference:", b_se, "\n")
```

```{r}
#| code-fold: true
n <- nobs(regression_model)/2 
sigma(regression_model)*sqrt(1/n + 1/n) 
```


## Group specific estimates

- regression models are prediction (y-hat) machines
- default in R is to use "dummy scores" which assign 0 or 1 to categorical variables

```{r}
library(marginaleffects)
predictions(regression_model, newdata = datagrid(group = unique))

```





## Contrasts

```{r}
class(data$group)
```

```{r}
contrasts(data$group)
```

## Deviation/ effect/sum coding

```{r}
contrasts(data$group) <- contr.sum(2)
contrasts(data$group)
```

```{r}
regression_model2 <- lm(y ~ group, data = data)
summary(regression_model2)
```

------------------

```{r}
summary(regression_model)
```




## standard error of the mean

- Indexes how much precision we have in each of our (mean) estimates

```{r}
sigma(regression_model)/sqrt(n) 
```
- Why is this equal the the SE of the $b_0$ (assuming we use R defaults)?
- Not equivalent to effect coding because those estimates are asking a different question


## R2 test

```{r}
summary(regression_model)
```

## F-test

- F-statistic is calculated as a ratio of the variance explained by your model to the unexplained variance (the residuals). Specifically, it's the ratio of the Mean Squared Regression (MSR) to the Mean Squared Error (MSE).

- MSE is sort of in the summary output (15.67^2), but MSR is not. 

## ANOVA


```{r}
anova(regression_model)
```

```{r}
15.67^2
```

$$\sum (Y_i - \bar{Y})^2 = \sum (\hat{Y_i} -\bar{Y})^2 + \sum(Y_i - \hat{Y_i})^2$$


## SS total

$$\sum (Y_i - \bar{Y})^2 = \sum (\hat{Y_i} -\bar{Y})^2 + \sum(Y_i - \hat{Y_i})^2$$

```{r}
#| code-fold: true
anova(regression_model)
```

## degrees of freedom

MSR: p (for now)

MSE: n - p - 1 


## types of t-test

- Student's t-test is used when the variances of each sample are (roughly) the same. In many cases, slightly more powerful test but the pooled variance will be biased toward the sample with the larger sample size.

- Welch's t-test if the variances are NOT (roughly) the same. Slightly less powerful. You don't pool the variances in the same way

$$\hat{\sigma}_p = \sqrt{\frac{\hat{\sigma}^2_1}{N_1}+\frac{\hat{\sigma}^2_2}{N_2}}$$

## Welch's t-test

$$Y_i \sim \mathcal{N}(\mu, \sigma)$$

$$Y_i \sim \mathcal{N}(\beta_0 + \beta_1 X_i, \gamma_0 + \gamma_1 X_i)$$

$$\mu = \beta_0 + \beta_1 X_i$$

$$\sigma = \gamma_0 + \gamma_1 X_i$$

## Effect size

Cohen's D: 

$$\delta = \frac{\mu_1 - \mu_0}{\sigma} \approx d = \frac{\bar{X}_1-\bar{X}_2}{\hat{\sigma}_p}$$

- A Cohen’s d of .2 is considered “small”, .5 is considered “moderate”, and .8 is considered “large”

------------

```{r}
#| code-fold: true
coef(regression_model)[2]/sigma(regression_model)
```


```{r}
library(emmeans)
emm <- emmeans(regression_model, "group")
eff_size(emm, sigma = sigma(regression_model), edf = df.residual(regression_model))
```


## Wrap up

$b_0$  
$b_1$  
$R^2$  
$\hat{\sigma}$  
$SE({b_1)}$ 


