---
title: "t-tests"
format: revealjs
slide-number: true
editor: visual
execute:
  echo: true
html:
  code-fold: true
  code-summary: Show the code
  scrollable: true
---


## last time

- How do we get our regression estiamtes
- How do we evaluate our regression model


## this time

- Example of regression with 2 means aka a t-test



## t-test

- The data generating process is still assumed to be a normal

$$Y_i \sim \mathcal{N}(\mu, \sigma)$$


$$Y_i \ = \beta_0 + \beta_1 X_i$$

----------

```{r}
#| code-fold: true
# For reproducibility
set.seed(42)

# --- 1. Create Sample Data ---
# Create data for two groups
group0 <- rnorm(50, mean = 100, sd = 15)
group1 <- rnorm(50, mean = 110, sd = 15)

# Combine into a data frame
data <- data.frame(
  y = c(group0, group1),
  tx = factor(rep(c(0, 1), each = 50))
)

# --- 2. Run the Regression Model ---
regression_model <- lm(y ~ tx, data = data)

summary(regression_model)
```

-------------

$$t = \frac{\bar{X}_1-\bar{X}_2}{\hat{\sigma_d}}$$


```{r}
#| code-fold: true
# --- 3. Run the t-test ---
# Note: var.equal = TRUE makes it equivalent to standard OLS
t_test_result <- t.test(y ~ tx, data = data, var.equal = TRUE)

t_test_result
```

--------------


Standard error of the difference ($SE_{\hat{\beta}}$): 

$${\hat{\sigma}_d = SE_{\hat{\beta}} = \sqrt{\frac{(N_1-1)\hat{\sigma}^2_1 + (N_2-1)\hat{\sigma}^2_2}{N_1+N_2-2}} \sqrt{\frac{1}{N_1} + \frac{1}{N_2}}}$$

$${\hat{\sigma}_d = SE_{\hat{\beta}} =\hat{\sigma}_p * \sqrt{\frac{1}{N_1} + \frac{1}{N_2}}}$$

## pooled SD

$$ \hat{\sigma}_p = \sqrt{\frac{(N_1-1)\hat{\sigma}^2_1 + (N_2-1)\hat{\sigma}^2_2}{N_1+N_2 - 2}} $$

- Unlike the pooled SD $\hat{\sigma}_p$, the standard error of the difference $\hat{\sigma}_d$ decreases the larger your sample size is. 

## pooled SD

```{r}
#| code-fold: true
rse_from_regression <- sigma(regression_model)
cat("Residual Standard Error (from Regression):", rse_from_regression, "\n")

# --- 3. Manually Calculate the Pooled Standard Deviation ---
n1 <- length(group0)
n2 <- length(group1)
s1_sq <- var(group0) # Variance of group A
s2_sq <- var(group1) # Variance of group B

pooled_sd <- sqrt(((n1 - 1) * s1_sq + (n2 - 1) * s2_sq) / (n1 + n2 - 2))
cat("Manually Calculated Pooled SD (for t-test):", pooled_sd, "\n")
```

---------------

```{r}
#| code-fold: true
library(easystats)
model_parameters(regression_model)
```

$${\hat{\sigma}_d = SE_{\hat{\beta}} =\hat{\sigma}_p * \sqrt{\frac{1}{N_1} + \frac{1}{N_2}}}$$

```{r}
#| code-fold: true
n <- nobs(regression_model)/2 
sigma(regression_model)*sqrt(1/n + 1/n) 
```


## Group specific estimates

```{r}
library(marginaleffects)
predictions(regression_model, newdata = datagrid(tx = unique))

```

```{r}
avg_predictions(regression_model, by = "tx")
```


## standard error of the mean


```{r}
sigma(regression_model)/sqrt(n) 
```

## Plotting







## types of t-test

- Student's t-test is used when the variances of each sample are (roughly) the same. In many cases, slightly more powerful test but the pooled variance will be biased toward the sample with the larger sample size.

- Welch's t-test if the variances are NOT (roughly) the same. Slightly less powerful. You don't pool the variances in the same way

$$\hat{\sigma}_p = \sqrt{\frac{\hat{\sigma}^2_1}{N_1}+\frac{\hat{\sigma}^2_2}{N_2}}$$

## Welch's t-test

$$Y_i \sim \mathcal{N}(\mu, \sigma)$$

$$Y_i \sim \mathcal{N}(\beta_0 + \beta_1 X_i, \gamma_0 + \gamma_1 X_i)$$


## Effect size

Cohen's D: 

$$\delta = \frac{\mu_1 - \mu_0}{\sigma} \approx d = \frac{\bar{X}_1-\bar{X}_2}{\hat{\sigma}_p}$$

- A Cohen’s d of .2 is considered “small”, .5 is considered “moderate”, and .8 is considered “large”

------------

```{r}
#| code-fold: true
coef(regression_model)[2]/sigma(regression_model)
```


```{r}
library(emmeans)
emm <- emmeans(regression_model, "tx")
eff_size(emm, sigma = sigma(regression_model), edf = df.residual(regression_model))
```


# One sample t-test


