---
title: "ANOVA"
format: revealjs
slide-number: true
editor: visual
execute:
  echo: true
html:
  code-fold: true
  code-summary: Show the code
  scrollable: true
---


## This time

- Bring it together; glm as a functional tool
 
- comparing two or more means ie ANOVA

## two means (t-test)

$$Y_i \sim \mathcal{N}(\mu, \sigma)$$

$$\mu_i \ = \beta_0 + \beta_1 X_i$$

$$Y_i = \beta_0 + \beta_1 X_i + e_{i}$$

## two means

$$t = \frac{\bar{X}_1-\bar{X}_2}{{\sigma}_p * \sqrt{\frac{1}{N_1} + \frac{1}{N_2}}}$$
- difference / se(diff) aka se(regression).   
- difference / pooled SD * ~N aka sigma * ~N.  
- standard error of group mean = sigma / sqrt(N). 



## dummy/effect coding

- Given nominal variables we have to assign numbers to each group
- We do so based on the hypothesis we want to test
- Regardless of how we code it does not impact the overall model fit


## more than 2 means

- Same idea extends to when we have more than two means
- This is often called "anova" (analysis of varince) but referring to anova vs t-test vs regression makes it appear these are all different rather than being the same


## more than 2 means

- Why is it called analysis of "VARIANCE" when we are interested in more than 2 means?

$$\sum (Y_i - \bar{Y})^2 = \sum (\hat{Y_i} -\bar{Y})^2 + \sum(Y_i - \hat{Y_i})^2$$

$$\sum (Y_i - \bar{Y})^2 = \sum (\bar{\text{Group}_k} -\text{Grand mean})^2 + \sum(Y_{ik} - \bar{\text{Group}_k})^2$$

- And we will often use ANOVA tables (which looks at this variance decomposition) rather than regression coefficients. 

## Example

```{r}
#| code-fold: true

library(ggplot2)
library(dplyr)
library(tibble) 

# --- 1. Set Simulation Parameters ---

set.seed(74) # Makes your simulation reproducible

n_per_group <- 50  # 50 participants per group
residual_sd <- 1.0 # This is the within-group SD

# Define the true means for our three groups.
# We'll set them apart to ensure the ANOVA finds a significant effect.

mean_group_A <- 4.0
mean_group_B <- 5.0
mean_group_C <- 6.0

# --- 2. Generate the Data ---

# Create the data for each group from a normal distribution

group_A_data <- rnorm(n_per_group, mean = mean_group_A, sd = residual_sd)
group_B_data <- rnorm(n_per_group, mean = mean_group_B, sd = residual_sd)
group_C_data <- rnorm(n_per_group, mean = mean_group_C, sd = residual_sd)

# Combine into a single "long" data frame, which is required for ANOVA
simulated_data <- data.frame(
  dv = c(group_A_data, group_B_data, group_C_data),
  group = factor(rep(c("Group A", "Group B", "Group C"), each = n_per_group))
)

```


```{r}
#| code-fold: true
simulated_data
```

---------------


```{r}
#| code-fold: true
ggplot(simulated_data, aes(x = group, y = dv, fill = group)) +
  geom_violin(trim = FALSE, alpha = 0.6) +
  geom_jitter(width = 0.15, alpha = 0.4, color = "black") +
  geom_boxplot(width=0.1) +
  labs(
    y = "Happiness",
    x = "Group"
  ) +
  theme_minimal(base_size = 14) +
  scale_fill_brewer(palette = "Set2") +
  guides(fill = "none") 
```


--------------

```{r}
#| code-fold: true
library(ggdist)
ggplot(simulated_data, aes(x = group, y = dv, fill = group)) +
stat_eye(alpha = .6)+ 
  geom_jitter(width = 0.15, alpha = 0.4, color = "black") +
  labs(
    y = "Happiness",
    x = "Group"
  ) +
  theme_minimal(base_size = 14) +
  scale_fill_brewer(palette = "Set2") +
  guides(fill = "none") 
```


--------------

```{r}
mod.1 <- lm(dv ~ group, simulated_data)
summary(mod.1)
```

-----------------------


$$Y_i = \beta_0 + \beta_1 D1_i + \beta_2 D2_i + e_{i}$$



## Why 2 for 3 groups? 

- If you have 3 means, how can you represent that in regression model? 

- k(number of levels/groups) - 1

- If you have an intercept, you have an additional mean so technically you are modeling 4 means if you try to model 3 coefficients, one for each group, plus an intercept.


## Reference group

- What ever is coded zero for both coeffiecents, will be the intercept


```{r}
contrasts(simulated_data$group)
```

| Group  | $b_1$ | $b_2$ | 
|---------|:-----|------:|
| Group A | 0   |    0 |  
| Group B   | 1  |   0 |  
| Group C   | 0  |     1 |  



## If we omit an intercept


```{r}
mod.2 <- lm(dv ~ 0 + group, simulated_data)
summary(mod.2)
```

----------

- Creates correct coefficients, but it is harder to evaluate. 

- R2 and F is off. Hard to learn from this model. 

- This coding is done with some machine learning models that do not have intercepts (i.e. one hot encoding)


## coding contrasts are up to you

- deviation/efffect/sum coding
```{r}
contr.sum(3)
contrasts(simulated_data$group) <- contr.sum(3)
contrasts(simulated_data$group)
```


----------------

```{r}
mod.3 <- lm(dv ~ 1 + group, simulated_data)
summary(mod.3)
```


## deviation/efffect/sum coding

- Uses -1, 1, and zeros
- Coefs compare to a grand mean
- Often used for ANOVA models
- Intercept is now the grand mean rather than reference group mean

Mean(A) = β₀ + β₁

Mean(B) = β₀ + β₂

Mean(C) = β₀ - β₁ - β₂ 



## ANOVA Dummy coding 

```{r}
anova(mod.1)
```


$$\sum (Y_i - \bar{Y})^2 = \sum (\bar{\text{Group}_k} -\text{Grand mean})^2 + \sum(Y_{ik} - \bar{\text{Group}_k})^2$$

## degrees of freedom

MS Regression: p 

MS Error: n - p - 1


## What does the F test? 

$$H_0: \mu_1 = \mu_2 = \mu_3 $$
$$H_1: \text{One group (or more) differs from another} $$

## Why not run many t-tests?

- parsimony
- pooling
- p-value inflation (increase family wise error rate)


## family wise error rate

- The probability that one or more of your "family" of multiple tests is false

P(making at least 1 error in m tests) = $1 - (1-\alpha)^m$

$1-(1-.05)^6$ = .265

- ANOVA omnibus test "helps" control this

--------------------

- Two step approach to FWER: omnibus + post hoc comparisons

- It only controls the family-wise error rate in a weak sense. 
- It guarantees that you have only a 5% chance of making any error if the null hypothesis is 100% true (i.e., A=B=C). 
- It does not control the error rate for the specific post hoc pairwise comparisons
- If you are interested in only some subset of possible comparisons eg Ho: A = C, then the omnibus is not useful. 
- Thus many say omnibus test is mostly useless, and one should just focus on post-hoc comparisons. 


## ANOVA effect coding

```{r}
anova(mod.3)
```

## Compare with ANOVA dummy coding

```{r}
anova(mod.1)
```

## More terminology

- In ANOVA speak, people often use MS between and MS within 

- These are exactly the same as MS model / MS regression and MS error / MS residual. 

## Can you visualize ANOVA? 

- What does MSregression represent? 

- What does MSerror represent? 



## Please don't

```{r}
aov.1 <- aov(dv ~ group, simulated_data)
summary(aov.1)
```



## Predictions revisited

- As with all models, we want to learn from what we built.

- Often, this means we are going to create predictions from the model.

$$\hat{Y_i} = \beta_0 + \beta_1 D1_i + \beta_2 D2_i $$

------------------

```{r}
library(broom)
m1_aug<- augment(mod.1)
m1_aug
```

## EM means

- Estimated marginal means i.e., predicted values

- Useful for complex models where we want to average over (marginalize) some other variables e.g., what is the predicted value for a 74 year old from Missouri who identifies as female?

- Used for figures, interpreting models, and post hoc comparisons


## post-hoc tests

- Does our reference group (in dummy coding) differ from other groups? 

- Does our "left out" group (in effect/sum coding) differ from the grand mean? 


-----------

```{r}
library(emmeans)
m1.emm<- emmeans(mod.1, ~group)
m1.emm
```

## Same predicted means for effect coding

```{r}
emmeans(mod.3, ~group)
```



---------

```{r}
pairs(m1.emm)
```

## Give me CIs

```{r}
pairs(m1.emm) |> 
     confint()
```

## What is Tukey's method? 

- Used for pairwise comparisons

$$q = \frac{\text{Mean 1 - Mean 2}}{\sqrt{\frac{\text{MS Within}}{\text{N per group}}}} \\
\text{Mean 1 - Mean 2} = q\sqrt{\frac{\text{MS Within}}{\text{N per group}}}$$

- Asks: What is the minimal difference between two means needed to declare something significantly different?


## False Discovery Rate (FDR)

```{r}
pairs(m1.emm, adjust = "fdr") 
```

## adjusting the alpha or adjusting the p-value

- Old school way is to adjust an alpha eg i now need to have my p-value below .023 to be considered significant. P-value was the same. 

- New way is to adjust the p-value (sometimes called q-values). Easier to do this. Technically the same.  

## Rant

- I hate this new way, even though technically the same. 

- It is not interpreted as a probability from a standard null, and thus it is one more source of confusion of what a p-value is. 

- FDR comes from multiple distributions (as it depends on the other p-values observed), Tukey comes from a different null (Studentized Range Distribution (q-distribution)). 

- If I declare every test with a q-value of 0.05 or less to be 'significant,' I can expect that about 5% of the tests on this list of 'discoveries' are actually false positives. But any individual FDR q-value is difficult to impossible to interpret

## p-value correction impacts CIs


```{r}
pairs(m1.emm, adjust = "fdr") |> confint()
```

## p-value correction impacts CIs

```{r}
pairs(m1.emm) |> confint()
```


```{r}
pairs(m1.emm, adjust = "none") |> confint()
```

-----------

Cohen's d. 

```{r}
eff_size(m1.emm, sigma = sigma(mod.1), edf = 147)
```


## Graphing

```{r}
plot(m1.emm)
```

--------


```{r}
#| code-fold: true
library(ggplot2)
dat.em <- as.data.frame(m1.emm) 
simulated_data |>   
ggplot(aes(x = group, y = dv)) + 
  geom_violin()+
  geom_point(position = position_jitter())+
  geom_errorbar(data = dat.em, aes(x = group, y = emmean,ymin = lower.CL, ymax = upper.CL), width = 0.2)
  
  
```



