---
title: "Assumptions"
format: revealjs
slide-number: true
editor: visual
execute:
  echo: true
html:
  code-fold: true
  code-summary: Show the code
  scrollable: true
---


## Assumptions of regression

- No measurement error

- Correctly specified form

- No omitted variables

- Homoscedasticity

- Independence among the residuals

- Normally distributed residuals

## What happens


| Violated Regression Assumption | Coefficients | Standard Errors |   
|--------------------------------|--------------|-----------------|
| 1. Measured without error      | Biased       | Biased             |
| 2. Correctly specified form    | Biased       | Biased             |  
| 3. Correctly specified model   | Biased       | Biased              |
| 4. Homoscedasticity            | Biased       |                    |
| 5. Independent Errors          | Biased       |                    |
| 6. Normality of the Errors     | Biased       |                   |   


## Detection

| Assumption                   | Detection                   | 
|------------------------------|-----------------------------|
| 1. Measured without error    | Reliability                 |  
| 2. Correctly specified form  | Residuals against predicted |   
| 3. Correctly specified model | Theory & Prediction plot    |
| 4. Homoscedasticity          | Residuals against predicted  |
| 5. Independent Errors        | Research Design               |
| 6. Normality of the Errors   | q-q plot or distribution      |

## Reliability

- If there is measurement error, our coefficient estimates will always UNDER-estimate the true parameter. This is because stochastic randomness (error) cannot relate with anything. 

$$r_{xy} = \rho\sqrt{r_{xx}r_{yy}}$$
- Measurement error inflates our standard errors, because they add...error


## Correctly specified form

- Categorical when categorical. Linear when linear. 


-----------------

```{r}
#| code-fold: true
library(performance)
library(readr)
anxiety <- read_csv("anxiety.csv")
m1 <- lm(Anxiety ~ Support, anxiety)
check_model(m1)
```

--------------

```{r}
#| code-fold: true
# return a list of single plots
diagnostic_plots <- plot(check_model(m1, panel = FALSE))
# linearity
diagnostic_plots[[2]]
```





## What about categorical predictor?

```{r}
#| code-fold: true
m2 <- lm(Support ~ as.factor(group), anxiety)
d_plots <- plot(check_model(m2,panel = FALSE))
# linearity
d_plots[[2]]
```



## Correctly specified model

- Correct Likelihood function. 

- Not overfitting or underfitting


--------------

```{r}
#| code-fold: true
diagnostic_plots[[1]]
```


---------------

```{r}
#| code-fold: true
anxiety$skewed <- rpois(118, 1)
m3 <- lm(skewed ~ Support, anxiety)
dia_plots <- plot(check_model(m3, panel = FALSE))
dia_plots[[1]]
```


------------

```{r}
# categorical
d_plots[[1]]
```




## Homoscedasticity


```{r}
#| code-fold: true
library(ggplot2)
d <- data.frame(
  x = fitted(m1),
  y = residuals(m1),
  grp = as.factor(residuals(m1) >= 0)
)
ggplot(d, aes(x, y)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) + xlab("fitted") +ylab("redisuals")
```

-----------------------


```{r}
#| code-fold: true
library(ggplot2)
d <- data.frame(
  x = fitted(m1),
  y = residuals(m1),
  grp = as.factor(residuals(m1) >= 0)
)
ggplot(d, aes(x, y, colour = grp)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE)+ xlab("fitted") +ylab("redisuals")
```

---------

```{r}
#| code-fold: true
# homoscedasticiy - homogeneity of variance
diagnostic_plots[[3]]
```


## How to fix homoscedasticity

- Calculating heteroscedasticity-consistent standard errors accounts for the larger variation, better reflecting the increased uncertainty. 

- The heteroscedasticity can be modeled directly (MELSMs)

- Transforming the response variable(eg log)

- Different estimation (eg WLS) which weights observations differently


## Independent Errors  

- Based on design and sampling
- Can tell by ICC if you know grouping and or residual based on time if time is a factor. 




## Normally distributed residuals/errors

```{r}
ggplot(m1) +
  stat_qq(aes(sample = .stdresid)) +
  geom_abline() +
  labs(title = "Q-Q Plot",
       subtitle = "for Normality of Errors") +
  theme_bw(base_size = 20)
```

---------

```{r}
diagnostic_plots[[5]]
```



## Note about normality

- None of the assumptions are about normally distributed variables
- But haven't I assumed at normal DGP? 
- Yes, but the assumption is that it is conditionally normal. 

So all you have to do is justify Gaussian as good default over another theoretically probablity distribution. And, check to see if residuals are normally distributed


---------------

| Assumption                   | Fix                                          |
|------------------------------|----------------------------------------------|
| 1. Measured without error    | SEM, factor scores, more data, better design |
| 2. Correctly specified form  | Different model terms                              |
| 3. Correctly specified model | Theory; new likelihood; Specificity tests                         |
| 4. Homoscedasticity          | Robust SEs, WLS, transformations, model it             |
| 5. Independent Errors        | Use different analysis method (MLM)                |
| 6. Normality of the Errors   | Additional IVs, different form               |


## When data go bad

- Are "Outliers" expected? 

- How do you know something is an outlier

- Do you do anything about it? 

## Outliers

- Each conveys a sense of the magnitude of outliery-ness the case exhibits. 

1. Leverage -- How unusual is this case from the rest of the cases in terms of predictors?
2. Distance -- How distant is the observed case from the predicted value?
3. Influence -- How much the does regression coefficient change if case were removed?

## Leverage 

- Leverage tells us how far observed values for a case are from mean values on the set of IVs (centroid).

- Not dependent on Y values

- Often arbitrarily at 3 sd


## Distance 

- Distance is the distance from prediction, or how far a case's observed value is from its predicted value i.e., residual

- In units of Y.

