---
title: "factorial anova"
format: revealjs
slide-number: true
editor: visual
execute:
  echo: true
html:
  code-fold: true
  code-summary: Show the code
  scrollable: true
---


## This time

- More than one IV! 
- Two+ categorical predictors

## nomenclature

- Between subjects design 
- Two-way
- 3 Dose levels X 2 Treatment levels

|        | Control  | Treatment | 
|---------|-----|------|-------| 
| Dose A |  $\hat{Y_{AC}}$   | $\hat{Y_{ATx}}$         |
| Dose B |  $\hat{Y_{BC}}$   | $\hat{Y_{BTx}}$         | 
| Dose C |  $\hat{Y_{CC}}$   |  $\hat{Y_{CTx}}$        | 


## Cell vs Marginal Means

|        | Control  | Treatment | Marginal Mean|
|---------|-----|------|-------| 
| Dose A |  $\hat{Y_{AC}}$   | $\hat{Y_{ATx}}$         |  A MM | 
| Dose B |  $\hat{Y_{BC}}$   | $\hat{Y_{BTx}}$         | B MM | 
| Dose C |  $\hat{Y_{CC}}$   |  $\hat{Y_{CTx}}$        | C MM | 
| Marginal Mean |Control MM | Tx MM|
  
## Why two + variables? 

- Efficiency
- New questions like dependencies (ie interactions)


## Variance decomposition one way

$$\sum (Y_i - \bar{Y})^2 = \sum (\hat{Y_i} -\bar{Y})^2 + \sum(Y_i - \hat{Y_i})^2$$


$$\sum (Y_i - \bar{Y})^2 = \sum (\bar{\text{Group}_k} -\text{Grand mean})^2$$

$$+ \sum(Y_{ik} - \bar{\text{Group}_k})^2$$

## Variance decomposition two way?

- What does the model look like for a two way anova? 

------------------

```{r}
#| code-fold: true
# Load necessary libraries
library(ggplot2)
library(dplyr)
library(tibble) # For a cleaner summary print

# --- 1. Set Simulation Parameters ---

set.seed(123) # Makes your simulation reproducible

# Design: 3 (Dose) x 2 (Condition)
n_per_cell <- 50 # 300 total participants / 6 cells = 50 per cell

# Set the standard deviation for the "noise" or error.
# A smaller SD makes the effects easier to detect (i.e., "significant").
residual_sd <- 2.0

# --- 2. Define the "True" Population Means ---
# We will set the 6 cell means to create a crossover interaction AND main effects.
#
# Interaction: The effect of Condition will be different at each level of Dose.
#   - For Dose A, the mean goes UP (+4)
#   - For Dose B, the mean goes UP (+2)
#   - For Dose C, the mean is FLAT (0)
#   This non-parallelism creates the interaction.
#
# We also set them so the marginal means are different.
mean_A_Control <- 6
mean_A_TX <- 10

mean_B_Control <- 9
mean_B_TX <- 11

mean_C_Control <- 12
mean_C_TX <- 12

# --- 3. Generate the Data ---

# Create the data for each of the 6 cells
data_A_Control <- rnorm(n_per_cell, mean = mean_A_Control, sd = residual_sd)
data_A_TX <- rnorm(n_per_cell, mean = mean_A_TX, sd = residual_sd)
data_B_Control <- rnorm(n_per_cell, mean = mean_B_Control, sd = residual_sd)
data_B_TX <- rnorm(n_per_cell, mean = mean_B_TX, sd = residual_sd)
data_C_Control <- rnorm(n_per_cell, mean = mean_C_Control, sd = residual_sd)
data_C_TX <- rnorm(n_per_cell, mean = mean_C_TX, sd = residual_sd)

# Combine into a single "long" data frame
simulated_data <- data.frame(
  dv = c(data_A_Control, data_A_TX, data_B_Control, data_B_TX, data_C_Control, data_C_TX),
  Dose = factor(rep(c("A", "A", "B", "B", "C", "C"), each = n_per_cell)),
  Condition = factor(rep(c("Control", "TX", "Control", "TX", "Control", "TX"), each = n_per_cell))
)

simulated_data

```

-----------------

```{r}
 simulated_data |> 
  group_by(Dose, Condition) |> 
  summarise(Mean = mean(dv), .groups = 'drop')

```

---------

```{r}
simulated_data %>%
  group_by(Dose) %>%
  summarise(Marginal_Mean = mean(dv), .groups = 'drop')
```



```{r}
simulated_data %>%
  group_by(Condition) %>%
  summarise(Marginal_Mean = mean(dv), .groups = 'drop')
```

----------

```{r}
#| code-fold: true
 simulated_data |> 
  group_by(Dose, Condition) |> 
  summarise(Mean = mean(dv), .groups = 'drop') |> 
ggplot(aes(x = Condition, y = Mean, color = Dose, group = Dose)) +
  geom_line(size = 1.2) +
  geom_point(size = 3) +
  labs(
    title = "Cell means",
    y = "DV",
    x = "Condition",
    color = "Dose"
  ) +
  theme_minimal(base_size = 14)

```


------------------

$$Y_i = \beta_0 + \beta_1 \text{Dose1}_i + \beta_2 \text{Dose2}_i +  \beta_3 \text{Condition}_{i}$$


-------------

```{r}
fit.1 <- lm(dv ~ Dose + Condition, simulated_data)
summary(fit.1)
```

------------

```{r}
anova(fit.1)
```

## Sums of squares decomposition mk2


$$\sum (Y_i - \bar{Y})^2 = \sum (\bar{\text{Dose}_k} -\text{Dose G.M.})^2 +$$
$$\sum (\bar{\text{Condition}_j} -\text{Condition G.M.})^2 + \sum(Y_{ikj} -  \hat{Y_{jk}})^2$$

Where $\hat{Y_{jk}}$ = Grand mean + Marginal Effect of Dose + Marginal effect of Condition. (technically cell means; but without an interaction we only focus on MMs) 

------------

```{r}
#| code-fold: true
grand_mean <- mean(simulated_data$dv)

dose_means <- simulated_data |> 
  group_by(Dose) |> 
  summarise(ME_dose = mean(dv))

condition_means <- simulated_data |> 
  group_by(Condition) |> 
  summarise(ME_condition = mean(dv))

simulated_data |> 
  left_join(dose_means, by = "Dose") |> 
  left_join(condition_means, by = "Condition") |> 
  mutate(GM = grand_mean) |> 
  mutate(
    Effect_Dose = ME_dose - GM,
    Effect_Condition = ME_condition - GM) |> 
  mutate(Predicted_Value = GM + Effect_Dose + Effect_Condition)  |> 
  mutate(Residual = dv - Predicted_Value) |> 
  mutate(Squared_Residual = Residual^2) |> 
 summarise(
  SS_Dose_Manual = sum(Effect_Dose^2), # SS Dose
  SS_Condition_Manual = sum(Effect_Condition^2), # SS Dose
  SS_Within_Manual = sum(Squared_Residual)) # SS within
```

## Differences from one way anova 

- What would SS dose be if we ran a one way without condition? 

- What would SS residuals look like in this model? 

- What would SS residual look like, relative to two factor model, if dose was a horrible predictor? 

-------------

```{r}
fit.dose <- lm(dv~Dose, simulated_data)
anova(fit.dose)
```

## Model comparison

- What if I wanted to test a dose only model vs a two-way model with condition? 
- The model comparison between SS error of the two models is equal to: 




---------

```{r}
anova(fit.dose, fit.1)
```

--------

- The model comparison between SS error is equal to: 
  1. The difference between SS error Full (Dose and condition) vs reduced (Dose only)
  2. The SS condition
  3. F for SS condition
  

  
----------  
  
  
## Model Effect size

- Coef of determination ($R^2$) is percent variance explains. 
- Eta-squared ($\eta^2$) is the same, but historically used for ANOVA models

```{r}
library(effectsize)
eta_squared(fit.dose)
```


---------

```{r}
summary(fit.dose)
```

-------------

$\eta^2$ = $R^2$ = SSRegression/SStotal

```{r}
836.32/(836.32 + 1660.89)
```


```{r}
anova(fit.dose)
```


-----------

- Effect size for model works when we have a single factor with multiple levels. But what if I wanted the effect size for condition?  


```{r}
anova(fit.1)
```


## Partial Eta squared


```{r}
eta_squared(fit.1)
```


------------------
 
$$R^2 \neq \eta_{dose}^2 + \eta_{condition}^2 $$

```{r}
summary(fit.1)
```

-------------------

```{r}
836.32 / (836.32 + 1213.71) # Dose
```

```{r}
447.17 / (447.17 + 1213.71) # Condistion
```

```{r}
anova(fit.1)
```


--------------

- In the two factor model $\eta_{dose}^2$ = .41 whereas in the one-factor model the effect of dose was .33. 

- Why? 


## partial eta squared interpretation

- Percent variance explained after accounting for other variables in the model. 
- If you put a good predictor in the model that explains the DV well your other variables are going to increase their $\eta^2$

- Thus $\eta^2$ is not one number and it depends on what else is in the model. If you do not know what else is in the model and how strongly it relates to the DV, then $\eta^2$ has little utility.  


## Omega squared

- A less biased estimate of proportion of variance explained. 

$$\omega^2 = \frac{df_{effect} \times (MS_{effect} - MS_{error})}{SS_{total} +  MS_{error} }$$

```{r}
omega_squared(fit.1)
```

## post hoc tests

marginal means 

```{r}
library(emmeans)
emm.1<- emmeans(fit.1, ~ Dose)
emm.1
```

  
-----------

```{r}
library(emmeans)
emm.2<- emmeans(fit.1, ~Condition)
emm.2
```
  
------- 

```{r}
pairs(emm.2)
```  
  
```{r}
pairs(emm.1) |> confint()
```   
  
---------

```{r}
summary(fit.1)
```

---------

- predicted mean vs actual mean for intercept

```{r}
 simulated_data |> 
  group_by(Dose, Condition) |> 
  summarise(Mean = mean(dv), .groups = 'drop') 
```

------ 

- Coefs are marginal means, not cell means (if balanced sample size)

```{r}
8.492199	- 6.068807	# dose B difference from A in control
```
- 1.6041 is regression est.

```{r}
pairs(emm.1) |> confint()
```


---------

- When we have more than two predictors the coefficients cannot be computed directly from the 6 raw cell means (and vice versa).

- The model we fit assumes that all we need are the marginal means. Or that the cell mean difference, for example, between Tx and Con is the same across all levels of dose. 

- How do we weaken this assumption?

## Interactions

- Thus far our interpretations of the model have been at the marginal means. 

- If we think the difference between control and tx DEPENDS on the dose, then we have an interaction. 

|        | Control  | Treatment | Marginal Mean|
|---------|-----|------|-------| 
| Dose A |  $\hat{Y_{AC}}$   | $\hat{Y_{ATx}}$         |  A MM | 
| Dose B |  $\hat{Y_{BC}}$   | $\hat{Y_{BTx}}$         | B MM | 
| Dose C |  $\hat{Y_{CC}}$   |  $\hat{Y_{CTx}}$        | C MM | 
| Marginal Mean |Control MM | Tx MM|


## Marginal vs Cell means

- When no interactions are present, the focus is on marginal means

- When interactions are modeled, the focus is on cell means (or how the main effects differ across levels of the other variable; simple main effects)



## model formula

You take each term and multiply it by the other factor. 

$$Y_i = \beta_0 + \beta_1 \text{Dose1}_i + \beta_2 \text{Dose2}_i +  \beta_3 \text{Condition}_{i}$$

$$Y_i = \beta_0 + \beta_1 \text{Dose1}_i + \beta_2 \text{Dose2}_i +  \beta_3 \text{Condition}_{i}$$ 
$$ + \beta_4 \text{(Dose1*Condition)}_i + \beta_5 \text{(Dose2*Condition)}_i$$

--------------

|ID | Dose1 | Dose2 | Control | D1*C | D2*C| Cell | 
|---|-------|------ |-------| ------|-------|------| 
| 1 |  0    | 0     |  0    |   0   |  0    | ACon|
| 2 |  1    | 0     | 0     |   0   |  0    |BCon |
| 3 |  0    | 1     | 0     |   0   |  0    |CCon |  
| 4 |  0    | 1     | 1     |   0   |  1    |Ctx |
| 5 |  0    | 0     | 1     |   0   |  0     |Atx |
| 6 |  1    | 0     | 1     |   1   |  0     |Btx |



## recreates our six cell means

|        | Control | Treatment | Marginal Mean|
|---------|-----|------|-------| 
| Dose A |  $\hat{Y_{AC}}$   | $\hat{Y_{ATx}}$         |  A MM | 
| Dose B |  $\hat{Y_{BC}}$   | $\hat{Y_{BTx}}$         | B MM | 
| Dose C |  $\hat{Y_{CC}}$   |  $\hat{Y_{CTx}}$        | C MM | 
| Marginal Mean |Control MM | Tx MM|



-----------

```{r}
fit.2 <- lm(dv ~ 1 + Dose * Condition, simulated_data)
summary(fit.2)
```


## How to interpret


$\beta_0$ = Cell mean for Dose A and Control. 

$\beta_1\text{Dose1}_i$ = Simple main effect of Dose B. It's the difference between Dose B and Dose A when in control. 

$\beta_2\text{Dose2}_i$ = Simple main effect of Dose C. It's the difference between Dose C and Dose A when in control. 

$\beta_3 \text{Condition}_{i}$ = Simple main effect of Condition. It's the difference between Tx and Control when dose = 0.   

## Simple main effects

- Not main effects (marginal means). They compare the effect of one variable at a specific level of the moderator variable.

- That specific level is always when the moderator is at zero. Why zero? Because that is how the math works. 

- In the full model, if I want to ask what is the effect of condition, it is hard to say because condition is in $\beta_3$, $\beta_4$, and $\beta_5$ terms. But if I look at condition = 0, then  $\beta_4$ and $\beta_5$ terms drop out. 

- With interactions, lower order terms are always "conditional" -- they depend on other terms in the model. 

----------------

$\beta_4 \text{(Dose1*Condition)}_i$ = is the simple interaction effect (a contrast/comparison of an already existing contrast). A difference of a difference. It's the difference between Dose B and Dose A when in control $\beta_2$ compared to the  difference between Dose B and Dose A when in Treatment. 

$\beta_5 \text{(Dose2*Condition)}_i$ = is the simple interaction effecet interpreted as the difference between Dose C and Dose A when in control $\beta_3$ compared to the difference between Dose C and Dose A when in Treatment. 

## simple interaction effect

- If there is no difference between simple main effects at different levels of the moderator (ie a simple interaction effect) then there is no interaction. I.e., the effect for one variable does not DEPEND on the moderator variable. Can thus simplify to marginal means

- If there is an interaction it suggests that the interpretation of simple main effects DEPEND on the levels of the moderator variable. And thus you have to look at cell means to interpret. 



## Predicted scores i.e., cell means

| Dose1 | Dose2 | Control | D1*C | D2*C| Cell | 
|-------|------ |-------| ------|-------|------| 
 |  0    | 0     |  0    |   0   |  0    | ACon|
|  1    | 0     | 0     |   0   |  0    |BCon |
|  0    | 1     | 0     |   0   |  0    |CCon |  
|  0    | 1     | 1     |   0   |  1    |Ctx |
|  0    | 0     | 1     |   0   |  0     |Atx |
 |  1    | 0     | 1     |   1   |  0     |Btx |
 
 $$Y_i = \beta_0 + \beta_1 \text{Dose1}_i + \beta_2 \text{Dose2}_i +  \beta_3 \text{Condition}_{i}$$ 
$$ + \beta_4 \text{(Dose1*Condition)}_i + \beta_5 \text{(Dose2*Condition)}_i$$

## ANOVA

- Where did SS Dose:Condition come from? 
```{r}
anova(fit.2)
```



## degrees of freedom

Effect: #levels - 1 (or number of coefficients)

Interaction: df effect 1 * df effect 2

Error: N - df effects - df interactions - 1


## Sums of squares decomp interaction

$$\sum (Y_i - \bar{Y})^2 = \sum (\bar{\text{Dose}_k} -\text{Dose G.M.})^2 +$$

$$\sum (\bar{\text{Condition}_j} -\text{Condition G.M.})^2 +$$
Variation of the cell means that cannot be explained by M.M.

$$\sum (\hat{Y_{jk}} -  \text{G.M.})^2 - SS_\text{Dose} - SS_\text{condition}$$

$$ + \sum(Y_{ikj} -  \hat{Y_{jk}})^2$$

## SS interaction

- Can be thought of as whether I can predict the cell means using just the marginal means and ask if simple main effects are the same. 

|                           | Condition |    |                      |
|----------------|----------|-------------:|------------------|
|Dose            | Control  | TX |MM (Dose) |
| A                         | 10 (17.5) | 20 (22.5) | 15         |
| B                         | 20 (22.5)   | 30 (27.5) | 25       |
| C                         | 30 (27.5)  | 40 (32.5)| 35           |
| MM (Condition) | 20       | 30 | Grand Mean = 25      |


----------

```{r}
anova(fit.1, fit.2)
```

## Simple main effects in emmeans

- Be careful. With an interaction you are interested in cell means. Do not interpret marginal means

```{r}
qq <- emmeans(fit.2, ~  Dose )
qq 
```

---------------

```{r}
pairs(qq)
```


------------------

```{r}

qq <- emmeans(fit.2, ~  Dose | Condition)
pairs(qq)

```

-----------

```{r}
pairs(qq, simple = "Condition") 
```

--------------

```{r}
summary(fit.2)
```

## interaction contrasts

```{r}
qqq <- emmeans(fit.2, ~ Dose * Condition)
CON <- pairs(qqq, by = "Dose")
contrast(CON, "pairwise", by = NULL)
```


---------------


```{r}

CON <- pairs(qq, by = "Condition")
contrast(CON, "consec", by = NULL)

```



## Paired comparisons in emmeans


```{r}
emmeans(fit.2, pairwise  ~  Dose * Condition)
```

## effect coding

```{r}

qq.f<- emmeans(fit.2, eff  ~  Dose * Condition)
qq.f
```

-----------

```{r}
emmeans(fit.2, eff ~  Dose | Condition)
```

---------------


```{r}

emmeans(fit.2, eff ~  Condition | Dose)
```





----------

```{r}
contrasts(simulated_data$Dose) <- contr.sum(3)
contrasts(simulated_data$Condition) <- contr.sum(2)

fit.3 <- lm(dv ~ 1 + Dose * Condition, simulated_data)

summary(fit.3)

```


