---
title: GLM basics 
format: revealjs
slide-number: true
editor: visual
execute:
  echo: true
html:
  code-fold: true
  code-summary: Show the code
  scrollable: true
---


## This time

- Getting comfortable with regression/glm 
- More in depth on each component of the model

## Thinking in terms of models

-   Our DV (here forth Y) is what we are trying to understand

-   We hypothesize it has some relationship with your IV(s) (here forth Xs), with what is left over described as error (E)

$Y = b_0 + b_{1}X + e$

- Your model is how you think the world works

## How can we visualize data to make sense of it?

```{r}
#| code-fold: true
library(broom)
library(tidyverse)
set.seed(123)
x.1 <- rnorm(100, 0, 1)
e.1 <- rnorm(100, 0, 2)
y.1 <- .5 + .55 * x.1 + e.1
d.1 <- data.frame(x.1,y.1)
m.1 <- lm(y.1 ~ x.1, data = d.1)
d1.f<- augment(m.1)
d.1
```

------------------------------------------------------------------------

```{r}
#| code-fold: true
ggplot(d1.f , aes(x=x.1, y=y.1)) +
    geom_point(size = 2) +
  geom_smooth(method = lm, se = FALSE) 
```

## General linear model (GLM)

-   This model (equation) can be very simple as in a treatment/control experiment

-   It can be very complex in terms of trying to understand something like academic achievement

-   The majority of our models fall under the umbrella of a general(ized) linear model (often referred to as regression models)

-   Models imply our theory about how the data are generated (ie how the world works)

## Parts of the model

$$Y_i = b_{0} + b_{1}X_i + e_i$$

-   Each individual has a unique Y value an X value and a residual/error term\
-   The model only has a single $b_{0}$ and $b_{1}$ term. These are the regression parameters. $b_{0}$ is the intercept and $b_{1}$ quantifies the relationship between your model of the world and the DV.



## in R

```{r, eval = FALSE}

fit <- lm(Y ~ X, data)
```

1.  Name regression object
2.  Use linear model function, lm
3.  Variable for your DV/Y
4.  \~ is interpretted as regressed on
5.  Variable for your IV/X

## What do the estimates tell us?

$$Y_i = b_{0} + b_{1}X_i + e_i$$

```{r}
#| code-fold: true
coef(m.1)
```

- Estimates are specific values for our coefficients (B or $\hat{\beta}$) which represent our best *guess* of the true population parameter ($\beta$) 

## How to interpret regression estimates

-   The entire class will go over different ways to interpret these estimates/parameters/coefficients

-   Intercept (b0) signifies the level of Y when your model IVs (Xs) are zero

-   Regression (b1) signifies the difference for a one unit change in your X

## Regression coefficient

**"For a one unit change in X, there is a b1 predicted change in Y."**

$$\large b_{1} = \frac{cov_{XY}}{s_{x}^{2}} = r_{xy} \frac{s_{y}}{s_{x}}$$ $$\large r_{xy} = \frac{s_{xy}}{s_xs_y}$$

------------------------------------------------------------------------

```{r}
#| code-fold: true
d.1 |> 
cor()
```

```{r}
#| code-fold: true
library(report)
report_sample(d.1)
```

```{r}
.2058 *(1.97/.91) 

```


## pop quiz

If I flipped the x and y variables, would I get the same regression coefficient?


## pop quiz #2

What would the regression coefficient be if I standardized (z-scored) the variables? The intercept? 

------------------------------------------------------------------------

```{r}
#standardized by hand
d.2 <- d.1 |> 
  mutate(x.z = (x.1 - mean(x.1))/sd(x.1)) |> 
  mutate(y.z = (y.1 - mean(y.1))/sd(y.1))

d.2
```

------------------------------------------------------------------------

```{r}
library(easystats)
m.2 <- lm(y.z ~ x.z, data = d.2)
model_parameters(m.2)
```

## Intercept

-   The intercept serves to adjust for differences in means between x and y.

-   What is the intercept in a standardized regression?

-   What happens to the intercept if I add 5 to x?

------------------------------------------------------------------------

```{r}
d.3 <- d.1 |> 
  mutate(x.5 = (x.1 + 5) ) 
d.3
```

------------------------------------------------------------------------

```{r}
m.3 <- lm(y.1 ~ x.5, data = d.3)
model_parameters(m.3)
```

------------------------------------------------------------------------

```{r}
#| code-fold: true
ggplot(d.3 , aes(x=x.5, y=y.1)) +
    geom_point(size = 2) +
  geom_smooth(method = lm, se = FALSE) 
```

------------------------------------------------------------------------

```{r}
#| code-fold: true
ggplot(d1.f , aes(x=x.1, y=y.1)) +
    geom_point(size = 2) +
  geom_smooth(method = lm, se = FALSE) 
```

------------------------------------------------------------------------

```{r}
model_parameters(m.1)
```

- Only the intercept changes. Not the slope. 
- What happens if we add 5 to Y variable? 

## Centroid

-   The mean of X and the mean of Y must always pass through the regression line

-   We will revisit this later in the semester as we discuss evaluating our models. The mean will be our baseline "guess" for using X to predict Y.

## Predictions

-   The regression line is made up of predictions
-   We will often call them $\hat{Y}$
-   "Best guess of Y, given some X"

```{r}
#| code-fold: true
ggplot(d1.f , aes(x=x.1, y=y.1)) +
    geom_point(size = 2) +
  geom_smooth(method = lm, se = FALSE) 
```

## Predictions

$$\hat{Y_i} = b_{0} + b_{1}X_i $$ 
- People who have the same IV, get the same DV. No error in our equation.


## Predictions

-   $\hat{Y_i}$ are also known as fitted values

```{r}
library(broom)
d1.f<- augment(m.1)
d1.f
```

## Predictions

- A regression model can be best thought of as prediction machine. Add hypothetical Xs in, get $\hat{Y}$ in return

- Regression can be thought of as telling us E(Y|X)

## Residuals

The residuals can be thought of as how far off we are in our predictions.  What is left over in Y that is not explained by our model (i.e. Xs).  

$$Y_i = b_{0} + b_{1}X_i + e_i$$

$$\hat{Y_i} = b_{0} + b_{1}X_i$$ 
$${Y_i} - \hat{Y_i} = e_i $$

## Residuals

```{r}
#| code-fold: true
ggplot(d1.f , aes(x=x.1, y=y.1)) +
    geom_point(size = 2) +
  geom_smooth(method = lm, se = FALSE) 
```

------------------------------------------------------------------------

```{r}
#| code-fold: true
ggplot(d1.f , aes(x=x.1, y=y.1)) +
    geom_point(size = 2) +
  geom_point(aes(y = .fitted), shape = 1, size = 2) 
```

------------------------------------------------------------------------

```{r}
#| code-fold: true
ggplot(d1.f , aes(x=x.1, y=y.1)) +
    geom_point(size = 2) +
  geom_point(aes(y = .fitted), shape = 1, size = 2) +
  geom_segment(aes( xend = x.1, yend = .fitted))
```

------------------------------------------------------------------------

```{r}
#| code-fold: true
ggplot(d1.f , aes(x=x.1, y=y.1)) +
    geom_point(size = 2) +
  geom_smooth(method = lm, se = FALSE) +
  geom_point(aes(y = .fitted), shape = 1, size = 2) +
  geom_segment(aes( xend = x.1, yend = .fitted))
```

------------------------------------------------------------------------

-   Thinking ahead: do we want large or small residuals?

```{r}
#| code-fold: true
new.i = 1.1
new.slope = -0.7
d1.f$new.fitted = 1.1 -0.7*d1.f$x.1

ggplot(d1.f , aes(x=x.1, y=y.1)) +
    geom_point(size = 2) +
  geom_abline(intercept = new.i, slope = new.slope, color = "blue", linewidth = 1) +
  geom_point(aes(y = new.fitted), shape = 1, size = 2) +
  geom_segment(aes( xend = x.1, yend = new.fitted))
```

## Residuals

-   Association between $\hat{Y_i}$ and e?

```{r}
#| code-fold: true
d1.f |> 
ggplot(aes(x = .fitted, y = .resid)) + geom_point() + geom_smooth(se = F, method = "lm") 
```

------------------------------------------------------------------------

-   x and residual?

```{r}
#| code-fold: true
d1.f |> 
ggplot(aes(x = x.1, y = .resid)) + geom_point() + geom_smooth(se = F, method = "lm") 
```

------------------------------------------------------------------------

-   x and $\hat{Y_i}$ represent the same information. We use x to create $\hat{Y_i}$

```{r}
#| code-fold: true
d1.f |> 
ggplot(aes(x = x.1, y = .fitted)) + geom_point() + geom_smooth(se = F, method = "lm") 
```


## Error vs residual

- Much like we have sample statistics and population parameters with $\bar{X}, \mu$

- Residual (e or $\hat{\epsilon}$)is the difference between our estimated regression line and our observed data.

- Error ($\epsilon$) is unknown parameter we estimate via the residual. It is the difference between our observed scores and our true population line.  

## Sigma

-   How do we know whether we have a lot or a little error in our model? 

```{r}
#| code-fold: true
d1.f<- augment(m.1)
mean(d1.f$.resid)
```
- Dispersion estimates! 

-   Sigma (σ) is a population parameter that represents the spread of error around the regression line. In our sample it reflects the SD of residuals. 


----------

```{r}
#| code-fold: true


# --- 1. Generate Sample Data ---
set.seed(123) # For reproducibility
n <- 100
x <- runif(n, 0, 50)
y <- 10 + 2 * x + rnorm(n, mean = 0, sd = 15)
data <- data.frame(x = x, y = y)

# --- 2. Fit the OLS Model ---
model <- lm(y ~ x, data = data)
data$predicted <- predict(model)
residual_sd <- sigma(model) # Get the residual standard deviation

# --- 3. Create the Base Plot ---
p <- ggplot(data, aes(x = x, y = y)) +
  geom_point(color = "gray50", alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE, color = "black",size = 1) +
  labs(
    title = "Regression Line with Residual Distributions",
    x = "X",
    y = "Y"
  ) 

# --- 4. Function to Generate Density Curve Data ---
# This function creates the points needed to draw a normal curve at a specific point
# on the regression line.
generate_density_data <- function(x_val, model, sd, scale = 5) {
  # Predict the y-value on the regression line
  y_hat <- predict(model, newdata = data.frame(x = x_val))
  
  # Generate points for a normal distribution curve
  y_curve <- seq(y_hat - 3 * sd, y_hat + 3 * sd, length.out = 100)
  density <- dnorm(y_curve, mean = y_hat, sd = sd)
  
  # Return a data frame with coordinates for the polygon
  data.frame(
    x_poly = x_val + density * scale, # Offset x by the density value
    y_poly = y_curve
  )
}

# --- 5. Add Density Curves to the Plot ---
# Choose a few x-values where you want to show the distributions
x_locations <- c(10, 15, 20, 25, 30, 35)

# Generate the data for each density curve
density_df1 <- generate_density_data(x_locations[1], model, residual_sd, scale = 80)
density_df2 <- generate_density_data(x_locations[2], model, residual_sd, scale = 80)
density_df3 <- generate_density_data(x_locations[3], model, residual_sd, scale = 80)
density_df4 <- generate_density_data(x_locations[4], model, residual_sd, scale = 80)
density_df5 <- generate_density_data(x_locations[5], model, residual_sd, scale = 80)
density_df6 <- generate_density_data(x_locations[6], model, residual_sd, scale = 80)

# Add the polygons to the plot
p + geom_polygon(data = density_df1, aes(x = x_poly, y = y_poly), alpha = 0.2) + geom_polygon(data = density_df2, aes(x = x_poly, y = y_poly), alpha = 0.2) + geom_polygon(data = density_df3, aes(x = x_poly, y = y_poly), alpha = 0.2) + geom_polygon(data = density_df4, aes(x = x_poly, y = y_poly), alpha = 0.2) + geom_polygon(data = density_df5, aes(x = x_poly, y = y_poly), alpha = 0.2) + geom_polygon(data = density_df6, aes(x = x_poly, y = y_poly), alpha = 0.2)

```

## Sigma

- Can be thought of as how much left over in Y that we cannot explain by our model. Often discussed in terms of Sums of Squares or variance (for math simplicity). 

$$\sum (Y_i - \bar{Y})^2 = \sum (\hat{Y_i} -\bar{Y})^2 + \sum(Y_i - \hat{Y_i})^2$$

- Total variation = regression (model) variation + residual variation


## Sum of Squared Errors (SSE)

- Reflects the "total" error in our model 

- SSE = $\sum(Y_i - \hat{Y_i})^2$

- To the extent that our actual values are close to the predicted values our model is doing better

## MSE & RSE

- Mean squared error is the "average" error in our model. RSE/sigma is our average error in original units. 

- Our estimate of sigma ($\hat{\sigma}$) is also known as Residual Standard Error (RSE), which is equal to the square root of the Mean Square Error (MSE).

- $\hat{\sigma}$ = $\sqrt{MSE}$ = $\sqrt{\frac{\sum(Y_i - \hat{Y_i})^2}{n-k-1}}$


## From population to sample

$$Y_i \sim \mathcal{N}(\mu, \sigma)$$

-   The DV, $Y$ is assumed to be distributed as a Gaussian normal, made up of $Y_i$, with a mean of $\mu$ and a standard deviation of $\sigma$

$$Y_i = b_{0} + b_{1}X_i + e_i$$

$$Y_i \sim \mathcal{N}(\beta_0 + \beta_1 X_i, \sigma)$$

## From population to sample

```{r}
#| code-fold: true
# Load the ggplot2 library for plotting
library(ggplot2)

# --- 1. Define the True Population Model ---
# These are the "true" parameters that we are trying to discover.
true_beta0 <- 10
true_beta1 <- 2
true_sigma <- 15
population_size <- 10000

# Create the entire population based on the true model
population <- data.frame(
  x = runif(population_size, 0, 50)
)
population$y <- true_beta0 + true_beta1 * population$x + rnorm(population_size, mean = 0, sd = true_sigma)

# --- 2. Draw a Random Sample from the Population ---
sample_size <- 50
sample_data <- population[sample(1:population_size, sample_size), ]

# --- 3. Fit a Regression Model to the SAMPLE ---
sample_model1 <- lm(y ~ x, data = sample_data)
sample_beta0_hat <- coef(sample_model1)[1]
sample_beta1_hat <- coef(sample_model1)[2]

# --- 4. Visualize the Relationship ---

pop <- ggplot(population, aes(x = y)) +
  geom_density(fill = "lightblue", alpha = 0.8, color = "black") +
  labs(
    title = "Distribution of the Outcome Variable (Y) in the Population",
    subtitle = "The overall distribution is approximately normal",
    x = "Outcome (Y)",
    y = "Density"
  ) +
  theme_minimal(base_size = 14)


pop

```

-------------

```{r}
#| code-fold: true
ggplot(population, aes(x = x, y = y)) +
# Plot the entire population as faint gray points
geom_point(color = "gray80", alpha = 0.5) +
# Add the TRUE population regression line in red
geom_abline(
intercept = true_beta0,
slope = true_beta1,
color = "black",
size = 1.5,
linetype = "dashed"
) +
# Highlight the points from our random sample in blue
geom_point(data = sample_data, color = "blue", size = 2.5) +
# Add the ESTIMATED sample regression line in blue
geom_abline(
intercept = sample_beta0_hat,
slope = sample_beta1_hat,
color = "green",
size = 1.5
) +
labs(subtitle = "The green line is our sample estimate of the population (black)",
x = "Predictor (X)",
y = "Outcome (Y)"
) +
theme_minimal(base_size = 14)
```




## From population to sample

```{r}
#| code-fold: true
# Load the ggplot2 library for plotting
library(ggplot2)
library(dplyr) # For data manipulation

# --- 1. Define the True Population Model ---
# These are the "true" parameters.
true_beta0 <- 50  # Mean of the group where X=0
true_beta1 <- 20  # The true DIFFERENCE between the group means
true_sigma <- 15
population_size <- 20000

# Create the population
population <- data.frame(
  # Create a binary predictor variable (0 or 1)
  x = factor(sample(c(0, 1), population_size, replace = TRUE))
)
# The model Y = b0 + b1*X + error still works.
# If X=0, Y = b0. If X=1, Y = b0 + b1.
population$y <- true_beta0 + true_beta1 * as.numeric(as.character(population$x)) + rnorm(population_size, mean = 0, sd = true_sigma)

# --- 2. Draw a Random Sample from the Population ---
sample_size <- 60
sample_data <- population[sample(1:population_size, sample_size), ]

# --- 3. Fit a Regression Model to the SAMPLE ---
sample_model <- lm(y ~ x, data = sample_data)
# Get the estimated intercept and slope (difference)
sample_beta0_hat <- coef(sample_model)[1]
sample_beta1_hat <- coef(sample_model)[2]

# --- 4. Calculate True and Estimated Group Means ---
# True population means
true_means <- data.frame(
  x = factor(c(0, 1)),
  mean_y = c(true_beta0, true_beta0 + true_beta1)
)
# Estimated means from our sample
estimated_means <- data.frame(
  x = factor(c(0, 1)),
  mean_y = c(sample_beta0_hat, sample_beta0_hat + sample_beta1_hat)
)

# --- 5. Visualize  

ggplot(population, aes(x = y)) +
  geom_density(fill = "lightblue", alpha = 0.8, color = "black") +
  labs(
    title = "Distribution of the Outcome Variable (Y) in the Population",
    subtitle = "The overall distribution is approximately normal",
    x = "Outcome (Y)",
    y = "Density"
  ) +
  theme_minimal(base_size = 14)



```


-----------------

```{r}
#| code-fold: true
ggplot(population, aes(x = x, y = y)) +
  # Jitter plot of the entire population
  geom_jitter(color = "gray80", alpha = 0.3, width = 0.2) +
  
  # Highlight the points from our random sample
  geom_jitter(data = sample_data, color = "blue", width = 0.2, size = 2) +
  
  # Add the TRUE population means in black
  geom_point(data = true_means, aes(y = mean_y), color = "Black", size = 28, shape = 95) +
  
  # Add the ESTIMATED sample means in blue
  geom_point(data = estimated_means, aes(y = mean_y), color = "green", size = 28, shape = 95) +
  
  labs(
    title = "Regression with a Dichotomous Predictor",
    subtitle = "Sample estimates (green bars) are close to the population black bars",
    x = "Group (Predictor X)",
    y = "Outcome (Y)"
  ) +
  theme_minimal(base_size = 14)
```


------------

What would happen if we increased sigma? 


$$Y_i \sim \mathcal{N}(\mu, \sigma)$$


$$Y_i \sim \mathcal{N}(\beta_0 + \beta_1 X_i, \sigma)$$

-----------

```{r}
#| code-fold: true
# Load the ggplot2 library for plotting
library(ggplot2)

set.seed(24)

# --- 1. Define the True Population Model ---
# These are the "true" parameters that we are trying to discover.
true_beta0 <- 10
true_beta1 <- 2
true_sigma <- 75
population_size <- 10000

# Create the entire population based on the true model
population <- data.frame(
  x = runif(population_size, 0, 50)
)
population$y <- true_beta0 + true_beta1 * population$x + rnorm(population_size, mean = 0, sd = true_sigma)

# --- 2. Draw a Random Sample from the Population ---
sample_size <- 50
sample_data <- population[sample(1:population_size, sample_size), ]

# --- 3. Fit a Regression Model to the SAMPLE ---
sample_model2 <- lm(y ~ x, data = sample_data)
sample_beta0_hat <- coef(sample_model2)[1]
sample_beta1_hat <- coef(sample_model2)[2]

ggplot(population, aes(x = x, y = y)) +
# Plot the entire population as faint gray points
geom_point(color = "gray80", alpha = 0.5) +
# Add the TRUE population regression line in red
geom_abline(
intercept = true_beta0,
slope = true_beta1,
color = "black",
size = 1.5,
linetype = "dashed"
) +
# Highlight the points from our random sample in blue
geom_point(data = sample_data, color = "blue", size = 2.5) +
# Add the ESTIMATED sample regression line in blue
geom_abline(
intercept = sample_beta0_hat,
slope = sample_beta1_hat,
color = "green",
size = 1.5
) +
labs(subtitle = "The green line is our sample estimate of the population (black)",
x = "Predictor (X)",
y = "Outcome (Y)"
) +
theme_minimal(base_size = 14)
```

----------

```{r}
sigma(sample_model2)
```

```{r}
sigma(sample_model1)
```
- Our regression line better reflects the population in our first simulation compared to this last one


## Sigma

- Our sigma reflects the conditional spread in the population. If there is low sigma there is low noise. We can estimate the population relatively easy with less sample size. (signal vs noise)

- Slope estimate is noisier and less precise with larger sigma, but not biased! If we sampled an infinite number of times the average slope would be correct

## Sigma

- Best to think of sigma as a conditional value that is dependent on what is in your model. As your model gets better, there is less unexplained noise in the data. 

$$Y_i \sim \mathcal{N}(\mu, \sigma)$$

$$Y_i \sim \mathcal{N}(\beta_0 + \beta_1 X_i, \sigma)$$
- We want smaller sigma, because that means a better model

## Sigma

Reasons for larger vs smaller sigma are aplenty: 
- Omitted variables 
- Measurement error/poor experimental design
- Population heterogeneity
- Inherent randomness


## Wrap-up

- Interpretation of coefficients
- The line is made up of predictions
- Predictions
- Residuals
- Sigma/SSE/MSE/RSE



