---
title: GLM basics 
format: revealjs
slide-number: true
editor: visual
execute:
  echo: true
html:
  code-fold: true
  code-summary: Show the code
  scrollable: true
---



## Thinking in terms of models

-   Our DV (here forth Y) is what we are trying to understand

-   We hypothesize it has some relationship with your IV(s) (here forth Xs), with what is left over described as error (E)

$y = b_0 + b_{1}X + e$

## How can we visualize data to make sense of it?

```{r}
#| code-fold: true
library(broom)
library(tidyverse)
set.seed(123)
x.1 <- rnorm(100, 0, 1)
e.1 <- rnorm(100, 0, 2)
y.1 <- .5 + .55 * x.1 + e.1
d.1 <- data.frame(x.1,y.1)
m.1 <- lm(y.1 ~ x.1, data = d.1)
d1.f<- augment(m.1)
d.1
```

------------------------------------------------------------------------

```{r}
#| code-fold: true
ggplot(d1.f , aes(x=x.1, y=y.1)) +
    geom_point(size = 2) +
  geom_smooth(method = lm, se = FALSE) 
```
## General linear model (GLM)

-   This model (equation) can be very simple as in a treatment/control experiment

-   It can be very complex in terms of trying to understand something like academic achievement

-   The majority of our models fall under the umbrella of a general(ized) linear model (often referred to as regression models)

-   Models imply our theory about how the data are generated (ie how the world works)


## Parts of the model

$$Y_i = b_{0} + b_{1}X_i + e_i$$ 

-   Each individual has a unique Y value an X value and a residual/error term\
-   The model only has a single $b_{0}$ and $b_{1}$ term. These are the regression parameters. $b_{0}$ is the intercept and $b_{1}$ quantifies the relationship between your model of the world and the DV.


## in R

```{r, eval = FALSE}

fit <- lm(Y ~ X, data)
```

1. Name regression object 
2. Use linear model function, lm
3. Variable for your DV/Y
4. ~ is interpretted as regressed on
5. Variable for your IV/X

## What do the estimates tell us?

$$Y_i = b_{0} + b_{1}X_i + e_i$$ 

```{r}
coef(m.1)
```


## How to interpret regression estimates

-   The entire class will go over different ways to interpret these estimates/parameters/coefficients

-   Intercept (b0) signifies the level of Y when your model IVs (Xs) are zero

-   Regression (b1) signifies the difference for a one unit change in your X




## Regression coefficient

**"For a one unit change in X, there is a b1 predicted change in Y."**


$$\large b_{1} = \frac{cov_{XY}}{s_{x}^{2}} = r_{xy} \frac{s_{y}}{s_{x}}$$
$$\large r_{xy} = \frac{s_{xy}}{s_xs_y}$$


--------------

```{r}
d.1 |> 
cor()
```

```{r}
library(report)
report_sample(d.1)
```

## pop quiz

If I flipped the x and y variables, would I get the same regression coefficient? 


## pop quiz #2

What would the regression coefficient be if I standardized (z-scored) the variables? 


--------------------

```{r}
d.2 <- d.1 |> 
  mutate(x.z = (x.1 - mean(x.1))/sd(x.1)) |> 
  mutate(y.z = (y.1 - mean(y.1))/sd(y.1))

d.2
```

-------------

```{r}
library(easystats)
m.2 <- lm(y.z ~ x.z, data = d.2)
model_parameters(m.2)
```


## Intercept

- The intercept serves to adjust for differences in means between x and y. 

- What is the intercept in a standardized regression? 

- What happens to the intercept if I add 5 to x? 

-------

```{r}
d.3 <- d.1 |> 
  mutate(x.5 = (x.1 + 5) )
d.3
```

----------

```{r}
m.3 <- lm(y.1 ~ x.5, data = d.3)
model_parameters(m.3)
```




---------

```{r}
#| code-fold: true
ggplot(d.3 , aes(x=x.5, y=y.1)) +
    geom_point(size = 2) +
  geom_smooth(method = lm, se = FALSE) 
```



----------

```{r}
#| code-fold: true
ggplot(d1.f , aes(x=x.1, y=y.1)) +
    geom_point(size = 2) +
  geom_smooth(method = lm, se = FALSE) 
```

--------------

```{r}
model_parameters(m.1)
```


## Centroid

- The mean of X and the mean of Y must always pass through the regression line

- We will revisit this later in the semester as we discuss evaluating our models. The mean will be our baseline "guess" for using X to predict Y. 


## Predictions

- The regression line is made up of predictions
- We will often call them $\hat{Y}$

```{r}
#| code-fold: true
ggplot(d1.f , aes(x=x.1, y=y.1)) +
    geom_point(size = 2) +
  geom_smooth(method = lm, se = FALSE) 
```

## Predictions 


$$\hat{Y_i} = b_{0} + b_{1}X_i $$ 
- People who have the same IV, get the same DV. No error in our equation. 


## Predictions

- $\hat{Y_i}$ are also known as fitted values

```{r}
d1.f<- augment(m.1)
d1.f
```




## Residuals

The residuals can be thought of as what is left over in Y that is not explained by our model. ie difference between actual and predicted 

$$Y_i = b_{0} + b_{1}X_i + e_i$$ 

$$\hat{Y_i} = b_{0} + b_{1}X_i$$ 
$${Y_i} - \hat{Y_i} = e_i $$



## Residuals

```{r}
ggplot(d1.f , aes(x=x.1, y=y.1)) +
    geom_point(size = 2) +
  geom_smooth(method = lm, se = FALSE) 
```

---

```{r, echo=FALSE}
ggplot(d1.f , aes(x=x.1, y=y.1)) +
    geom_point(size = 2) +
  geom_point(aes(y = .fitted), shape = 1, size = 2) 
```


---

```{r, echo=FALSE}
ggplot(d1.f , aes(x=x.1, y=y.1)) +
    geom_point(size = 2) +
  geom_point(aes(y = .fitted), shape = 1, size = 2) +
  geom_segment(aes( xend = x.1, yend = .fitted))
```



---

```{r, echo=FALSE, wanring = FALSE, message = FALSE}
ggplot(d1.f , aes(x=x.1, y=y.1)) +
    geom_point(size = 2) +
  geom_smooth(method = lm, se = FALSE) +
  geom_point(aes(y = .fitted), shape = 1, size = 2) +
  geom_segment(aes( xend = x.1, yend = .fitted))
```


------

- Thinking ahead: do we want large or small residuals? 

```{r, echo = F}
new.i = 1.1
new.slope = -0.7
d1.f$new.fitted = 1.1 -0.7*d1.f$x.1

ggplot(d1.f , aes(x=x.1, y=y.1)) +
    geom_point(size = 2) +
  geom_abline(intercept = new.i, slope = new.slope, color = "blue", linewidth = 1) +
  geom_point(aes(y = new.fitted), shape = 1, size = 2) +
  geom_segment(aes( xend = x.1, yend = new.fitted))
```


## Residuals

- Association between $\hat{Y_i}$ and e?

```{r}
d1.f |> 
ggplot(aes(x = .fitted, y = .resid)) + geom_point() + geom_smooth(se = F, method = "lm") 
```



-------

- x and residual? 

```{r}
d1.f |> 
ggplot(aes(x = x.1, y = .resid)) + geom_point() + geom_smooth(se = F, method = "lm") 
```


--------

- x and $\hat{Y_i}$ represent the same information. We use x to create $\hat{Y_i}$

```{r}
d1.f |> 
ggplot(aes(x = x.1, y = .fitted)) + geom_point() + geom_smooth(se = F, method = "lm") 
```



## Sigma

- How do we know whether we have a lot or a little residuals? 

- Sigma (σ) is the SD of residuals. It can be thought of as how much left over in Y that we cannot explain by our model.



