---
title: GLM basics 
format: revealjs
slide-number: true
editor: visual
execute:
  echo: true
html:
  code-fold: true
  code-summary: Show the code
  scrollable: true
---


## This time

- Getting comformatable with regression/glm 
- More in depth on each component of the model

## Thinking in terms of models

-   Our DV (here forth Y) is what we are trying to understand

-   We hypothesize it has some relationship with your IV(s) (here forth Xs), with what is left over described as error (E)

$Y = b_0 + b_{1}X + e$

*- Your model is how you think the world works*

## How can we visualize data to make sense of it?

```{r}
#| code-fold: true
library(broom)
library(tidyverse)
set.seed(123)
x.1 <- rnorm(100, 0, 1)
e.1 <- rnorm(100, 0, 2)
y.1 <- .5 + .55 * x.1 + e.1
d.1 <- data.frame(x.1,y.1)
m.1 <- lm(y.1 ~ x.1, data = d.1)
d1.f<- augment(m.1)
d.1
```

------------------------------------------------------------------------

```{r}
#| code-fold: true
ggplot(d1.f , aes(x=x.1, y=y.1)) +
    geom_point(size = 2) +
  geom_smooth(method = lm, se = FALSE) 
```

## General linear model (GLM)

-   This model (equation) can be very simple as in a treatment/control experiment

-   It can be very complex in terms of trying to understand something like academic achievement

-   The majority of our models fall under the umbrella of a general(ized) linear model (often referred to as regression models)

-   Models imply our theory about how the data are generated (ie how the world works)

## Parts of the model

$$Y_i = b_{0} + b_{1}X_i + e_i$$

-   Each individual has a unique Y value an X value and a residual/error term\
-   The model only has a single $b_{0}$ and $b_{1}$ term. These are the regression parameters. $b_{0}$ is the intercept and $b_{1}$ quantifies the relationship between your model of the world and the DV.

## in R

```{r, eval = FALSE}

fit <- lm(Y ~ X, data)
```

1.  Name regression object
2.  Use linear model function, lm
3.  Variable for your DV/Y
4.  \~ is interpretted as regressed on
5.  Variable for your IV/X

## What do the estimates tell us?

$$Y_i = b_{0} + b_{1}X_i + e_i$$

```{r}
#| code-fold: true
coef(m.1)
```

- Estimates are specific values for our coefficients (B or $\hat{\beta}$) which represent our best *guess* of the true population parameter ($\beta$) 

## How to interpret regression estimates

-   The entire class will go over different ways to interpret these estimates/parameters/coefficients

-   Intercept (b0) signifies the level of Y when your model IVs (Xs) are zero

-   Regression (b1) signifies the difference for a one unit change in your X

## Regression coefficient

**"For a one unit change in X, there is a b1 predicted change in Y."**

$$\large b_{1} = \frac{cov_{XY}}{s_{x}^{2}} = r_{xy} \frac{s_{y}}{s_{x}}$$ $$\large r_{xy} = \frac{s_{xy}}{s_xs_y}$$

------------------------------------------------------------------------

```{r}
#| code-fold: true
d.1 |> 
cor()
```

```{r}
#| code-fold: true
library(report)
report_sample(d.1)
```

```{r}
.2058 *(1.97/.91) 

```


## pop quiz

If I flipped the x and y variables, would I get the same regression coefficient?


## pop quiz #2

What would the regression coefficient be if I standardized (z-scored) the variables? The intercept? 

------------------------------------------------------------------------

```{r}
#standardized by hand
d.2 <- d.1 |> 
  mutate(x.z = (x.1 - mean(x.1))/sd(x.1)) |> 
  mutate(y.z = (y.1 - mean(y.1))/sd(y.1))

d.2
```

------------------------------------------------------------------------

```{r}
library(easystats)
m.2 <- lm(y.z ~ x.z, data = d.2)
model_parameters(m.2)
```

## Intercept

-   The intercept serves to adjust for differences in means between x and y.

-   What is the intercept in a standardized regression?

-   What happens to the intercept if I add 5 to x?

------------------------------------------------------------------------

```{r}
d.3 <- d.1 |> 
  mutate(x.5 = (x.1 + 5) ) 
d.3
```

------------------------------------------------------------------------

```{r}
m.3 <- lm(y.1 ~ x.5, data = d.3)
model_parameters(m.3)
```

------------------------------------------------------------------------

```{r}
#| code-fold: true
ggplot(d.3 , aes(x=x.5, y=y.1)) +
    geom_point(size = 2) +
  geom_smooth(method = lm, se = FALSE) 
```

------------------------------------------------------------------------

```{r}
#| code-fold: true
ggplot(d1.f , aes(x=x.1, y=y.1)) +
    geom_point(size = 2) +
  geom_smooth(method = lm, se = FALSE) 
```

------------------------------------------------------------------------

```{r}
model_parameters(m.1)
```

- Only the intercept changes. Not the slope. 
- What happens if we add 5 to Y variable? 

## Centroid

-   The mean of X and the mean of Y must always pass through the regression line

-   We will revisit this later in the semester as we discuss evaluating our models. The mean will be our baseline "guess" for using X to predict Y.

## Predictions

-   The regression line is made up of predictions
-   We will often call them $\hat{Y}$
-   "Best guess of Y, given some X"

```{r}
#| code-fold: true
ggplot(d1.f , aes(x=x.1, y=y.1)) +
    geom_point(size = 2) +
  geom_smooth(method = lm, se = FALSE) 
```

## Predictions

$$\hat{Y_i} = b_{0} + b_{1}X_i $$ 
- People who have the same IV, get the same DV. No error in our equation.


## Predictions

-   $\hat{Y_i}$ are also known as fitted values

```{r}
library(broom)
d1.f<- augment(m.1)
d1.f
```

## Predictions

- A regression model can be best thought of as prediction machine. Add hypothetical Xs in, get $\hat{Y}$ in return

- Regression can be thought of as telling us E(Y|X)

## Residuals

The residuals can be thought of as how far off we are in our predictions.  What is left over in Y that is not explained by our model (i.e. Xs).  

$$Y_i = b_{0} + b_{1}X_i + e_i$$

$$\hat{Y_i} = b_{0} + b_{1}X_i$$ 
$${Y_i} - \hat{Y_i} = e_i $$

## Residuals

```{r}
#| code-fold: true
ggplot(d1.f , aes(x=x.1, y=y.1)) +
    geom_point(size = 2) +
  geom_smooth(method = lm, se = FALSE) 
```

------------------------------------------------------------------------

```{r}
#| code-fold: true
ggplot(d1.f , aes(x=x.1, y=y.1)) +
    geom_point(size = 2) +
  geom_point(aes(y = .fitted), shape = 1, size = 2) 
```

------------------------------------------------------------------------

```{r}
#| code-fold: true
ggplot(d1.f , aes(x=x.1, y=y.1)) +
    geom_point(size = 2) +
  geom_point(aes(y = .fitted), shape = 1, size = 2) +
  geom_segment(aes( xend = x.1, yend = .fitted))
```

------------------------------------------------------------------------

```{r}
#| code-fold: true
ggplot(d1.f , aes(x=x.1, y=y.1)) +
    geom_point(size = 2) +
  geom_smooth(method = lm, se = FALSE) +
  geom_point(aes(y = .fitted), shape = 1, size = 2) +
  geom_segment(aes( xend = x.1, yend = .fitted))
```

------------------------------------------------------------------------

-   Thinking ahead: do we want large or small residuals?

```{r}
#| code-fold: true
new.i = 1.1
new.slope = -0.7
d1.f$new.fitted = 1.1 -0.7*d1.f$x.1

ggplot(d1.f , aes(x=x.1, y=y.1)) +
    geom_point(size = 2) +
  geom_abline(intercept = new.i, slope = new.slope, color = "blue", linewidth = 1) +
  geom_point(aes(y = new.fitted), shape = 1, size = 2) +
  geom_segment(aes( xend = x.1, yend = new.fitted))
```

## Residuals

-   Association between $\hat{Y_i}$ and e?

```{r}
#| code-fold: true
d1.f |> 
ggplot(aes(x = .fitted, y = .resid)) + geom_point() + geom_smooth(se = F, method = "lm") 
```

------------------------------------------------------------------------

-   x and residual?

```{r}
#| code-fold: true
d1.f |> 
ggplot(aes(x = x.1, y = .resid)) + geom_point() + geom_smooth(se = F, method = "lm") 
```

------------------------------------------------------------------------

-   x and $\hat{Y_i}$ represent the same information. We use x to create $\hat{Y_i}$

```{r}
#| code-fold: true
d1.f |> 
ggplot(aes(x = x.1, y = .fitted)) + geom_point() + geom_smooth(se = F, method = "lm") 
```


## Error vs residual

- Much like we have sample statistics and population parameters with $\bar{X}, \mu$

- Residual (e or $\hat{\epsilon}$)is the difference between our estimated regression line and our observed data.

- Error ($\epsilon$) is unknown parameter we estimate via the residual. It is the difference between our observed scores and our true population line.  

## Sigma

-   How do we know whether we have a lot or a little error in our model? 

```{r}
#| code-fold: true
d1.f<- augment(m.1)
mean(d1.f$.resid)
```
- Dispersion estimates! 

-   Sigma (σ) is a population parameter that represents the spread of error around the regression line. In our sample it reflects the SD of residuals. 


----------

```{r}
# Load the ggplot2 library for plotting
library(ggplot2)

# --- 1. Generate Sample Data ---
set.seed(123) # For reproducibility
n <- 100
x <- runif(n, 0, 50)
y <- 10 + 2 * x + rnorm(n, mean = 0, sd = 15)
data <- data.frame(x = x, y = y)

# --- 2. Fit the OLS Model ---
model <- lm(y ~ x, data = data)
data$predicted <- predict(model)
residual_sd <- sigma(model) # Get the residual standard deviation

# --- 3. Create the Base Plot ---
p <- ggplot(data, aes(x = x, y = y)) +
  geom_point(color = "gray50", alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE, color = "black",size = 1) +
  labs(
    title = "Regression Line with Residual Distributions",
    x = "X",
    y = "Y"
  ) 

# --- 4. Function to Generate Density Curve Data ---
# This function creates the points needed to draw a normal curve at a specific point
# on the regression line.
generate_density_data <- function(x_val, model, sd, scale = 5) {
  # Predict the y-value on the regression line
  y_hat <- predict(model, newdata = data.frame(x = x_val))
  
  # Generate points for a normal distribution curve
  y_curve <- seq(y_hat - 3 * sd, y_hat + 3 * sd, length.out = 100)
  density <- dnorm(y_curve, mean = y_hat, sd = sd)
  
  # Return a data frame with coordinates for the polygon
  data.frame(
    x_poly = x_val + density * scale, # Offset x by the density value
    y_poly = y_curve
  )
}

# --- 5. Add Density Curves to the Plot ---
# Choose a few x-values where you want to show the distributions
x_locations <- c(10, 15, 20, 25, 30, 35)

# Generate the data for each density curve
density_df1 <- generate_density_data(x_locations[1], model, residual_sd, scale = 80)
density_df2 <- generate_density_data(x_locations[2], model, residual_sd, scale = 80)
density_df3 <- generate_density_data(x_locations[3], model, residual_sd, scale = 80)
density_df4 <- generate_density_data(x_locations[4], model, residual_sd, scale = 80)
density_df5 <- generate_density_data(x_locations[5], model, residual_sd, scale = 80)
density_df6 <- generate_density_data(x_locations[6], model, residual_sd, scale = 80)

# Add the polygons to the plot
p + geom_polygon(data = density_df1, aes(x = x_poly, y = y_poly), alpha = 0.2) + geom_polygon(data = density_df2, aes(x = x_poly, y = y_poly), alpha = 0.2) + geom_polygon(data = density_df3, aes(x = x_poly, y = y_poly), alpha = 0.2) + geom_polygon(data = density_df4, aes(x = x_poly, y = y_poly), alpha = 0.2) + geom_polygon(data = density_df5, aes(x = x_poly, y = y_poly), alpha = 0.2) + geom_polygon(data = density_df6, aes(x = x_poly, y = y_poly), alpha = 0.2)

```

## Sigma

- Can be thought of as how much left over in Y that we cannot explain by our model. Often discussed in terms of Sums of Squares or variance (for math simplicity). 

$$\sum (Y_i - \bar{Y})^2 = \sum (\hat{Y_i} -\bar{Y})^2 + \sum(Y_i - \hat{Y_i})^2$$

- Total variation = regression (model) variation + residual variation


## Sum of Squared Errors (SSE)

- Reflects the "total" error in our model 

- SSE = $\sum(Y_i - \hat{Y_i})^2$

- To the extent that our actual values are close to the predicted values our model is doing better

## MSE & RSE

- Mean squared error is the "average" error in our model. RSE/sigma is our average error in original units. 

- Our estimate of sigma ($\hat{\sigma}$) is also known as Residual Standard Error (RSE), which is equal to the square root of the Mean Square Error (MSE).

- $\hat{\sigma}$ = $\sqrt{MSE}$ = $\sqrt{\frac{\sum(Y_i - \hat{Y_i})^2}{n-k-1}}$

## Wrap-up

- Interpretation of coefficients
- The line is made up of predictions
- Predictions
- Residuals
- Sigma/SSE/MSE/RSE



