---
title: "hw3"
format: html
---

# Instructions

Please complete this assignment using a quarto or RMarkdown file. Be sure to include your name in the YAML/preamble. When you turn in the assignment, include ***both*** the .qmd or .Rmd and knitted .html files. Please turn in both documents to the **ASSIGNMENTS** page on Canvas.

Put all R code in code chunks. They will behave as normal R code. If you are asked to explain something or write a few sentences, please do that *outside* of the code chunks. That way, when you knit, your sentences will show up as normal text, but all of your code will be in a designated code chunk.

You are allowed to Google or use your preferred AI-overlord. *However*, if you use code from an outside source (meaning not the lecture slides or the R help documentation), then you must provide a **working link** to the website you used. You don't need to use APA formatting or anything. Just a working url. You may discuss homework assignments with your classmates; however, it is important that you complete each assignment on your own and **do not simply copy someone else’s code**. If we believe one student has copied another’s work, both students will receive a 0 on the homework assignment. Please also list at the top of the document who you worked with.

In the instructions, when it says ***by hand***, what we mean is that you should not use a function that is built into R to complete the task. For example, if I ask you to find the mean *by hand*, you should **not** use the `mean()` function. However, you can (and should!) use your variables. So for instance, if I ask you to find the `mean()` by hand, you may do something like `sum(variableX)/length(variableX)`. This would be an appropriate *by hand* calculation.

**Data:** You will simulate your own data. FUN.


1. Describe the relationship between standard error and power estimates.  

2. Why is 80% power often considered a desirable target? What does this benchmark imply about the balance between Type I and Type II errors?

3. Can something be "over" powered? 

4. Someone, lets call them J. Jackson, thinks that power puts too much emphasis on dichotomous decisions due to p-values. Instead, he suggests that one should do "precision estimates" rather than "power estimates."   a. What is one pro and one con to this approach?  
 b. For your own research topic, describe an association between two variables and the range of CI (in original units) that would be meaningful to your research question. 
 
5. Below is a power simulation for a regression (and a reason to force you to run a for loop).  Run it. 

```{r}

# --- Parameters ---
n_simulations <- 1000
sample_size <- 100
alpha <- 0.05
true_intercept <- 0
true_slope <- 0.25 # Assumed true effect size
true_error_sd <- .9375 # Residual standard error
predictor_sd <- 1 # SD of the predictor 'x'
predictor_mean <- 0 # Mean of 'x' (doesn't affect power)

# --- Simulation ---
p_values_slope <- numeric(n_simulations)
slope <- numeric(n_simulations)
std_errors_slope <- numeric(n_simulations)

for (i in 1:n_simulations) {
  x <- rnorm(sample_size, mean = predictor_mean, sd = predictor_sd)
  y <- true_intercept + true_slope * x + rnorm(sample_size, mean = 0, sd = true_error_sd)
  simulated_data <- data.frame(x, y)
  model <- lm(y ~ x, data = simulated_data)
  model_summary <- summary(model)
  p_values_slope[i] <- model_summary$coefficients[2, 4]
  slope[i] <- model_summary$coefficients[2, 1]
  std_errors_slope[i] <- model_summary$coefficients[2, 2]
}

# --- Calculate Power & Average SE ---
power <- mean(p_values_slope < alpha)
avg_slope <- mean(slope)
avg_coeff_se <- mean(std_errors_slope)


```

A. What is the power? Is this good or bad? What min sample size would you need for this? 

B. What do the vectors p_values_slope, slope, std_errors_slope represent? Hint: are they all the same size?  

C. What CI would you put around the average simulation est? Does it overlap with zero or not? Is this problematic, reassuring or asking a different question? 

D. Graph a visualization of the power using this code from class. You will need to add stuff to it to make it run as well as delete the "eval = FALSE" in the first line. Then run a second visualization with a simulation that provides "good power." How do the two differ? 


```{r, eval = FALSE}

distributions_df <- tribble(
  ~distribution,       ~mu, ~sigma,
  "Null (H0)",         ??,   ??,
  "Alternative (H1)",   ??,   ??
)

# --- Step 2: Define the Critical Value ---
critical_value <- 1.96

ggplot(distributions_df, aes(
    xdist = distributional::dist_normal(mu = mu, sigma = sigma)
  )) +  stat_dist_slab(
    alpha = 0.4
  ) +

  stat_dist_slab(
    data = . %>% filter(distribution == "Alternative (H1)"),
    alpha = 0.6,
    fill = "green") +
  
  geom_vline(xintercept = critical_value, linetype = "dashed", color = "darkred", size = 1) +
  
  labs(
    title = "Beta (type 2 error) is AUC left of C.V.",
    subtitle = "Power is AUC right of C.V.",
    x = "Sample Mean",
    y = "Density"
  )

```


