---
title: "Statistical Inference"
format: revealjs
slide-number: true
editor: visual
execute:
  echo: true
html:
  code-fold: true
  code-summary: Show the code
  scrollable: true
---

## Last section

- First third of the class

1. Construction of model (make it work)
2. Inference from model (learn from it)
3. Additions to basic model (make it suit our needs)

## This time (this section)

-   Sampling distributions

-   NHST as evaluating our hypothesis

-   Estimation as alternative

-   Model comparison as third alternative

## Estimation of what?

-   We To do that we need to use an *Estimator* to get *Estimates*, which are our best guess of the population parameters.

-   The population parameters are your *Estimand* -- what you want to know but have to estimate

-   Picture of perfect cake (Estimand), Cake recipe (Estimator), your cake (Estimate)

## How good our are estimates?

-   Thus far we used the standard error to help us define the precision around the estimates... and I just gave you the equation

-   Where does this come from, and how can I think about the process?

-   Remember, we only get one shot to estimate the population.

------------------------------------------------------------------------

```{r}
#| code-fold: true
# Load the ggplot2 library for plotting
library(ggplot2)

# --- 1. Define the True Population Model ---
# These are the "true" parameters that we are trying to discover.
true_beta0 <- 10
true_beta1 <- 2
true_sigma <- 15
population_size <- 20000

# Create the entire population based on the true model
population <- data.frame(
  x = runif(population_size, 0, 50)
)
population$y <- true_beta0 + true_beta1 * population$x + rnorm(population_size, mean = 0, sd = true_sigma)

# --- 2. Draw a Random Sample from the Population ---
sample_size <- 50
sample_data <- population[sample(1:population_size, sample_size), ]

# --- 3. Fit a Regression Model to the SAMPLE ---
sample_model1 <- lm(y ~ x, data = sample_data)
sample_beta0_hat <- coef(sample_model1)[1]
sample_beta1_hat <- coef(sample_model1)[2]

# --- 4. Visualize the Relationship ---

pop <- ggplot(population, aes(x = y)) +
  geom_density(fill = "lightblue", alpha = 0.8, color = "black") +
  labs(
    title = "Distribution of the Outcome Variable (Y) in the Population",
    subtitle = "Y is determined by y = 10 + 2X + rnorm(0,15) ",
    x = "Outcome (Y)",
    y = "Density"
  ) +
  theme_minimal(base_size = 14)


pop

```

------------------------------------------------------------------------

```{r}
#| code-fold: true
ggplot(population, aes(x = x, y = y)) +
# Plot the entire population as faint gray points
geom_point(color = "gray80", alpha = 0.5) +
# Add the TRUE population regression line in red
geom_abline(
intercept = true_beta0,
slope = true_beta1,
color = "black",
size = 1.5,
linetype = "dashed"
) +
# Highlight the points from our random sample in blue
geom_point(data = sample_data, color = "blue", size = 2.5) +
# Add the ESTIMATED sample regression line in blue
geom_abline(
intercept = sample_beta0_hat,
slope = sample_beta1_hat,
color = "green",
size = 1.5
) +
labs(subtitle = "The green line is our sample estimate of the population (black)",
x = "Predictor (X)",
y = "Outcome (Y)"
) +
theme_minimal(base_size = 14)
```

## What we normally see

```{r}
#| code-fold: true
ggplot(sample_data, aes(x = x, y = y))+
geom_point() +
geom_smooth(method='lm', se = FALSE) 
```

------------------------------------------------------------------------

-   The major goal that we have in statistical inference is to make confident claims about the population based on a small representation of it, the sample. We use features of the sample (statistics) to inform us about features of the population (parameters).

-   Any sample will be off the mark in how well it captures the important features of a population due to *sampling variability*.

-   The sampling distribution tells us how far off the mark we can expect a sample statistic to be.

## Sampling distribution - empirical

-   Run 5000 experiments from a known population. Make a distribution of the 5000 regression coefs to tell us what we should expect from any 1 experiment.

```{r}
#| code-fold: true
library(ggplot2)

# --- Step 1: Create a "True" Population ---

true_intercept <- 10
true_slope <- 2
true_error_sd <- 15 # The true standard deviation of the residuals (sigma)
population_size <- 20000

# Create the entire population based on our true model
set.seed(74) # for reproducibility
population_data <- data.frame(
  x = runif(population_size, 0, 50)
)
population_data$y <- true_intercept + true_slope * population_data$x + rnorm(population_size, mean = 0, sd = true_error_sd)



# --- Step 2: Simulate the Process of Repeated Sampling ---

n_simulations <- 5000 # Let's take 5000 different samples
sample_size <- 100  # Each sample will have 10 data points
all_my_slopes <- numeric(n_simulations) # An empty vector to store our results

all_my_models <- data.frame(
  intercept = numeric(n_simulations),
  slope = numeric(n_simulations)
)

for (i in 1:n_simulations) {
  # Draw one random sample of 100 data points from our population
  current_sample <- population_data[sample(1:population_size, sample_size), ]
  
  # Fit a regression model to this one small sample
  sample_model <- lm(y ~ x, data = current_sample)
  
  # Extract and store the estimated slope from this sample's model
  all_my_models$intercept[i] <- coef(sample_model)[1]
  all_my_models$slope[i] <- coef(sample_model)[2]
}


# --- Step 3: Build and Analyze the Sampling Distribution ---
all_my_slopes <- all_my_models$slope
mean_of_slopes <- mean(all_my_slopes)


cat("The true slope is 2.0. The mean of our 5000 estimated slopes is:", round(mean_of_slopes, 4), "\n")

simulated_standard_error <- sd(all_my_slopes)
cat("The standard deviation of our slopes (the Simulated Standard Error) is:", round(simulated_standard_error, 4), "\n\n")


```

## Sampling distribution

-   For each sample, we calculate a statistic, and then create a distribution of sample statistics

```{r}
#| code-fold: true
# --- Step 5: Visualize Everything ---
# A histogram clearly shows the distribution of our estimates.
sampling_dist_plot <- ggplot(data.frame(slopes = all_my_slopes), aes(x = slopes)) +
  geom_histogram(aes(y = ..density..), bins = 30, fill = "skyblue", color = "black") +
  geom_density(color = "red", size = 1) +
  geom_vline(xintercept = true_slope, color = "blue", linetype = "dashed", size = 1.5) +
  labs(
    title = "Sampling Distribution of the Slope (β₁)",
    subtitle = paste("Based on", n_simulations, "samples of size", sample_size),
    x = "Estimated Slope Coefficient (β̂₁)",
    y = "Density"
  ) + 
  # geom_vline(2.014) + 
  theme_minimal()

print(sampling_dist_plot)

```

------------------------------------------------------------------------

-   We normally do not have access to this data, but if we did we now have a *distribution* of plausible statistics drawn from the population.

-   This distribution tells us what we would expect if we repeated the experiment again, and again, and again

- Regression coef is an unbiased estimator because we can find the true mean (if we repeat experiment infinite times)

-   Range of scores are SUPER helpful to gauge our precision of the estimate.

------------------------------------------------------------------------

-   If we look at just one sample (like normal experiments), we won't exactly estimate the population intercept (10) and slope (2).

```{r}
#| code-fold: true
# --- Step 4:Connecting to a Single Regression Summary ---
summary(sample_model) # Using the last model from our loop

```

## Sampling distribution

```{r}
#| code-fold: true
# --- Step 5: Visualize Everything ---
# A histogram clearly shows the distribution of our estimates.
sampling_dist_plot <- ggplot(data.frame(slopes = all_my_slopes), aes(x = slopes)) +
  geom_histogram(aes(y = ..density..), bins = 30, fill = "skyblue", color = "black") +
  geom_density(color = "red", size = 1) +
  geom_vline(xintercept = true_slope, color = "blue", linetype = "dashed", size = 1.5) +
  geom_vline(xintercept = 1.87082, color = "black", linetype = "solid", size = 1.5) +
  labs(
    title = "Sampling Distribution of the Slope (β₁)",
    subtitle = paste("Based on", n_simulations, "samples of size", sample_size),
    x = "Estimated Slope Coefficient (β̂₁)",
    y = "Density"
  ) + 
  # geom_vline(2.014) + 
  theme_minimal()

print(sampling_dist_plot)

```

- Did we get unlucky with our one experiment? Is this just part of the uncertainty we have to live with? 

## Sampling distribution

-   Can be for any statistic. Mean, SD, r, b, etc.

http://shiny.calpoly.sh/Sampling_Distribution/

-   If unbiased, the mean of the sampling distribution is equal to the population parameter

-   The standard deviation of the sampling distribution is called standard error, and helps us gauge our precision of the statistic.

## Sampling distribution

-   Given we don't have access to the population, nor do we have the time and resources to conduct thousands of replication studies, what do we do?

-   Our sampling distribution will be theoretical... and it will be normal.

## Central limit theorm take 2

-   Central Limit Theorem: the sum of a large number of independent random variables will be approximately normally distributed, even if original distributions are not normal.

-   We previously used this to show that the normal was everywhere. Here we are going to use it to define the shape of our sampling distribution.

-   The sample size necessary to be "approximately normal" depends on the nature of the underlying data. The less normal it is, the larger the sample size necessary in order for the sampling distribution of the means to become normal.

------------------------------------------------------------------------

```{r}
#| code-fold: true
# Load the required libraries
library(ggplot2)
library(patchwork)

# --- 1. Create a Skewed Population ---
# We'll use an exponential distribution, which is very non-normal.
population_size <- 100000
population <- data.frame(value = rexp(population_size, rate = 0.5))

# --- 2. Set Simulation Parameters ---
sample_size <- 40       # The size of each sample we draw from the population
n_simulations <- 5000  # The number of samples we will take

# Create a vector to store the mean of each sample
sample_means <- numeric(n_simulations)

# --- 3. Run the Simulation ---
for (i in 1:n_simulations) {
  # Draw a random sample from the population
  my_sample <- sample(population$value, size = sample_size)
  
  # Calculate the mean of that sample and store it
  sample_means[i] <- mean(my_sample)
}

# Create a data frame of the results
sampling_dist_data <- data.frame(mean = sample_means)

# --- 4. Visualize the Results ---

# Plot 1: The original, skewed population distribution
plot_population <- ggplot(population, aes(x = value)) +
  geom_density(fill = "lightcoral", alpha = 0.8) +
  labs(title = "1. Population Distribution",
       subtitle = "Highly skewed (Exponential)",
       x = "Value", y = "Density") +
  theme_minimal(base_size = 14)

# Plot 2: The distribution of sample means 
plot_sampling_dist <- ggplot(sampling_dist_data, aes(x = mean)) +
  geom_density(fill = "skyblue", alpha = 0.8) +
  # Overlay a perfect normal curve for comparison
  stat_function(fun = dnorm, args = list(mean = mean(sample_means), sd = sd(sample_means)),
                color = "red", size = 1) +
  labs(title = "2. Distribution of Means from repeated samples",
       x = "Sample Mean", y = "Density") +
  theme_minimal(base_size = 14)

# Combine the plots side-by-side
plot_population + plot_sampling_dist
```

## Sampling distribution

-   Based on central limit theorem we know that the shape will be normal.

-   We know the area under the curve and other properties about the normal based on PDF

-   However, we dont have time and resources to sample again and again, so how does this help us?

## Sampling mean = population mean 

- for a mean statistic 
$$\begin{equation}
\bar{x} = \frac{(x_1 + x_2 + ... x_N)}{N} \\
E(\bar{x}) = E\bigg(\frac{(x_1 + x_2 + ... x_N)}{N}\bigg) \\
E(\bar{x}) = \frac{E(x_1) + E(x_2) + ... E(x_N)}{E(N)} \\
E(\bar{x}) = \frac{\mu + \mu + ...\mu}{N} \\
E(\bar{x}) = \frac{N\mu}{N} \\
E(\bar{x}) = \mu
\end{equation}$$

- Similar proofs for other statistics

## Sampling mean = population mean

- What does this mean? (pun intended) 

- That we can assume our statistic is an unbiased best guess. We then want to ask our self "How good is that guess?" Or "How confident are we in this guess"

## sampling SD ≠ population SD

-   The dispersion parameter is what we are going to be focused on the most rather than the mean.

-   The SD of the sampling distribution is the standard error of the statistic (mean, b, r, SD, etc).

-   The width of the sampling distribution depends on the population variance but also the sample size!

-   We have an estimate of population variance (and any other population parameters) and we defined sample size!

------------------------------------------------------------------------

```{r}
#| code-fold: true
library(dplyr)

# --- Step 1: Create a "True" Population ---

true_population_mean <- 20
true_population_sd <- 5
population_size <- 20000

# Create the population data (a skewed gamma distribution)
set.seed(456) # for reproducibility
population_data <- rgamma(population_size, shape = (true_population_mean^2)/(true_population_sd^2), scale = (true_population_sd^2)/true_population_mean)


# --- Step 2: Simulate Sampling for DIFFERENT Sample Sizes ---

n_simulations <- 5000 # We'll run 5000 simulations for each sample size
sample_sizes <- c(5, 30, 100)  # The different sample sizes we want to test

# Create an empty data frame to store all of our results
results_df <- data.frame()

# Loop through each sample size we want to test
for (n in sample_sizes) {
  
  # Create a temporary vector to store the means for the current sample size
  current_means1 <- numeric(n_simulations)
  
  for (i in 1:n_simulations) {
    # Draw one random sample of size 'n'
    current_sample1 <- sample(population_data, n)
    
    # Calculate the mean of this sample and store it
    current_means1[i] <- mean(current_sample1)
  }
  
  # Add the means for this sample size to our main results dataframe
  results_df <- bind_rows(results_df, data.frame(mean = current_means1, sample_size = n))
}

# Convert sample_size to a factor for better plotting
results_df$sample_size <- as.factor(results_df$sample_size)

#  Visualize the Sampling Distributions Side-by-Side ---
sampling_dist_plot <- ggplot(results_df, aes(x = mean)) +
  # Create a separate histogram plot for each sample size
  facet_wrap(~ sample_size, scales = "free_y") + 
  geom_histogram(aes(y = ..density..), bins = 40, fill = "skyblue", color = "black") +
  geom_density(color = "red", size = 1) +
  # Add a vertical line for the true population mean on all plots
  geom_vline(xintercept = true_population_mean, color = "blue", linetype = "dashed", size = 1.2) +
  labs(
    title = "Sampling Distribution of the Mean for Different Sample Sizes",
    subtitle = "Drawn from the same skewed population. The blue line is the true population mean.",
    x = "Sample Mean",
    y = "Density"
  ) +
  theme_minimal()

print(sampling_dist_plot)
```

------------------------------------------------------------------------

-   We can see this too by sampling randomly from a known population

```{r}

sd(replicate(1000, mean(rnorm(5, 100, 15))))
sd(replicate(1000, mean(rnorm(25, 100, 15))))
sd(replicate(1000, mean(rnorm(100, 100, 15))))
sd(replicate(1000, mean(rnorm(1000, 100, 15))))

```


--------

-   Mean of sampling distribution is not systematically influenced by sample size 

```{r}

mean(replicate(1000, mean(rnorm(5, 100, 15))))
mean(replicate(1000, mean(rnorm(25, 100, 15))))
mean(replicate(1000, mean(rnorm(100, 100, 15))))
mean(replicate(1000, mean(rnorm(1000, 100, 15))))

```


## Sampling distribution standard error

-   mean $SE_m = \frac{\hat{\sigma}}{\sqrt{N}}$

-   sd $SE_s = \frac{\hat{\sigma}}{\sqrt{2(n-1)}}$

-   r $$SE_r =\sqrt{\frac{1-r^2}{n-2}}$$

-   regression $$\text{SE}(\hat{\beta}_1) = \frac{\text{RSE}}{\sqrt{\sum(X_i - \bar{X})^2}}$$


## sampling distribution of the mean

```{r}
sd(replicate(1000, mean(rnorm(25, 100, 15))))
```

```{r}
sd(rnorm(25, 100, 15))/sqrt(25)
```

```{r}
set.seed(74)
sd(rnorm(25, 100, 15))/sqrt(25)
```

-   Sometimes we are unlucky. The good thing (?) is we will never know.

## sampling distribution of regression coefficient

- From our simulation: 
```{r}
sd(all_my_slopes)
```

- From our single experiment: 
```{r}
summary(sample_model)
```

## SE implies different regression lines

```{r}
#| code-fold: true
regression_lines_plot <- ggplot(current_sample, aes(x = x, y = y)) +
  geom_point(alpha = 0.6) +
  # Add the 50 regression lines from our first 50 simulations
  geom_abline(
    data = all_my_models[1:50, ], 
    aes(intercept = intercept, slope = slope), 
    color = "gray", alpha = 0.5
  ) +
  # Add the "true" population line in blue so we can see how our samples vary around it
  geom_abline(intercept = true_intercept, slope = true_slope, color = "blue", size = 1.5) +
  labs(
    title = "Regression Lines from 50 Different Samples",
    x = "Predictor (x)",
    y = "Outcome (y)"
  ) +
  theme_minimal()

print(regression_lines_plot)
```

## Regression Band

```{r}
#| code-fold: true
ggplot(sample_data, aes(x = x, y = y))+
geom_point() +
geom_smooth(method='lm', se = TRUE) +
  theme_minimal()
```


## Using Sampling distributions

- Sampling distributions are used to make probabilistic statements about statistics in the same way that the standard normal distribution is used to make probabilistic statements about scores.

- Can propose different values for the population parameter and ask how rare the sample statistic would be if that population parameter were true (NHST). 

- We can determine the range within which the population mean is likely to be with a particular level of confidence (ie CI).



## Inferences from data

1. Null hypothesis significance testing
2. Precision estimation of a particular parameter
3. Comparing alternative models to see which fits the data better

## NHST

Logic:

- We are interested in estimating some statistic, for which we do not know the true population parameter. 

- We are going to assume some true value (null hypothesis), to see if our statistic is likely (or not) based on this assumption. 

- We then either reject the null (and accept the alternative), or fail to reject (and then are back at square one. The theory lives to fight another day)


## Problem of induction

- Ideally we want deductive reasoning. If the premises are true, the conclusion must be true. e.g., all people are mortal...

- Learning about the world necessitates induction. Start with specific observations and move to a general conclusion. e.g., all swans are white

- We use our sample to try and make general claims about the population, but we can never observe everything. Thus we can't prove our theories are true through induction alone.

## Falsification

- Popper states science does not progress by proving theories true (verification), but by proving theories false (falsification). 

- Research Hypothesis (H₁): This is your scientific theory 
- Null Hypothesis (H₀): A testable, and falsifiable straw man
- Use deductive logic, grounded in probability theory, to challenge the null. 
- Then make a tentative inductive inference about (H₁)


## NHST

1. Assume some population value you want to reject (H₀) 
2. Collect a sample 
3. Use sample statistics to construct sampling distribution
4. Compare the sample "test" statistic to a predefined "critical" value on the sampling distribution. 
5. If test statistic > critical value, reject null (H₀)
6. If test statistic < critical value, fail to reject null (H₁)


## Null and alternative hypothesis

- (H₀) is often a specific parameter value. Typically it is zero, which is often referred to as the nil hypothesis. 
- It can be a range, but a specific (worse case) value is needed to create the sampling distribution

- (H₁) is ALL other values. Even if you have a specific Estimate in mind

## NHST in action

```{r}
#| code-fold: true
library(ggplot2)



# --- Step 1: Create a "True" Population ---


set.seed(123) # For reproducibility
population_size <- 100000

# World where the Null Hypothesis is TRUE (No difference between groups)
population_h0_control <- rnorm(population_size, mean = 100, sd = 15)
population_h0_treatment <- rnorm(population_size, mean = 100, sd = 15) # Identical to control


# --- Step 2: Build the Sampling Distribution of the COEFFICIENT ---
# We'll simulate thousands of experiments, and for each one, we will store the
# estimated regression coefficient (the slope).

n_simulations <- 5000
sample_size <- 30
# Create an empty vector to store the SLOPE from each simulated experiment
slope_dist_h0 <- numeric(n_simulations)

for (i in 1:n_simulations) {
  # Draw samples from the world where the NULL IS TRUE
  sample_control <- sample(population_h0_control, sample_size)
  sample_treatment <- sample(population_h0_treatment, sample_size)
  
  # Combine data and create a grouping variable for the linear model
  combined_data <- data.frame(
    value = c(sample_control, sample_treatment),
    group = factor(rep(c("control", "treatment"), each = sample_size), levels = c("control", "treatment"))
  )
  
  # Run a linear model
  lm_result <- lm(value ~ group, data = combined_data)
  
  # Extract and store the SLOPE for the group predictor
  # The slope is in the 2nd row, 1st column of the coefficients table
  slope_dist_h0[i] <- summary(lm_result)$coefficients[2, 1]
}


# --- Step 3: Analyze the Sampling Distribution of the Slope ---
# The standard deviation of this distribution is our "true" Standard Error.

mean_of_slopes <- mean(slope_dist_h0)
simulated_standard_error <- sd(slope_dist_h0)


# --- Step 4: Connecting to a Single Regression Summary ---
# Now, let's run just ONE experiment and see how its summary output compares to our simulation.

our_sample_control <- sample(population_h0_control, sample_size)
our_sample_treatment <- sample(population_h0_treatment, sample_size)

our_data <- data.frame(
  value = c(our_sample_control, our_sample_treatment),
  group = factor(rep(c("control", "treatment"), each = sample_size), levels = c("control", "treatment"))
)

our_lm_test <- lm(value ~ group, data = our_data)
summary_our_lm <- summary(our_lm_test)

# Extract the Standard Error from the summary output
estimated_se_from_summary <- summary_our_lm$coefficients[2, 2]

# --- Step 5: Visualize the Sampling Distribution ---
# Let's plot the distribution of the slopes we generated.

plot_data <- data.frame(slopes = slope_dist_h0)

se_plot <- ggplot(plot_data, aes(x = slopes)) +
  geom_histogram(aes(y = ..density..), bins = 40, fill = "skyblue", color = "black") +
  geom_density(color = "red", size = 1) +
  geom_vline(xintercept = 0, color = "blue", size = 1.5, linetype = "dashed") +
  labs(
    title = "Sampling Distribution of the Regression Coefficient (Slope)",
    subtitle = "The distribution of estimates from 5000 experiments. The blue line is the true slope (0).",
    x = "Estimated Slope (Difference between group means)",
    y = "Density"
  ) +
  annotate("text", x = 0.5, y = 0.05, label = "True Slope = 0", color = "blue", fontface = "bold", hjust = 0) +
  theme_minimal()

print(se_plot)

```



## Sampling distribution regression estimate

- Centered around the null. 

- "The standard deviation of the sampling distribution is the standard error of the statistic"

```{r}
round(simulated_standard_error, 4)
```

## Our sample statistics

```{r}
summary(our_lm_test)
```

## NHST in action

- I want to see if our statistic (-1.529) is unlikley or likely assuming the null. 


```{r}
#| code-fold: true
ggplot(plot_data, aes(x = slopes)) +
  geom_histogram(aes(y = ..density..), bins = 40, fill = "skyblue", color = "black") +
  geom_density(color = "red", size = 1) +
  geom_vline(xintercept = 0, color = "blue", size = 1.5, linetype = "dashed") +
  geom_segment(aes(x = -1.529, xend = -1.529, y = 0, yend = .1), size = 1.5, linetype = "dashed") + 
  labs(
    title = "Sampling Distribution of the Regression Coefficient (Slope)",
    subtitle = "The distribution of estimates from 5000 experiments. The blue line is the null slope (0).",
    x = "Estimated Slope (Difference between group means)",
    y = "Density"
  ) +
  annotate("text", x = -8, y = 0.09, label = "Observed slope", color = "black", fontface = "bold", hjust = 0) +
  theme_minimal()
```

## NHST in action

- Next we need to define critical values. 

- Can be anywhere. Convention is to define extreme values as happening less than .05 percent of time. 

- Since extreme can be both above and below the mean, we set area under curve at 2.5% on each tail. 

```{r}
alpha <- 0.05
qnorm(1 - alpha/2,  0, 3.883)
qnorm(alpha/2,  0, 3.883)
```

---------------

```{r}
#| code-fold: true
# ---  Define the Rejection Region on the ESTIMATE's Scale ---
alpha <- 0.05
degrees_freedom <- 100
standard_error <- 3.883
# First, find the critical t-values
critical_t_upper <- qt(1 - alpha/2, df = degrees_freedom)
critical_t_lower <- qt(alpha/2, df = degrees_freedom)

# Now, convert these t-values back to the scale of your original estimate
critical_value_upper <- critical_t_upper * standard_error
critical_value_lower <- critical_t_lower * standard_error


 ggplot(plot_data, aes(x = slopes)) +
  geom_histogram(aes(y = ..density..), bins = 40, fill = "skyblue", color = "black") +
  geom_density(color = "red", size = 1) +
  geom_vline(xintercept = 0, color = "blue", size = 1.5, linetype = "dashed") +
  geom_segment(aes(x = -1.529, xend = -1.529, y = 0, yend = .1), size = 1.5, linetype = "dashed") + 
  labs(
    title = "Sampling Distribution of the Regression Coefficient (Slope)",
    subtitle = "The distribution of estimates from 5000 experiments. The blue line is the null slope (0).",
    x = "Estimated Slope (Difference between group means)",
    y = "Density"
  ) +
  annotate("text", x = -7.9, y = 0.09, label = "Observed slope", color = "black", fontface = "bold", hjust = 0) +
  geom_rect(aes(xmin = critical_value_upper, xmax = Inf, ymin = 0, ymax = Inf), fill = "grey", alpha = 0.1) +
  geom_rect(aes(xmin = -Inf, xmax = critical_value_lower, ymin = 0, ymax = Inf), fill = "grey", alpha = 0.1) +
  annotate("text", x = critical_value_upper + 2, y = 0.05, label = "Rejection\nRegion", color = "red", fontface = "bold") +
  annotate("text", x = critical_value_lower - 2, y = 0.05, label = "Rejection\nRegion", color = "red", fontface = "bold") 


```

## Decision

- Given that our statistic was not in the rejection region, we fail to reject the null. 

- Our value was consistent with a process that occurred assuming no effect. 

- This doesn't meant the null is true, just that the statistic from our sample was indistinguishable from a sampling process from a null distribution. 

## NHST in action

- Rather than work with our sampling distribution in original units, we are going to translate this distribution into a standardized form. We do this by dividing by the standard error (which is a standard deviation). 

- The information is the same, this is just convention. Because we have to estimate the SE, we use the t distribution

$$\frac{(\hat{\beta} - H_0)} {\hat{\sigma}}$$

## Critical values 


```{r}
#| code-fold: true
critical_t_upper <- qt(1 - alpha/2, df = 100)
critical_t_lower <- qt(alpha/2, df = 100)

# Now, convert these t-values back to the scale of your original estimate
critical_o.unit_upper <- critical_t_upper * standard_error
critical_o.unit_lower <- critical_t_lower * standard_error

critical_t_lower 
critical_o.unit_lower 
```

## Sampling distribution of the test statistic 

```{r}
#| code-fold: true
library(ggplot2)

set.seed(123) # For reproducibility
population_size <- 100000

# World 1: The Null Hypothesis is TRUE (No difference between groups)
population_h0_control <- rnorm(population_size, mean = 100, sd = 15)
population_h0_treatment <- rnorm(population_size, mean = 100, sd = 15) # Identical to control

# World 2: The Alternative Hypothesis is TRUE (A real difference exists)
population_h1_control <- rnorm(population_size, mean = 100, sd = 15)
population_h1_treatment <- rnorm(population_size, mean = 105, sd = 15) # Mean is 5 points higher


# --- Step 2: Build the Sampling Distribution ASSUMING THE NULL IS TRUE ---

n_simulations <- 5000
sample_size <- 30
# Create an empty vector to store the t-statistic from each simulated experiment
t_dist_h0 <- numeric(n_simulations)

for (i in 1:n_simulations) {
  # Draw samples from the world where the NULL IS TRUE
  sample_control <- sample(population_h0_control, sample_size)
  sample_treatment <- sample(population_h0_treatment, sample_size)
  
  # Combine data and create a grouping variable for the linear model
  combined_data <- data.frame(
    value = c(sample_control, sample_treatment),
    group = factor(rep(c("control", "treatment"), each = sample_size))
  )
  
  # Run a linear model 
  lm_result <- lm(value ~ group, data = combined_data)
  
  # Extract and store the t-statistic for the group predictor
  # The t-statistic is in the 2nd row, 3rd column of the coefficients table
  t_dist_h0[i] <- summary(lm_result)$coefficients[2, 3]
}


# --- Step 3: Define Our Rejection Region ---
# We set our standard of proof, alpha (α). This defines our "zone of surprise."
# If a result falls in this zone, it's too unlikely to have happened by chance.

alpha <- 0.05
# Find the critical t-values that mark the boundaries of our rejection region
degrees_freedom <- (sample_size - 1) * 2
critical_value_upper <- qt(1 - alpha/2, df = degrees_freedom)
critical_value_lower <- qt(alpha/2, df = degrees_freedom)


# --- Step 4: Conduct "Our" Single, Real Experiment ---
our_sample_control <- sample(population_h1_control, sample_size)
our_sample_treatment <- sample(population_h1_treatment, sample_size)

# Combine our single experiment's data for the linear model
our_data <- data.frame(
  value = c(our_sample_control, our_sample_treatment),
  group = factor(rep(c("control", "treatment"), each = sample_size))
)


# Extract the t-statistic and p-value for the group predictor
our_observed_t <- summary_our_lm$coefficients[2, 3]
p_value <- summary_our_lm$coefficients[2, 4]

# --- Step 5: Visualize the Verdict ---
# Let's plot the sampling distribution (from Step 2) and see where our
# single result (from Step 4) lands.

# Create a dataframe for the plot
plot_data <- data.frame(t_values = t_dist_h0)

ggplot(plot_data, aes(x = t_values)) +
  # Create the histogram of the sampling distribution under the null
  geom_histogram(aes(y = ..density..), bins = 40, fill = "skyblue", color = "black") +
  stat_function(fun = dt, args = list(df = degrees_freedom), color = "blue", size = 1) 

```


------------

- Don't need to work with the simulated sampling distribution, as we just need sample statistics to estimate.

$$\text{SE}(\hat{\beta}_1) = \frac{\text{RSE}}{\sqrt{\sum(X_i - \bar{X})^2}}$$

```{r}
#| code-fold: true
ggplot(data.frame(x = c(-4, 4)), aes(x = x)) +
  stat_function(fun = dt, args = list(df = 50)) 
```


## Our sample statistics

- Different from above, like we ran a new experiment

```{r}
summary(our_lm_test)
```


## Test statistic

```{r}
#| code-fold: true
ggplot(plot_data, aes(x = t_values)) +
    stat_function(fun = dt, args = list(df = degrees_freedom))  + geom_segment(aes(x = our_observed_t, xend = our_observed_t, y = 0, yend = .4), size = 1.5, linetype = "dashed") +

  labs(title = "The null distribution",
    x = "t-statistic",
    y = "Density"
  ) +  annotate("text", x = our_observed_t, y = -0.035, label = paste("Our observed test t-statistic\n(t =", round(our_observed_t, 2), ")"), color = "black", fontface = "bold") +
  theme_minimal()


```

## Critical values and Rejection regions

- Convention is two look at more extreme regardless of direction, a two tailed test. 

- Convention is to set it at .05 AUC. This is your *alpha* level. 

- A "one tail" test does the same, such that you specify your direction prior to analzying. Your critical value is closer to the null because now 5% AUC is on one side. 

---------------------

```{r}
#| code-fold: true
ggplot(plot_data, aes(x = t_values)) +
    stat_function(fun = dt, args = list(df = degrees_freedom))  + geom_segment(aes(x = our_observed_t, xend = our_observed_t, y = 0, yend = .4), size = 1.5, linetype = "dashed") +
  annotate("rect", xmin=-Inf, xmax=critical_value_lower, ymin=0, ymax=Inf, alpha=0.2, fill="red") + annotate("rect", xmin=critical_value_upper, xmax=Inf, ymin=0, ymax=Inf, alpha=0.2, fill="red") 
  

```
## NHST in action

- Again, our test statistic is not more extreme than our critical value so we fail to reject the null. 

- The logic of rejecting or not rejecting should be it, as in nothing else is needed given it is a decision rule. 

- Unfortunately, current practice awkwardly combines two differnt views on NHST: Fisherian and Neymann-Pearson. 

- This results in people wanting to also interpret p-values. 

## p-values

- Fisher only cared about p-values and the null. There were no alternative hypotheses or decisions to be made. Provided a strength against the null. 

P-value: 

$$p(\text{data} | H_0)$$


"The probability of getting the sample value or more extreme, assuming the null hypothesis is true"

-----------------

```{r}
#| code-fold: true

ggplot(data.frame(x = c(-4, 4)), aes(x = x)) +
annotate("rect", xmin=-Inf, xmax=our_observed_t, ymin=0, ymax=Inf, alpha=0.2, fill="red") + annotate("rect", xmin=abs(our_observed_t), xmax=Inf, ymin=0, ymax=Inf, alpha=0.2, fill="red") +
  
  # Draw the t-distribution curve
  stat_function(fun = dt, args = list(df = degrees_freedom), color = "blue", size = 1) +
  
  # Add a vertical line to mark our specific t-statistic
  geom_vline(xintercept = our_observed_t, linetype = "dashed", size = 1)+
  annotate("text", x = -3, y = 0.15, label = "This area under the curve...", color = "black") +
  annotate("text", x = 3, y = 0.15, label = "...plus this area...", color = "black") +
  annotate("text", x = 0.2, y = 0.12, label = paste("...equals the p-value\n(p =", round(p_value, 2), ")"), color = "black", fontface = "bold", size = 5) +
  
  # Add labels and title
  labs(
    title = "Conceptualizing a P-Value",
    x = "t-statistic",
    y = "Density"
  ) +
  theme_minimal()
```

---------


```{r}
pt(-0.394, df = 100) + (1 - pt(0.394, df = degrees_freedom)) 
```

- What would 1 sided p value be if 2 sided is .7? 

```{r}
pt(-0.394, df = 100) 
```

## the p-value is NOT

- The probability that the null hypothesis is true p(Null |data) . This is the "inverse probability fallacy" 

- The probability of replicating your results. A p-value of .04 does not mean there is a 96% chance the effect is real and will replicate.

- The probability that you are wrong. A small p-value does not mean your conclusion has a low chance of being a fluke (prosecutors fallacy).


## Problems with p-values

- Doesn't tell you what you want to know, which leads to misunderstanding

"What's wrong with [NHST]? Well, among many other things, it does not tell us what we want to know, and we so much want to know what we want to know that, out of desperation, we nevertheless believe that it does!" – Cohen (1994)

- p(data | H0) when we really want p(H1 | data)


## Problems with p-values

- Awkward combo of two ideas. Fisher's continuous p-value with Neyman and Pearson NHST decision-making procedure. You set an alpha level before the experiment and make a binary "reject/fail to reject" decision. The exact p-value is irrelevant; only whether it crosses your critical value, based on alpha threshold, matters.

-  But people still try to make continuous by interperting p-values out of more ***. "More significant" is oxymoronic as it is either true or not based on priori decisions. 


## Problems with p-values

- Focus on p-values leads to weird incentives

- p < .05 as a condition for publication, publication as a condition for job/tenure/grant$, which leads to gaming the system given it all hinges on this one value. Makes this all you care about at the expense of other things. 

- Leads to p-hacking, file drawers, etc (more later)

## Problems with p-values

- Leads to dichotomous thinking. "The world is flat p < .05";  "Surely god loves .051 as much as .049"

- Fail vs not fail when the world is much more grey. Both in terms of original hypotheses (nil isn't risky) as well as interpreting results. 

- Lacks a measure of interpretable effect size and an estimate of precision 

## Problems with p-values

- Leads to uncritical thinking

- "I wouldn't trust a study with larger sample size than is standard (in the field), as that suggests they just gamed the system". - WashU big wig 

- As p is a function of sample size, with Large N, anything is significant and with low N, anything is insignificant. 

- Not reporting necessary information as long as p-value is there, ritualistic p-value corrections without thought, power estimates without thought, asterisk  

## Problems with p-values

- Assuming no effect, we are still wrong at best 1/20 times. Likely much more (low power, p-hacking, etc) in published lit. 

- Assuming an effect exists, do our results tell us anything? Are they risky hypotheses at all? Instead low value, low precision (“differences” are enough), almost no non-nil null hypotheses. 

## p-values and NHST

... are intrinsically tied. 

- can be used effectively. Just many pitfalls. 

- But there are other ways! 

## Estimation

Logic: 

- Sampling distribution provides us with information about how much we can trust the parameter estimate. 

- We can quantify this precision via a confidence interval (CI), created from the sampling distribution.

- Rather than focus on null/nil, focus on effect size estimate. 

- Can also do NHST, but with a focus on gauging uncertainty

## Confidence intervals

- The sampling distribution dispersion, represented by the SE, reflects uncertainty in the sample statistic as an estimate of the population parameter

- We can create an interval that tells us how "confident" we are in the population parameter being in the range.

- We can do this because we know that the sampling distribution is normal and...

----------

- Because if a sample mean is close to the true mean, then logically, the true mean must also be close to that sample mean. If Josh (sample) is standing close to Ben (pop), then we know Ben (pop) is standing close to Josh (sample).

- Based on our simulation, we know 95% of our samples will be within 2 SDs of our true population parameter. Or that 95% of our samples will be "close" to the true parameter. 

- We know that 95% of times our sample statistic is going to be close to the population parameter. Thus we know 95% of times our population parameter will be close to the sample statistic. 


--------

$$CI_{95} = \text{statistic} \pm (1.96 * SE)$$

$$CI_{95} = \bar{X} \pm [1.96*\frac{\hat{\sigma}}{\sqrt{N}}]$$
$$CI_{95} = \hat{\beta_{1}} \pm (1.96*\frac{\text{RSE}}{\sqrt{\sum(X_i - \bar{X})^2}})$$

-----------

- The 95% does not mean there is a 95% probability that the true mean is in your specific interval. We could be unlucky. 

- It means that if you were to repeat your study many times, you would expect 95% of the confidence intervals you calculate to contain the true mean. 

-------------

```{r}
#| code-fold: true
library(ggplot2)

# --- Step 1: Define Our "True" Population ---
# In the real world, we never know the true population mean. But in a simulation,
# we can define it. This is the target our confidence intervals will try to capture.

true_population_mean <- 100
population_sd <- 15
population_size <- 20000

# Create the entire population
set.seed(123) # for reproducibility
population_data <- rnorm(population_size, mean = true_population_mean, sd = population_sd)


# --- Step 2: Simulate Thousands of Experiments ---
# We will repeatedly take small samples and calculate a 95% CI for each one.

n_simulations <- 10000
sample_size <- 30

# Create a dataframe to store the results of each simulation
simulation_results <- data.frame(
  simulation_id = 1:n_simulations,
  lower_bound = numeric(n_simulations),
  upper_bound = numeric(n_simulations),
  captured_true_mean = logical(n_simulations)
)

for (i in 1:n_simulations) {
  # 1. Draw one random sample from our population
  current_sample <- sample(population_data, sample_size)
  
  # 2. Calculate the sample's statistics
  sample_mean <- mean(current_sample)
  sample_sd <- sd(current_sample)
  standard_error <- sample_sd / sqrt(sample_size)
  degrees_freedom <- sample_size - 1
  
  # 3. Find the critical t-value for a 95% CI
  critical_t <- qt(0.975, df = degrees_freedom)
  
  # 4. Calculate the confidence interval for this one sample
  margin_of_error <- critical_t * standard_error
  simulation_results$lower_bound[i] <- sample_mean - margin_of_error
  simulation_results$upper_bound[i] <- sample_mean + margin_of_error
  
  # 5. Check if this interval "captured" the true population mean
  simulation_results$captured_true_mean[i] <- 
    (simulation_results$lower_bound[i] < true_population_mean) && 
    (simulation_results$upper_bound[i] > true_population_mean)
}


# --- Step 3: Calculate the Final Success Rate ---
# Let's see what percentage of our 10,000 CIs actually contained the true mean.

capture_rate <- mean(simulation_results$captured_true_mean) * 100


# --- Step 4: Visualize the First 100 Confidence Intervals ---
# This plot provides an intuitive look at what "95% confidence" means.

# We'll just plot the first 100 simulations so we can see them clearly
plot_data <- simulation_results[1:4, ]

ggplot(plot_data, aes(x = simulation_id, ymin = lower_bound, ymax = upper_bound, color = captured_true_mean)) +
  # Draw a vertical line for each confidence interval
  geom_errorbar(width = 0.5) +
  
  # Draw a horizontal line representing the true population mean
  geom_hline(yintercept = true_population_mean, color = "black", linetype = "dashed", size = 1) +
  
  # Manually set the colors (green for capture, red for miss)
  scale_color_manual(values = c("TRUE" = "gray50", "FALSE" = "red")) +
  
  # Add labels and a title
  labs(
    title = "Visualizing 95% Confidence Intervals",
    subtitle = "Each vertical line is one 95% CI from a different sample. The dashed line is the true population mean.",
    x = "Simulation Number (First 4 Shown)",
    y = "Value"
  ) +
  theme_minimal() +
  theme(legend.position = "none") + # Hide the legend for a cleaner look
  annotate("text", x = 80, y = true_population_mean + 1, label = "True Population Mean", color = "black", fontface = "bold") 

```

-------------


```{r}
#| code-fold: true
plot_data <- simulation_results[1:100, ]

ggplot(plot_data, aes(x = simulation_id, ymin = lower_bound, ymax = upper_bound, color = captured_true_mean)) +
  # Draw a vertical line for each confidence interval
  geom_errorbar(width = 0.5) +
  
  # Draw a horizontal line representing the true population mean
  geom_hline(yintercept = true_population_mean, color = "black", linetype = "dashed", size = 1) +
  
  # Manually set the colors (green for capture, red for miss)
  scale_color_manual(values = c("TRUE" = "gray50", "FALSE" = "red")) +
  
  # Add labels and a title
  labs(
    title = "Visualizing 95% Confidence Intervals",
    subtitle = "Each vertical line is one 95% CI from a different sample. The dashed line is the true population mean.",
    x = "Simulation Number (First 100 Shown)",
    y = "Value"
  ) +
  theme_minimal() +
  theme(legend.position = "none") + # Hide the legend for a cleaner look
  annotate("text", x = 80, y = true_population_mean + 1, label = "True Population Mean", color = "black", fontface = "bold") 
```

-------------------

- Are all CIs the same? Why or why not? 



------------

```{r}
#Number of Simulated Experiments
n_simulations

#Percentage of CIs that Captured the True Mean
 round(capture_rate, 2)
```
95% CI: "If we repeated our experiment an infinite amount of times, 95% of CIs will contain the true population parameter"

## CIs

- Can make the range any level you want, just have to know the critical value under your theoretical probability distribution

$$CI_{95} = \text{statistic} \pm (\text{crit value}* SE)$$
p <.01 or 99% CI
```{r}
qnorm(.01/2)
```
89% CI
```{r}
qnorm(.11/2)
```


## CI NHST

- Do not (necessarily) require a null hypothesis. We can focus on our "best guess" and the precision around it. 

- "One weird trick to significance. PhDs hate it."

- The standard error is treated as a fixed property of the sample, regardless of what the true mean is

------------

```{r}
#| code-fold: true
library(ggplot2)

# --- Step 1: Define a Hypothetical "Significant" Result ---
# Let's imagine we ran an experiment and got these results.
# We chose an estimate that is more than 2 standard errors away from 0,
# so we know it should be statistically significant.
observed_estimate <- 5.0
standard_error <- 2.0
degrees_freedom <- 50 # A reasonable sample size

# --- Step 2: Calculate the NHST and CI Components ---
null_value <- 0
alpha <- 0.05

# --- NHST Components ---
# The t-statistic standardizes our result: How many SEs from the null?
t_statistic <- (observed_estimate - null_value) / standard_error
p_value <- 2 * pt(abs(t_statistic), df = degrees_freedom, lower.tail = FALSE)

# Find the critical t-values that define the rejection region
critical_t <- qt(1 - alpha/2, df = degrees_freedom)

# --- Confidence Interval Components ---
margin_of_error <- critical_t * standard_error
ci_lower <- observed_estimate - margin_of_error
ci_upper <- observed_estimate + margin_of_error

# --- Plot 1: The NHST Perspective (on the t-statistic scale) ---
# This plot answers: "Does our t-statistic fall in the rejection region?"

nhst_plot <- ggplot(data.frame(x = c(-4, 4)), aes(x = x)) +
  # Shade the rejection regions
  geom_rect(aes(xmin = critical_t, xmax = Inf, ymin = 0, ymax = Inf), fill = "red", alpha = 0.3) +
  geom_rect(aes(xmin = -Inf, xmax = -critical_t, ymin = 0, ymax = Inf), fill = "red", alpha = 0.3) +
  
  # Draw the t-distribution (the "world of no effect")
  stat_function(fun = dt, args = list(df = degrees_freedom), color = "blue", size = 1) +
  
  # Add a line for our observed t-statistic
  geom_vline(xintercept = t_statistic, color = "darkgreen", size = 1.5, linetype = "dashed") +
  
  labs(
    title = "Perspective 1: Null Hypothesis Test",
    subtitle = "Our t-statistic falls in the rejection region, so we reject the null hypothesis.",
    x = "t-statistic",
    y = "Density"
  ) +
  annotate("text", x = t_statistic + 0.1, y = 0.2, label = paste("Our t-statistic\n(t =", round(t_statistic, 2), ")"), color = "darkgreen", fontface = "bold", hjust = 0) +
  theme_minimal()

print(nhst_plot)


```

-------------

Remember, for NHST we convert into a standardized form (usually t if we are estimating s). For CIs we are working with original units (as it has more meaning). 

```{r}

round(t_statistic, 2)
round(p_value, 3)
round(ci_lower, 2)
round(ci_upper, 2)
```




------------

```{r}
#| code-fold: true
# --- Plot 2: The Confidence Interval Perspective (on the estimate's scale) ---

ci_plot <- ggplot(data.frame(x = c(-10, 15)), aes(x = x)) +
  # Draw the sampling distribution if the null were true (centered at 0)
  stat_function(fun = dnorm, args = list(mean = null_value, sd = standard_error), color = "blue", size = 1) +

  # Add a vertical line for the null hypothesis value
  geom_vline(xintercept = null_value, color = "red", size = 1.5, linetype = "dashed") +
  
  # Add our observed estimate as a point
  geom_point(aes(x = observed_estimate, y = 0), color = "darkgreen", size = 4) +
  
  # Add the 95% Confidence Interval as an error bar
  geom_errorbarh(aes(xmin = ci_lower, xmax = ci_upper, y = 0), height = 0.01, color = "darkgreen", size = 1.5) +
  
  labs(
    title = "Perspective 2: Confidence Interval Estimation",
    subtitle = "Our 95% CI does not overlap with the null value (0), so the result is significant.",
    x = "Value of the Estimate",
    y = "Density"
  ) +
  annotate("text", x = null_value - 0.5, y = 0.1, label = "Null Value (0)", color = "red", fontface = "bold", hjust = 1) +
  annotate("text", x = observed_estimate, y = 0.015, label = "Our Estimate", color = "darkgreen", fontface = "bold") +
  theme_minimal()

print(ci_plot)

```



------------

```{r}
#| code-fold: true
# --- Plot 3: The CI as the "Zone of Plausible Results" ---


ci_logic_plot <- ggplot(data.frame(x = c(-10, 15)), aes(x = x)) +
  # Shade the area of the confidence interval
  geom_rect(aes(xmin = ci_lower, xmax = ci_upper, ymin = 0, ymax = Inf), fill = "gray", alpha = 0.5) +
  
  # Draw the sampling distribution centered on OUR estimate
  stat_function(fun = dnorm, args = list(mean = observed_estimate, sd = standard_error), color = "darkgreen", size = 1) +
  
  # Add the null distribution for comparison
  stat_function(fun = dnorm, args = list(mean = null_value, sd = standard_error), color = "blue", linetype = "dotted", size = 1) +
  
  # Add a vertical line for our observed estimate
  geom_vline(xintercept = observed_estimate, color = "darkgreen", size = 1.5, linetype = "dashed") +
  
  # Add the 95% Confidence Interval as an error bar at the bottom
  geom_errorbarh(aes(xmin = ci_lower, xmax = ci_upper, y = 0), height = 0.01, color = "black", size = 1.5) +
  
  labs(
    title = "The null distribution has same shape as sampling centered around estimate",
    x = "Value of the Estimate",
    y = "Density"
  ) +
  annotate("text", x = observed_estimate + 0.2, y = 0.1, label = "Distribution centered\non our estimate", color = "darkgreen", fontface = "bold", hjust = 0) +
  annotate("text", x = observed_estimate, y = -0.015, label = "The 95% Confidence Interval", color = "black", fontface = "bold") +
  annotate("text", x = null_value - 0.2, y = 0.1, label = "Null\nDistribution", color = "blue", fontface = "bold", hjust = 1) +
  theme_minimal()

print(ci_logic_plot)

```



## regression example

```{r}
#| code-fold: true
library(ggplot2)

# --- Step 1: Simulate a Single Experiment with a "Significant" Result ---

# Define the "true" population model
true_intercept <- 10
true_slope <- 0.5 # A positive relationship between x and y
true_error_sd <- 15
sample_size <- 50

# Generate one random sample based on this model
set.seed(123) # for reproducibility
sample_data <- data.frame(
  x = runif(sample_size, 0, 50)
)
sample_data$y <- true_intercept + true_slope * sample_data$x + rnorm(sample_size, mean = 0, sd = true_error_sd)

# Fit the linear regression model to our single sample
model <- lm(y ~ x, data = sample_data)
model_summary <- summary(model)

# --- Step 2: Extract Results and Calculate NHST/CI Components ---
# We are interested in the coefficient for the predictor 'x'
observed_slope <- coef(model)[2]
standard_error <- model_summary$coefficients[2, 2]
degrees_freedom <- model_summary$df[2]

null_value <- 0 # The null hypothesis is that the true slope is 0
alpha <- 0.05

# --- NHST Components ---
t_statistic <- model_summary$coefficients[2, 3]
p_value <- model_summary$coefficients[2, 4]
critical_t <- qt(1 - alpha/2, df = degrees_freedom)

# --- Confidence Interval Components ---
ci <- confint(model, "x", level = 0.95)
ci_lower <- ci[1]
ci_upper <- ci[2]
```


```{r}
#| code-fold: true
 summary(model)
```

---------

```{r}

0.6145 + 1.96*0.1369

0.6145 - 1.96*0.1369
```

```{r}
confint(model, "x", level = 0.95)
```

-------------

```{r}
#| code-fold: true
# --- Plot 1: The NHST Perspective (on the t-statistic scale) ---

nhst_plot <- ggplot(data.frame(x = c(-4, 4)), aes(x = x)) +
  geom_rect(aes(xmin = critical_t, xmax = Inf, ymin = 0, ymax = Inf), fill = "red", alpha = 0.3) +
  geom_rect(aes(xmin = -Inf, xmax = -critical_t, ymin = 0, ymax = Inf), fill = "red", alpha = 0.3) +
  stat_function(fun = dt, args = list(df = degrees_freedom), color = "blue", size = 1) +
  geom_vline(xintercept = t_statistic, color = "darkgreen", size = 1.5, linetype = "dashed") +
  labs(
    title = "Perspective 1: Null Hypothesis Test for a Regression Slope",
    subtitle = "Our t-statistic falls in the rejection region, so we reject the null hypothesis.",
    x = "t-statistic for the Slope (β₁)",
    y = "Density"
  ) +
  annotate("text", x = 2.5, y = 0.2, label = paste("Our t-statistic\n(t =", round(t_statistic, 2), ")"), color = "darkgreen", fontface = "bold", hjust = 0) +
  theme_minimal()

print(nhst_plot)


```


------------

- What is standard error for the standardized sampling distribution? 

- How does it relate to our original unit sampling distribution?

-------------------


```{r}
# t = 4.487 
# t = est/se

4.49 * 0.1369 # mean estiamte

0.6145 / 4.49   # se
```



-------------

```{r}
#| code-fold: true
# --- Plot 2: The Confidence Interval Perspective (on the slope's scale) ---

ci_plot <- ggplot(data.frame(x = c(-1.5, 1.5)), aes(x = x)) +
  stat_function(fun = dnorm, args = list(mean = null_value, sd = standard_error), color = "blue", size = 1) +
  geom_vline(xintercept = null_value, color = "red", size = 1.5, linetype = "dashed") +
  geom_point(aes(x = observed_slope, y = 0), color = "darkgreen", size = 4) +
  geom_errorbarh(aes(xmin = ci_lower, xmax = ci_upper, y = 0), height = 0.05, color = "darkgreen", size = 1.5) +
  labs(
    title = "Confidence Interval for a Regression Slope",
    y = "Density"
  ) +
  annotate("text", x = null_value - 0.05, y = 0.5, label = "Null Value (Slope = 0)", color = "red", fontface = "bold", hjust = 1) +
  annotate("text", x = observed_slope, y = 0.1, label = "Our Estimated Slope (.614)", color = "darkgreen", fontface = "bold") + scale_x_continuous(breaks = scales::pretty_breaks(n = 10)) +
  theme_minimal() 

print(ci_plot)
```



-------------------

```{r}
#| code-fold: true
ci_logic_plot <- ggplot(data.frame(x = c(-1, 1.5)), aes(x = x)) +
  geom_rect(aes(xmin = ci_lower, xmax = ci_upper, ymin = 0, ymax = Inf), fill = "gray", alpha = 0.5) +
  stat_function(fun = dnorm, args = list(mean = observed_slope, sd = standard_error), color = "darkgreen", size = 1) +
  stat_function(fun = dnorm, args = list(mean = null_value, sd = standard_error), color = "blue", linetype = "dotted", size = 1) +
  geom_vline(xintercept = observed_slope, color = "darkgreen", size = 1.5, linetype = "dashed") +
  geom_errorbarh(aes(xmin = ci_lower, xmax = ci_upper, y = 0), height = 0.05, color = "black", size = 1.5) +
  labs(
    title = "Confidence Interval for a Regression Slope",
    x = "Value of the Slope Estimate (β₁)",
    y = "Density"
  ) +
  annotate("text", x = observed_slope + 0.02, y = 0.5, label = "Distribution centered\non our estimate", color = "darkgreen", fontface = "bold", hjust = 0) +
  annotate("text", x = observed_slope, y = -0.1, label = "The 95% Confidence Interval", color = "black", fontface = "bold") +
  annotate("text", x = null_value - 0.02, y = 0.5, label = "Null\nDistribution", color = "blue", fontface = "bold", hjust = 1) + scale_x_continuous(breaks = scales::pretty_breaks(n = 10))+
  theme_minimal()

print(ci_logic_plot)
```


## CI interpretation

1. "If we repeated this experiment, ..."
2. Plausible range of values. (est could be any of these)
3. Precision in our estimate (is our Margin of Error (MOE) excessively large)


## CIs summary

- Remember we can always get unlucky 

```{r}
#| code-fold: true
plot_data <- simulation_results[1:100, ]

ggplot(plot_data, aes(x = simulation_id, ymin = lower_bound, ymax = upper_bound, color = captured_true_mean)) +
  # Draw a vertical line for each confidence interval
  geom_errorbar(width = 0.5) +
  
  # Draw a horizontal line representing the true population mean
  geom_hline(yintercept = true_population_mean, color = "black", linetype = "dashed", size = 1) +
  
  # Manually set the colors (green for capture, red for miss)
  scale_color_manual(values = c("TRUE" = "gray50", "FALSE" = "red")) +
  
  # Add labels and a title
  labs(
    title = "Visualizing 95% Confidence Intervals",
    subtitle = "Each vertical line is one 95% CI from a different sample. The dashed line is the true population mean.",
    x = "Simulation Number (First 100 Shown)",
    y = "Value"
  ) +
  theme_minimal() +
  theme(legend.position = "none") + # Hide the legend for a cleaner look
  annotate("text", x = 80, y = true_population_mean + 1, label = "True Population Mean", color = "black", fontface = "bold") 

```

## Conveying uncertainty

```{r}
#| code-fold: true
library(ggplot2)
library(dplyr)

# --- Step 1: Simulate Data for Two Groups ---
# We'll create a dataset with a continuous outcome and a two-level factor
# to represent a simple linear model with a dummy-coded predictor.

set.seed(42) # for reproducibility
n_per_group <- 50
control_mean <- 100
treatment_mean <- 110
common_sd <- 15

# Create the data
full_data <- data.frame(
  value = c(
    rnorm(n_per_group, mean = control_mean, sd = common_sd),
    rnorm(n_per_group, mean = treatment_mean, sd = common_sd)
  ),
  group = factor(rep(c("Control", "Treatment"), each = n_per_group))
)


# --- Step 2: Calculate Summary Statistics for Each Group ---
# We need to calculate the mean and the three different interval types.
# Using dplyr makes this straightforward.

summary_stats <- full_data %>%
  group_by(group) %>%
  summarise(
    mean = mean(value),
    sd = sd(value),
    n = n(),
    se = sd / sqrt(n),
    .groups = 'drop'
  ) %>%
  mutate(
    # Calculate the 95% CI using the t-distribution
    critical_t = qt(0.975, df = n - 1),
    ci_margin = critical_t * se,
    ci_lower = mean - ci_margin,
    ci_upper = mean + ci_margin
  )


model <- lm(value ~ group, data = full_data)


# --- Step 3: Create the Visualization ---
# We will create one plot that shows the raw data and overlays the
# three different error bars, jittered side-by-side for clarity.

# Define the amount to jitter the error bars
dodge_width <- 0.2

error_plot <- ggplot() +
  
  # Layer 1: The Raw Data Points (in the background)
  # REVISED: Using as.numeric(group) for a consistent continuous x-axis
  geom_jitter(
    data = full_data, 
    aes(x = as.numeric(group), y = value),
    width = 0.15, 
    color = "gray70", 
    alpha = 0.5
  ) +
  
  # Layer 2: Standard Deviation
  geom_errorbar(
    data = summary_stats,
    aes(x = as.numeric(group) - dodge_width, ymin = mean - sd, ymax = mean + sd, color = "Standard Deviation"),
    width = 0.1, size = 1
  ) +
  geom_point(data = summary_stats, aes(x = as.numeric(group) - dodge_width, y = mean, color = "Standard Deviation"), size = 4) +

  # Layer 3: 95% Confidence Interval
  geom_errorbar(
    data = summary_stats,
    aes(x = as.numeric(group), ymin = ci_lower, ymax = ci_upper, color = "95% Confidence Interval"),
    width = 0.1, size = 1
  ) +
  geom_point(data = summary_stats, aes(x = as.numeric(group), y = mean, color = "95% Confidence Interval"), size = 4) +
  
  # Layer 4: Standard Error
  geom_errorbar(
    data = summary_stats,
    aes(x = as.numeric(group) + dodge_width, ymin = mean - se, ymax = mean + se, color = "Standard Error"),
    width = 0.1, size = 1
  ) +
  geom_point(data = summary_stats, aes(x = as.numeric(group) + dodge_width, y = mean, color = "Standard Error"), size = 4) +
  
  # --- Formatting and Labels ---
  # Use scale_color_manual to define our legend
  scale_color_manual(
    name = "Uncertainty Measure",
    values = c("Standard Deviation" = "gray50", "95% Confidence Interval" = "blue", "Standard Error" = "darkgreen")
  ) +
  
  # Use scale_x_continuous to set the labels back to the group names
  scale_x_continuous(
    breaks = 1:length(levels(summary_stats$group)),
    labels = levels(summary_stats$group)
  ) +
  
  labs(
    title = "Comparing Measures of Spread and Precision for Group Means",
    subtitle = "Showing raw data with jittered error bars for SD, SE, and 95% CI.",
    x = "Group",
    y = "Value"
  ) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "bottom")

print(error_plot)



```


--------

```{r}
summary(model)
```

## smaller effect size example

```{r}
#| code-fold: true
set.seed(42) # for reproducibility
n_per_group <- 50
control_mean <- 100
treatment_mean <- 105
common_sd <- 15

# Create the data
full_data <- data.frame(
  value = c(
    rnorm(n_per_group, mean = control_mean, sd = common_sd),
    rnorm(n_per_group, mean = treatment_mean, sd = common_sd)
  ),
  group = factor(rep(c("Control", "Treatment"), each = n_per_group))
)


# --- Step 2: Calculate Summary Statistics for Each Group ---
# We need to calculate the mean and the three different interval types.
# Using dplyr makes this straightforward.

summary_stats <- full_data %>%
  group_by(group) %>%
  summarise(
    mean = mean(value),
    sd = sd(value),
    n = n(),
    se = sd / sqrt(n),
    .groups = 'drop'
  ) %>%
  mutate(
    # Calculate the 95% CI using the t-distribution
    critical_t = qt(0.975, df = n - 1),
    ci_margin = critical_t * se,
    ci_lower = mean - ci_margin,
    ci_upper = mean + ci_margin
  )


model <- lm(value ~ group, data = full_data)


# --- Step 3: Create the Visualization ---
# We will create one plot that shows the raw data and overlays the
# three different error bars, jittered side-by-side for clarity.

# Define the amount to jitter the error bars
dodge_width <- 0.2

error_plot <- ggplot() +
  
  # Layer 1: The Raw Data Points (in the background)
  # REVISED: Using as.numeric(group) for a consistent continuous x-axis
  geom_jitter(
    data = full_data, 
    aes(x = as.numeric(group), y = value),
    width = 0.15, 
    color = "gray70", 
    alpha = 0.5
  ) +
  
  # # Layer 2: Standard Deviation
  # geom_errorbar(
  #   data = summary_stats,
  #   aes(x = as.numeric(group) - dodge_width, ymin = mean - sd, ymax = mean + sd, color = "Standard Deviation"),
  #   width = 0.1, size = 1
  # ) +
  # geom_point(data = summary_stats, aes(x = as.numeric(group) - dodge_width, y = mean, color = "Standard Deviation"), size = 4) +

  # Layer 3: 95% Confidence Interval
  geom_errorbar(
    data = summary_stats,
    aes(x = as.numeric(group), ymin = ci_lower, ymax = ci_upper, color = "95% Confidence Interval"),
    width = 0.1, size = 1
  ) +
  geom_point(data = summary_stats, aes(x = as.numeric(group), y = mean, color = "95% Confidence Interval"), size = 4) +
  
  # Layer 4: Standard Error
  geom_errorbar(
    data = summary_stats,
    aes(x = as.numeric(group) + dodge_width, ymin = mean - se, ymax = mean + se, color = "Standard Error"),
    width = 0.1, size = 1
  ) +
  geom_point(data = summary_stats, aes(x = as.numeric(group) + dodge_width, y = mean, color = "Standard Error"), size = 4) +
  
  # --- Formatting and Labels ---
  # Use scale_color_manual to define our legend
  scale_color_manual(
    name = "Uncertainty Measure",
    values = c("Standard Deviation" = "gray50", "95% Confidence Interval" = "blue", "Standard Error" = "darkgreen")
  ) +
  
  # Use scale_x_continuous to set the labels back to the group names
  scale_x_continuous(
    breaks = 1:length(levels(summary_stats$group)),
    labels = levels(summary_stats$group)
  ) +
  
  labs(
    title = "Comparing Measures of Spread and Precision for Group Means",
    subtitle = "Showing raw data with jittered error bars for SD, SE, and 95% CI.",
    x = "Group",
    y = "Value"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")

print(error_plot)
```


---------------


```{r}
summary(model)
```

-------------

```{r}
#| code-fold: true
library(dplyr)

set.seed(42) # for reproducibility
n_per_group <- 50
control_mean <- 100
treatment_mean <- 105
common_sd <- 15

# Create the data
full_data <- data.frame(
  value = c(
    rnorm(n_per_group, mean = control_mean, sd = common_sd),
    rnorm(n_per_group, mean = treatment_mean, sd = common_sd)
  ),
  group = factor(rep(c("Control", "Treatment"), each = n_per_group))
)


# --- Step 2: Calculate Summary Statistics for Each Group ---
# We only need the mean and the 95% CI for this plot.
summary_stats <- full_data %>%
  group_by(group) %>%
  summarise(
    mean = mean(value),
    n = n(),
    se = sd(value) / sqrt(n),
    .groups = 'drop'
  ) %>%
  mutate(
    critical_t = qt(0.975, df = n - 1),
    ci_margin = critical_t * se,
    ci_lower = mean - ci_margin,
    ci_upper = mean + ci_margin
  )


# --- Step 3: Create the Focused Visualization ---
# This plot shows the raw data points with the group mean and 95% CI overlaid.
# REVISED: Aesthetics are now defined locally in each geom_ layer to prevent errors.
ci_plot <- ggplot() +  # Start with an empty canvas
  
  # Layer 1: The Raw Data Points (jittered so we can see them all)
  geom_jitter(
    data = full_data,
    aes(x = group, y = value), # Aesthetics are local to this layer
    width = 0.2, 
    color = "gray70", 
    alpha = 0.6
  ) +
  
  # Layer 2: The 95% Confidence Interval
  # This is centered and not jittered.
  geom_errorbar(
    data = summary_stats,
    aes(x = group, ymin = ci_lower, ymax = ci_upper), # Local aesthetics
    width = 0.1, 
    size = 1.2, 
    color = "blue"
  ) +
  
  # Layer 3: The Group Mean
  geom_point(
    data = summary_stats,
    aes(x = group, y = mean), # Local aesthetics
    color = "blue",
    size = 5
  ) +
  
  # --- Formatting and Labels ---
  labs(
    title = "Group Means with 95% Confidence Intervals",
    subtitle = "Showing the plausible range for the true population mean for each group.",
    x = "Group",
    y = "Value"
  ) +
  theme_minimal(base_size = 14)

print(ci_plot)



```

## Overlapping CIs

- If they ~touch~ then p ~= .01

- They can overlap, a little, and still be significantly different. If one edge is near the other's est, then they will be significant different. 

- If they do not overlap, they are significantly different from one another.


## Effect sizes

- practical versus statistical significance

- "One weird trick to significance. PhDs hate it."

- std b, correlation, cohen's d, odds ratio... 

## Effect sizes and p-values

- p-value is a function of the effect size and N

- Can think of p value = f(effect size / se), with the se indexing N. 


## Effect sizes


Cohen's D:

$$\delta = \frac{\mu_1 - \mu_0}{\sigma} \approx d = \frac{\bar{X}_1-\bar{X}_2}{\hat{\sigma}_p}$$

-   A Cohen’s d of .2 is considered “small”, .5 is considered “moderate”, and .8 is considered “large”

------------------------------------------------------------------------

```{r}
#| code-fold: true
coef(model)[2]/sigma(model)
```

```{r}
#| code-fold: true
library(emmeans)
emm <- emmeans(model, "group")
eff_size(emm, sigma = sigma(model), edf = df.residual(model))
```
 
## Effect sizes

Standardized: 

- r, R, $R^2$, std. b / b* 

Unstandardized:

- b 

## Meta-analytic thinking

- Estimation can be thought of as "getting the best guess (at population parameter), given my data, and conveying the uncertainty"

- Ideally, we would do this again, either via a direct replication or some "extension." Overtime, we would learn more, getting more and more precise about the parameter we are interested in (this is what a meta-analysis does).  

- In contrast, NHST focuses on compiling yeses and nos, which makes it harder to be "cumulative"

## Meta-analytic thinking

- Even if you do NHST make sure there is enough information for people to do "estimation"

- Provide sample statistics (means, SDs), Ns, se, RSE, etc.  (You would be amazed at what people neglect)

- Easily identifiable red flag, especially as models get more complex

## Model comparison

Logic:

- We should explicitly test theories against one another rather than weak null/nil hypotheses.  

- Build more than one regression model and pit two (or more) models against one another

- Compare which fits the best. The focus can be on the overall model fit or on specific coefficients. Each depend on sigma, which is used to estimate SE.

## Model comparison

- Simplest model comparison is for a "null" or "intercept only" model vs a model with a predictor

- We evaluate our models by SSE, so lets compare the SSE between two models

$$\sum (Y_i - \bar{Y})^2 = \sum (\hat{Y_i} -\bar{Y})^2 + \sum(Y_i - \hat{Y_i})^2$$

## Revisit ANOVA table

```{r}
anova(model)
```


```{r}
15.67^2
```

-------

```{r}
summary(model)
```

-----------

```{r}
#| code-fold: true
model.0 <- lm(value ~ 1, data = full_data)
summary(model.0)
```



-------------

```{r}
anova(model.0)
```

```{r}
15.98^2
```

-----------

```{r}
anova(model.0, model)
```

## SS needs to go somewhere

$$\sum (Y_i - \bar{Y})^2 = \sum (\hat{Y_i} -\bar{Y})^2 + \sum(Y_i - \hat{Y_i})^2$$

- Null model has 25292 SS residual, while model with group has 24051. That means we took 1241 and put it into model SS. 

```{r}
anova(model)
```

## Model comparison F test

- F here is the same as the previous F and p-value. 1241/245.42

```{r}
anova(model.0, model)
```
- Can be thought of as "Improvement divided by remaining error" or a metric of how much does sigma decrease


## Model comparison

- Testing a predictor in the model (evaluating b) is the exact same as testing the overall fit of a model with the predictor vs fit a model without the predictor. Because the only thing that differs between the models is the predictor. 

- This plays out in other ways, as the F-test for the overall model (tests does this model explain any variance) is the exact same as the test of a 1 predictor model (does this explain any variance)

---------

```{r}
summary(model)
```

## Equivalent tests (for 1 predictor)

- regression coef (b/se = t), R2 F test, regression CI, model comparison with null




## Visualizing F comparison

```{r}
#| code-fold: true
library(ggplot2)

# --- Step 1: Define the Parameters for the Visualization ---

# Degrees of freedom for the F-distribution
df1 <- 1
df2 <- 98

# The alpha level for our hypothesis test
alpha <- 0.05

# The observed F test statistic from your experiment
observed_f_statistic <- 5.0567

# --- Step 2: Calculate the Critical Value ---
# This is the value that our F-statistic must exceed to be "significant".
# It's the point on the distribution where 5% of the area is to its right.
critical_value <- qf(1 - alpha, df1 = df1, df2 = df2)


# --- Step 3: Create the Plot ---

f_distribution_plot <- ggplot(data.frame(x = c(0, 6)), aes(x = x)) +
  
  # Shade the rejection region (the area to the right of the critical value)
  geom_rect(
    aes(xmin = critical_value, xmax = Inf, ymin = 0, ymax = Inf),
    fill = "red",
    alpha = 0.3
  ) +
  
  # Draw the F-distribution curve
  stat_function(
    fun = df, 
    args = list(df1 = df1, df2 = df2),
    color = "blue",
    size = 1
  ) +
  
  # Add a vertical line for the critical value
  geom_vline(
    aes(xintercept = critical_value), 
    color = "darkred", 
    linetype = "dashed", 
    size = 1
  ) +
  
  # Add a vertical line for our observed test statistic
  geom_vline(
    aes(xintercept = observed_f_statistic), 
    color = "darkgreen", 
    size = 1.5
  ) +
  
  # --- Add Labels and Annotations ---
  labs(
    title = paste0("F-Distribution (df = ", df1, ", ", df2, ")"),
    subtitle = "Comparing an Observed Test Statistic to the Critical Value",
    x = "F-statistic",
    y = "Density"
  ) +
  annotate(
    "text", x = critical_value, y = 0.2, 
    label = paste("Critical Value\n(α = 0.05)\nF =", round(critical_value, 2)), 
    color = "darkred", fontface = "bold", hjust = 1.1
  ) +
  annotate(
    "text", x = observed_f_statistic, y = 0.1, 
    label = paste("Our Test Statistic\nF =", round(observed_f_statistic, 2)), 
    color = "darkgreen", fontface = "bold", hjust = 1.1
  ) +
  theme_minimal(base_size = 14)

print(f_distribution_plot)
```

## what is ANOVA? 

- ANOVA is just a Sums of Squares table

- Looks at how much SSE / Residual SS reduces compared to another model or how much SSE / Residual SS is relative to SS regression



## Nested vs non-nested

- Often our model comparisons are nested if we can think of one as a simplified version of another

- We can think of a nested version of y ~ $b_0$ + $b_1$*X + $e_i$ by fixing the slope to = 0. 

- The resulting model,  y ~ $b_0$ + $e_i$ is our intercept only model

## Non-nested comparisons? 

- We want to ask is happiness ~ $b_0$ + $b_1$ * friends + $e_i$ or happiness ~ $b_0$ + $b_1$*money  + $e_i$

- This is non-nested data and we cannot use F-tests/ANOVA  

## OLS vs Maximum Likelihood

- When we have nested models, we can use the F-test to compare. This works for linear models with residual sum of squares are calculated as the residuals are assumed to be Gaussian normal (and thus we can calculate MSE). 

- We can do the same procedure with ML. It is called a Chi-Square Difference Test (more formally, a Likelihood Ratio Test). It is more flexible because it uses likelihoods. Tests the difference in the log-likelihoods between the simple and complex models

- Asymptotically equivalent (we can see that because F and chi-square are related; and that log-likelihood is proportional to RSE. 


## Likelihoods

-   p(data \| θ ) is what we typically work with (or p(data \| H0))

-   Is this what we want to know?

-   Perhaps more helpful for scientists would be "what is the most likely parameter value, given our data"?

-   L(θ \| data )



## Binomial likelihood

```{r}
#| code-fold: true
library(ggplot2)
# Our observed data from the experiment
n_trials <- 50
k_successes <- 30 # Number of heads
mle_p = .6

# Create a sequence of candidate p values (from 0 to 1)
# THIS IS WHAT WE ARE TRYING TO ESTIMATE
p_values <- seq(0, 1, by = 0.001)

# Calculate the likelihood for each p using dbinom()
# dbinom(x, size, prob) calculates P(X=x) for a binomial distribution
likelihoods <- dbinom(x = k_successes, size = n_trials, prob = p_values)

# Store the results in a data frame
likelihood_df <- data.frame(p = p_values, likelihood = likelihoods)

# Load the ggplot2 library for nice plots
library(ggplot2)

ggplot(likelihood_df, aes(x = p, y = likelihood)) +
  geom_line(color = "blue", size = 1) +
  # Add a vertical line at the MLE to highlight the peak
  geom_vline(xintercept = mle_p, linetype = "dashed", color = "black") +
  labs(
    title = "Binomial Likelihood Function for Coin Flips",
    subtitle = "Likelihood of p given k=30 heads in n=50 flips",
    x = "Candidate Probability of Heads (p)",
    y = "Likelihood"
  ) +
  theme_minimal()




```


## ML for linear regression

-   Default DGP will be normal/Gaussian

$Y_i | X_i \sim \mathcal{N}(\beta_0 + \beta_1 X_i, \sigma^2)$

The probability density function (PDF) for a single data point: $f(Y_i | \beta_0, \beta_1, \sigma^2) = \frac{1}{\sqrt{2\pi\sigma^2}} \exp\left(-\frac {(Y_i - (\beta_0 + \beta_1 X_i))^2}{2\sigma^2}\right)$

The likelihood function $\mathcal{L}(θ | data )$ is based on our model parameters $\theta =(\beta_0 + \beta_1, \sigma^2)$ such that we find that values of our parameters that maximize this function:

$$\mathcal{L}(\beta_0, \beta_1, \sigma^2) = \prod_{i=1}^{n} \frac{1}{\sqrt{2\pi\sigma^2}} \exp\left(-\frac{(Y_i - (\beta_0 + \beta_1 X_i))^2}{2\sigma^2}\right)$$

## Log Likelihood

-   Maximizing the log-likelihood is the same as maximizing the likelihood, but it's mathematically and computationally much easier.

-   Likelihood of observing all your data points is the product of the likelihood of each individual point. Multiplying many small probs will lead to issues. So we log them to turn it into addition.

## Log Likelihood estimating a mean

-   define what values we want to calculate log likelihood for

```{r}
#| code-fold: true
library(psychTools)
library(tidyr)
library(tidyverse)
galton.data <- galton
grid <-
  crossing(mu = seq(from = 66, to = 69, length.out = 200), sigma = seq(from = 2, to = 3, length.out = 200))
grid
```

## Log Likelihood example

-   Need to calculate the log likelihood of our data (each child’s height) assuming each part of our grid is true.

-   Take the first line of our grid (mu = 66, sigma = 2) so we can see how likely each child (all 928) by dnorm(height, 66, 2)

-   We sum the LLs across all 928 participants for each spot in the grid to get the average log-likelihood for each grid spot

------------------------------------------------------------------------

40,000 grids \* 928 participants = 37,120,000 calculations

```{r}
#| code-fold: true
library(purrr)
grid_function <- function(mu, sigma) {
  dnorm(galton.data$child, mean = mu, sd = sigma, log = T) %>%
    sum() 
  }

p_grid <-
  grid %>% 
  mutate(log_likelihood = map2(mu, sigma, grid_function)) %>%
  unnest(log_likelihood)
  p_grid
```

------------------------------------------------------------------------

```{r}
#| code-fold: true
library(viridis)
p_grid %>% 
  ggplot(aes(x = mu, y = sigma, fill = log_likelihood)) + 
  geom_raster(interpolate = T) +
  scale_fill_viridis_c(option = "C") +
  labs(title = "log likelihood for mean and sd",
         x = expression(mu),
       y = expression(sigma)) +
  theme(panel.grid = element_blank())
```

------------------------------------------------------------------------

```{r}
#| code-fold: true
library(tidybayes)
p_grid %>% 
  ggplot(aes(x = mu, y = log_likelihood )) + 
    stat_lineribbon(.width = c(0), alpha = .4) +  theme(legend.position="none")
```

------------------------------------------------------------------------

```{r}
#| code-fold: true
p_grid %>% 
  ggplot(aes(x = sigma, y = log_likelihood )) + 
    stat_lineribbon(.width = c(0), alpha = .4) +  theme(legend.position="none")
```

------------------------------------------------------------------------

```{r}
#| code-fold: true
library(report)
report_sample(galton.data$child)
```

## Key take aways

- ML is iterative, in that we try out a bunch of different values and then select the one that is *most consistent with our data*

- The likelihood is a direct consequence of our data based on a theoretical probability distribution. 

- It does not mean it is the only answer, and we can see  other values that are almost as likely


## Maximum Likelihood

-   regression example

```{r}
#| code-fold: true
# Load the ggplot2 library for plotting
library(ggplot2)

# --- 1. Generate Sample Data ---
set.seed(42) # For reproducibility
n <- 100
true_beta0 <- 10  # True intercept
true_beta1 <- 2   # True slope
true_sigma <- 5   # True standard deviation of the error

x <- 1:n
errors <- rnorm(n, mean = 0, sd = true_sigma)
y <- true_beta0 + true_beta1 * x + errors

# --- 2. Define the Log-Likelihood Function ---
# This function calculates the total log-likelihood for a given set of parameters.
# params is a vector: c(beta0, beta1, sigma)
log_likelihood <- function(params, x, y) {
  beta0 <- params[1]
  beta1 <- params[2]
  sigma <- params[3]
  
  # Ensure sigma is positive, otherwise, the density is undefined
  if (sigma <= 0) return(-Inf)

  # Calculate the predicted values for y
  y_pred <- beta0 + beta1 * x

  # Calculate the log-likelihood using the normal distribution's PDF.
  # dnorm() with log = TRUE calculates the log-probability density of each residual.
  ll <- sum(dnorm(y, mean = y_pred, sd = sigma, log = TRUE))
  
  return(ll)
}

# --- 3. Set Up and Run the Grid Search ---
# To make sure our grid is in the right area, we can center it around the OLS estimates.
ols_fit <- lm(y ~ x)

# Define a range of plausible values for each parameter
beta0_grid <- seq(coef(ols_fit)[1] - 2, coef(ols_fit)[1] + 2, length.out = 50)
beta1_grid <- seq(coef(ols_fit)[2] - 0.1, coef(ols_fit)[2] + 0.1, length.out = 50)
sigma_grid <- seq(sigma(ols_fit) - 2, sigma(ols_fit) + 2, length.out = 50)

# Create a data frame containing every combination of parameters
param_grid <- expand.grid(beta0 = beta0_grid, beta1 = beta1_grid, sigma = sigma_grid)

# Calculate the log-likelihood for each combination in the grid. This may take a moment.
log_likelihood_values <- apply(param_grid, 1, log_likelihood, x = x, y = y)
param_grid$ll <- log_likelihood_values

# Find the row in the grid with the maximum log-likelihood
mle_grid_estimate <- param_grid[which.max(param_grid$ll), ]

param_grid
```

## magic is here

```{r, eval = F}
sum(dnorm(y, mean = y_pred, sd = sigma, log = TRUE))
```

-   Our data are fixed as we collected them. We test out which ones make better predictions by trying out different parameters, which will change the y_pred. Asks: How likely is y, using different parameters L($\theta$\|d).

```{r}
#better guess! 
dnorm(20, mean = 25, sd = 3.23, log = TRUE)
```

```{r}
dnorm(20, mean = 40, sd = 3.23, log = TRUE)
```

------------------------------------------------------------------------

```{r}
print(mle_grid_estimate)
```

```{r}
summary(ols_fit)
```


------------------------------------------------------------------------

```{r}
#| code-fold: true
# Filter the grid to only include the slice with the best sigma value
surface_data <- subset(param_grid, sigma == mle_grid_estimate$sigma)

ggplot(surface_data, aes(x = beta0, y = beta1, fill = ll)) +
  geom_tile() + # Creates the heatmap
  scale_fill_viridis_c() + # A visually appealing color scale
  # Add a point to mark the maximum
  geom_point(data = mle_grid_estimate, aes(x = beta0, y = beta1), 
             color = "red", size = 4, shape = 4, stroke = 1.5) +
  labs(
    title = "Log-Likelihood Surface",
    subtitle = "Red 'X' marks the Maximum Likelihood Estimate from the grid search",
    x = "Intercept (beta0)",
    y = "Slope (beta1)",
    fill = "Log-Likelihood"
  ) +
  theme_minimal()
```

## ML regression

- ML is the default estimation algorithm using the glm (generalized linear model) function. 

- We will use this anytime we cannot assume Gaussian normal. However, we can still use it for Gaussian normal DGPs

```{r, eval = FALSE}

fit <- glm( y ~ x, family = Gaussian, data = data)
```



---------


```{r}
lmod <- glm(y~x)
summary(lmod)
```

--------

- We don't get the same information below the coefs because we are no longer doing sum of squares decomposition.

$$\sum (Y_i - \bar{Y})^2 = \sum (\hat{Y_i} -\bar{Y})^2 + \sum(Y_i - \hat{Y_i})^2$$

- SSE/RSS as a measure of fit assumes that the errors around the regression line are continous and normally distributed

-------------------------

```{r}
logLik(lmod)
```

```{r}
mle_grid_estimate
```


## example revisited

```{r}
summary(model)
```


---------------

```{r}
m.1 <- glm(value ~ group, data = full_data)
summary(m.1)
```

------------

```{r}
m.0 <- glm(value ~ 1, data = full_data)
summary(m.0)
```

## Likelihood ratio test

- For use with nested comparisons. Cannot use ANOVA (well for any other type of glm, but since this is a gaussian one, we can). 

- LRT = -2($LL_\text{Simple}$ - $LL_\text{complex}$ )

```{r}
logLik(m.1)
```

```{r}
logLik(m.0)
```

-------------

```{r}
#| code-fold: true
chi_sq_statistic <- -2 * (as.numeric(logLik(m.0))-as.numeric(logLik(m.1)))

df_diff <- length(coef(m.1)) - length(coef(m.0))

# Calculate the p-value
p_value_manual <- pchisq(chi_sq_statistic, df = df_diff, lower.tail = FALSE)

chi_sq_statistic
df_diff
p_value_manual 
```

-----------

```{r}
#| code-fold: true
library(lmtest)
lrtest(m.0, m.1)
```

----------

```{r}
#| code-fold: true
df <- 1

# The alpha level for our hypothesis test
alpha <- 0.05

# The observed Chi-Square test statistic from your LRT
observed_chi_sq <- 5.031213

# --- Step 2: Calculate the Critical Value and p-value ---
# The critical value is the point on the distribution where 5% of the area is to its right.
critical_value <- qchisq(1 - alpha, df = df)

# The p-value is the probability of getting a result as extreme or more extreme than ours.
p_value <- pchisq(observed_chi_sq, df = df, lower.tail = FALSE)


# --- Step 3: Create the Plot ---

chi_sq_plot <- ggplot(data.frame(x = c(0, 8)), aes(x = x)) +
  
  # Shade the rejection region (the area to the right of the critical value)
  geom_rect(
    aes(xmin = critical_value, xmax = Inf, ymin = 0, ymax = Inf),
    fill = "red",
    alpha = 0.3
  ) +
  
  # Draw the Chi-Square distribution curve
  stat_function(
    fun = dchisq, 
    args = list(df = df),
    color = "blue",
    size = 1
  ) +
  
  # Add a vertical line for the critical value
  geom_vline(
    aes(xintercept = critical_value), 
    color = "darkred", 
    linetype = "dashed", 
    size = 1
  ) +
  
  # Add a vertical line for our observed test statistic
  geom_vline(
    aes(xintercept = observed_chi_sq), 
    color = "darkgreen", 
    size = 1.5
  ) +
  
  # --- Add Labels and Annotations ---
  labs(
    title = expression(paste("Chi-Square (", chi^2, ") Distribution with df = ", df)),
    subtitle = "Comparing an Observed LRT Statistic to the Critical Value",
    x = expression(paste(chi^2, " Statistic")),
    y = "Density"
  ) +
  annotate(
    "text", x = critical_value, y = 0.5, 
    label = paste("Critical Value\n(α = 0.05)\nχ² =", round(critical_value, 2)), 
    color = "darkred", fontface = "bold", hjust = 1.1
  ) +
  annotate(
    "text", x = observed_chi_sq, y = 0.1, 
    label = paste("Our Test Statistic\nχ² =", round(observed_chi_sq, 2), "\np-value =", format.pval(p_value, digits=2)), 
    color = "darkgreen", fontface = "bold", hjust = 1.1
  ) +
  theme_minimal(base_size = 14)

print(chi_sq_plot)
```

-------

- Does our lm function results have likelihoods? 

```{r}
anova(model.0, model, test="LRT")
```

```{r}
anova(model.0, model)
```


----------


```{r}
logLik(model)
```

```{r}

lrtest(model.0, model)
```


---------

Anova with glm models will give you deviances

```{r}
anova(m.0, m.1)
```


## Deviance


```{r}
summary(m.1)
```


```{r}
deviance(m.1)
deviance(m.0)
```

## Deviance

- While log likelihood is a measure of goodness of fit, deviance is a measure of poor fit (similar to SS error/Residual SS). 

- Deviance = -2 * Log-Likelihood

- Why do this at all, since they are the same? Multiplying the log-likelihood by -2 puts the measure of fit onto a scale that approximates a chi-square (For most models. For gaussian we will still use F-test)

------------


$$F = \frac{\frac{\text{Deviance}_{\text{simple}} - \text{Deviance}_{\text{complex}}}{\text{df}_{\text{simple}} - \text{df}_{\text{complex}}}}{\frac{\text{Deviance}_{\text{complex}}}{\text{df}_{\text{complex}}}}$$

```{r}
(deviance(m.0)-deviance(m.1))/(deviance(m.1)/98)
```

----------

- F will be roughly equal to a chi square (df = 1) when the df for the F(1, LARGE)

Chi-square = 
```{r}
log(deviance(m.0)/deviance(m.1))*100
```


## Fit measure summary

- MSE/RSE/sigma (low is better fit), $R^2$ & R (large is better fit) 

- log likelihood (ll; higher/less negative is better fit), deviance (lower is better fit)

- All related; all are non-standardized (meaningless without more information) except for Rs

- For Gaussian we test with F, for all else with Chi-square. Each compares fit.

## Non-nested comparisons

- Nested designs can be thought of improvement from some simpler comparison. Non-nested are comparing apples and oranges. 

- We cannot directly use SSE / MSE / RSE / ll / deviance to test. Instead we will focus on "goodness of fit" and occularly, rather than formally, compare

- We will use the likelihood + a penalty based on number of predictors

- Why penalty? Because models can be more complex and thus be easier to predict some outcome. 

## Information criteria

The smaller the number, the better the fit: 

- AIC (Akaike Information Criterion): AIC = -2 * log(Likelihood) + 2p(predictors)

- BIC (Bayesian Information Criterion): BIC = -2 * log(Likelihood) + p * log(n)

- BIC has a much harsher penalty for complexity, especially with large sample sizes (n), and will favor a simpler model

## AIC/BIC


```{r}
library("performance")
model_performance(m.1)
```

```{r}
compare_performance(m.1, m.0)
```

------------

```{r}
plot(compare_performance(m.1, m.0, rank = TRUE))
```

## Aic/bic

- On their own, the numbers are meaningless. Need to directly compare to another model.

- There is no formal test, though there are some informal guidelines about magnitude of difference

- Sometimes AIC will differ from BIC on which model it prefers. 

## Summary

1. NHST (primary outcome is decision)
2. Effect size + CI (primary outcome is precision)
3. Model Comparison (primary outcome is comparison)

