---
title: "Statistical Inference"
format: revealjs
slide-number: true
editor: visual
execute:
  echo: true
html:
  code-fold: true
  code-summary: Show the code
  scrollable: true
---

## Last section

- First third of the class

1. Construction of model (make it work)
2. Inference from model (learn from it)
3. Additions to basic model (make it suit our needs)

## This time (this section)

-   Sampling distributions

-   NHST as evaluating our hypothesis

-   Estimation as alternative

-   Model comparison as third alternative

## Estimation of what?

-   We To do that we need to use an *Estimator* to get *Estimates*, which are our best guess of the population parameters.

-   The population parameters are your *Estimand* -- what you want to know but have to estimate

-   Picture of perfect cake (Estimand), Cake recipe (Estimator), your cake (Estimate)

## How good our are estimates?

-   Thus far we used the standard error to help us define the precision around the estimates... and I just gave you the equation

-   Where does this come from, and how can I think about the process?

-   Remember, we only get one shot to estimate the population.

------------------------------------------------------------------------

```{r}
#| code-fold: true
# Load the ggplot2 library for plotting
library(ggplot2)

# --- 1. Define the True Population Model ---
# These are the "true" parameters that we are trying to discover.
true_beta0 <- 10
true_beta1 <- 2
true_sigma <- 15
population_size <- 20000

# Create the entire population based on the true model
population <- data.frame(
  x = runif(population_size, 0, 50)
)
population$y <- true_beta0 + true_beta1 * population$x + rnorm(population_size, mean = 0, sd = true_sigma)

# --- 2. Draw a Random Sample from the Population ---
sample_size <- 50
sample_data <- population[sample(1:population_size, sample_size), ]

# --- 3. Fit a Regression Model to the SAMPLE ---
sample_model1 <- lm(y ~ x, data = sample_data)
sample_beta0_hat <- coef(sample_model1)[1]
sample_beta1_hat <- coef(sample_model1)[2]

# --- 4. Visualize the Relationship ---

pop <- ggplot(population, aes(x = y)) +
  geom_density(fill = "lightblue", alpha = 0.8, color = "black") +
  labs(
    title = "Distribution of the Outcome Variable (Y) in the Population",
    subtitle = "Y is determined by y = 10 + 2X + rnorm(0,15) ",
    x = "Outcome (Y)",
    y = "Density"
  ) +
  theme_minimal(base_size = 14)


pop

```

------------------------------------------------------------------------

```{r}
#| code-fold: true
ggplot(population, aes(x = x, y = y)) +
# Plot the entire population as faint gray points
geom_point(color = "gray80", alpha = 0.5) +
# Add the TRUE population regression line in red
geom_abline(
intercept = true_beta0,
slope = true_beta1,
color = "black",
size = 1.5,
linetype = "dashed"
) +
# Highlight the points from our random sample in blue
geom_point(data = sample_data, color = "blue", size = 2.5) +
# Add the ESTIMATED sample regression line in blue
geom_abline(
intercept = sample_beta0_hat,
slope = sample_beta1_hat,
color = "green",
size = 1.5
) +
labs(subtitle = "The green line is our sample estimate of the population (black)",
x = "Predictor (X)",
y = "Outcome (Y)"
) +
theme_minimal(base_size = 14)
```

## What we normally see

```{r}
#| code-fold: true
ggplot(sample_data, aes(x = x, y = y))+
geom_point() +
geom_smooth(method='lm', se = FALSE) 
```

------------------------------------------------------------------------

-   The major goal that we have in statistical inference is to make confident claims about the population based on a small representation of it, the sample. We use features of the sample (statistics) to inform us about features of the population (parameters).

-   Any sample will be off the mark in how well it captures the important features of a population due to *sampling variability*.

-   The sampling distribution tells us how far off the mark we can expect a sample statistic to be.

## Sampling distribution - empirical

-   Run 5000 experiments from a known population. Make a distribution of the 5000 regression coefs to tell us what we should expect from any 1 experiment.

```{r}
#| code-fold: true
library(ggplot2)

# --- Step 1: Create a "True" Population ---

true_intercept <- 10
true_slope <- 2
true_error_sd <- 15 # The true standard deviation of the residuals (sigma)
population_size <- 20000

# Create the entire population based on our true model
set.seed(74) # for reproducibility
population_data <- data.frame(
  x = runif(population_size, 0, 50)
)
population_data$y <- true_intercept + true_slope * population_data$x + rnorm(population_size, mean = 0, sd = true_error_sd)



# --- Step 2: Simulate the Process of Repeated Sampling ---

n_simulations <- 5000 # Let's take 5000 different samples
sample_size <- 100  # Each sample will have 10 data points
all_my_slopes <- numeric(n_simulations) # An empty vector to store our results

all_my_models <- data.frame(
  intercept = numeric(n_simulations),
  slope = numeric(n_simulations)
)

for (i in 1:n_simulations) {
  # Draw one random sample of 100 data points from our population
  current_sample <- population_data[sample(1:population_size, sample_size), ]
  
  # Fit a regression model to this one small sample
  sample_model <- lm(y ~ x, data = current_sample)
  
  # Extract and store the estimated slope from this sample's model
  all_my_models$intercept[i] <- coef(sample_model)[1]
  all_my_models$slope[i] <- coef(sample_model)[2]
}


# --- Step 3: Build and Analyze the Sampling Distribution ---
all_my_slopes <- all_my_models$slope
mean_of_slopes <- mean(all_my_slopes)


cat("The true slope is 2.0. The mean of our 5000 estimated slopes is:", round(mean_of_slopes, 4), "\n")

simulated_standard_error <- sd(all_my_slopes)
cat("The standard deviation of our slopes (the Simulated Standard Error) is:", round(simulated_standard_error, 4), "\n\n")


```

## Sampling distribution

-   For each sample, we calculate a statistic, and then create a distribution of sample statistics

```{r}
#| code-fold: true
# --- Step 5: Visualize Everything ---
# A histogram clearly shows the distribution of our estimates.
sampling_dist_plot <- ggplot(data.frame(slopes = all_my_slopes), aes(x = slopes)) +
  geom_histogram(aes(y = ..density..), bins = 30, fill = "skyblue", color = "black") +
  geom_density(color = "red", size = 1) +
  geom_vline(xintercept = true_slope, color = "blue", linetype = "dashed", size = 1.5) +
  labs(
    title = "Sampling Distribution of the Slope (β₁)",
    subtitle = paste("Based on", n_simulations, "samples of size", sample_size),
    x = "Estimated Slope Coefficient (β̂₁)",
    y = "Density"
  ) + 
  # geom_vline(2.014) + 
  theme_minimal()

print(sampling_dist_plot)

```

------------------------------------------------------------------------

-   We normally do not have access to this data, but if we did we now have a *distribution* of plausible statistics drawn from the population.

-   This distribution tells us what we would expect if we repeated the experiment again, and again, and again

- Regression coef is an unbiased estimator because we can find the true mean (if we repeat experiment infinite times)

-   Range of scores are SUPER helpful to gauge our precision of the estimate.

------------------------------------------------------------------------

-   If we look at just one sample (like normal experiments), we won't exactly estimate the population intercept (10) and slope (2).

```{r}
#| code-fold: true
# --- Step 4:Connecting to a Single Regression Summary ---
summary(sample_model) # Using the last model from our loop

```

## Sampling distribution

```{r}
#| code-fold: true
# --- Step 5: Visualize Everything ---
# A histogram clearly shows the distribution of our estimates.
sampling_dist_plot <- ggplot(data.frame(slopes = all_my_slopes), aes(x = slopes)) +
  geom_histogram(aes(y = ..density..), bins = 30, fill = "skyblue", color = "black") +
  geom_density(color = "red", size = 1) +
  geom_vline(xintercept = true_slope, color = "blue", linetype = "dashed", size = 1.5) +
  geom_vline(xintercept = 1.87082, color = "black", linetype = "solid", size = 1.5) +
  labs(
    title = "Sampling Distribution of the Slope (β₁)",
    subtitle = paste("Based on", n_simulations, "samples of size", sample_size),
    x = "Estimated Slope Coefficient (β̂₁)",
    y = "Density"
  ) + 
  # geom_vline(2.014) + 
  theme_minimal()

print(sampling_dist_plot)

```

- Did we get unlucky with our one experiment? Is this just part of the uncertainty we have to live with? 

## Sampling distribution

-   Can be for any statistic. Mean, SD, r, b, etc.

http://shiny.calpoly.sh/Sampling_Distribution/

-   If unbiased, the mean of the sampling distribution is equal to the population parameter

-   The standard deviation of the sampling distribution is called standard error, and helps us gauge our precision of the statistic.

## Sampling distribution

-   Given we don't have access to the population, nor do we have the time and resources to conduct thousands of replication studies, what do we do?

-   Our sampling distribution will be theoretical... and it will be normal.

## Central limit theorm take 2

-   Central Limit Theorem: the sum of a large number of independent random variables will be approximately normally distributed, even if original distributions are not normal.

-   We previously used this to show that the normal was everywhere. Here we are going to use it to define the shape of our sampling distribution.

-   The sample size necessary to be "approximately normal" depends on the nature of the underlying data. The less normal it is, the larger the sample size necessary in order for the sampling distribution of the means to become normal.

------------------------------------------------------------------------

```{r}
#| code-fold: true
# Load the required libraries
library(ggplot2)
library(patchwork)

# --- 1. Create a Skewed Population ---
# We'll use an exponential distribution, which is very non-normal.
population_size <- 100000
population <- data.frame(value = rexp(population_size, rate = 0.5))

# --- 2. Set Simulation Parameters ---
sample_size <- 40       # The size of each sample we draw from the population
n_simulations <- 5000  # The number of samples we will take

# Create a vector to store the mean of each sample
sample_means <- numeric(n_simulations)

# --- 3. Run the Simulation ---
for (i in 1:n_simulations) {
  # Draw a random sample from the population
  my_sample <- sample(population$value, size = sample_size)
  
  # Calculate the mean of that sample and store it
  sample_means[i] <- mean(my_sample)
}

# Create a data frame of the results
sampling_dist_data <- data.frame(mean = sample_means)

# --- 4. Visualize the Results ---

# Plot 1: The original, skewed population distribution
plot_population <- ggplot(population, aes(x = value)) +
  geom_density(fill = "lightcoral", alpha = 0.8) +
  labs(title = "1. Population Distribution",
       subtitle = "Highly skewed (Exponential)",
       x = "Value", y = "Density") +
  theme_minimal(base_size = 14)

# Plot 2: The distribution of sample means 
plot_sampling_dist <- ggplot(sampling_dist_data, aes(x = mean)) +
  geom_density(fill = "skyblue", alpha = 0.8) +
  # Overlay a perfect normal curve for comparison
  stat_function(fun = dnorm, args = list(mean = mean(sample_means), sd = sd(sample_means)),
                color = "red", size = 1) +
  labs(title = "2. Distribution of Means from repeated samples",
       x = "Sample Mean", y = "Density") +
  theme_minimal(base_size = 14)

# Combine the plots side-by-side
plot_population + plot_sampling_dist
```

## Sampling distribution

-   Based on central limit theorem we know that the shape will be normal.

-   We know the area under the curve and other properties about the normal based on PDF

-   However, we dont have time and resources to sample again and again, so how does this help us?

## Sampling mean = population mean 

- for a mean statistic 
$$\begin{equation}
\bar{x} = \frac{(x_1 + x_2 + ... x_N)}{N} \\
E(\bar{x}) = E\bigg(\frac{(x_1 + x_2 + ... x_N)}{N}\bigg) \\
E(\bar{x}) = \frac{E(x_1) + E(x_2) + ... E(x_N)}{E(N)} \\
E(\bar{x}) = \frac{\mu + \mu + ...\mu}{N} \\
E(\bar{x}) = \frac{N\mu}{N} \\
E(\bar{x}) = \mu
\end{equation}$$

- Similar proofs for other statistics

## Sampling mean = population mean

- What does this mean? (pun intended) 

- That we can assume our statistic is an unbiased best guess. We then want to ask our self "How good is that guess?" Or "How confident are we in this guess"

## sampling SD ≠ population SD

-   The dispersion parameter is what we are going to be focused on the most rather than the mean.

-   The SD of the sampling distribution is the standard error of the statistic (mean, b, r, SD, etc).

-   The width of the sampling distribution depends on the population variance but also the sample size!

-   We have an estimate of population variance (and any other population parameters) and we defined sample size!

------------------------------------------------------------------------

```{r}
#| code-fold: true
library(dplyr)

# --- Step 1: Create a "True" Population ---

true_population_mean <- 20
true_population_sd <- 5
population_size <- 20000

# Create the population data (a skewed gamma distribution)
set.seed(456) # for reproducibility
population_data <- rgamma(population_size, shape = (true_population_mean^2)/(true_population_sd^2), scale = (true_population_sd^2)/true_population_mean)


# --- Step 2: Simulate Sampling for DIFFERENT Sample Sizes ---

n_simulations <- 5000 # We'll run 5000 simulations for each sample size
sample_sizes <- c(5, 30, 100)  # The different sample sizes we want to test

# Create an empty data frame to store all of our results
results_df <- data.frame()

# Loop through each sample size we want to test
for (n in sample_sizes) {
  
  # Create a temporary vector to store the means for the current sample size
  current_means1 <- numeric(n_simulations)
  
  for (i in 1:n_simulations) {
    # Draw one random sample of size 'n'
    current_sample1 <- sample(population_data, n)
    
    # Calculate the mean of this sample and store it
    current_means1[i] <- mean(current_sample1)
  }
  
  # Add the means for this sample size to our main results dataframe
  results_df <- bind_rows(results_df, data.frame(mean = current_means1, sample_size = n))
}

# Convert sample_size to a factor for better plotting
results_df$sample_size <- as.factor(results_df$sample_size)

#  Visualize the Sampling Distributions Side-by-Side ---
sampling_dist_plot <- ggplot(results_df, aes(x = mean)) +
  # Create a separate histogram plot for each sample size
  facet_wrap(~ sample_size, scales = "free_y") + 
  geom_histogram(aes(y = ..density..), bins = 40, fill = "skyblue", color = "black") +
  geom_density(color = "red", size = 1) +
  # Add a vertical line for the true population mean on all plots
  geom_vline(xintercept = true_population_mean, color = "blue", linetype = "dashed", size = 1.2) +
  labs(
    title = "Sampling Distribution of the Mean for Different Sample Sizes",
    subtitle = "Drawn from the same skewed population. The blue line is the true population mean.",
    x = "Sample Mean",
    y = "Density"
  ) +
  theme_minimal()

print(sampling_dist_plot)
```

------------------------------------------------------------------------

-   We can see this too by sampling randomly from a known population

```{r}

sd(replicate(1000, mean(rnorm(5, 100, 15))))
sd(replicate(1000, mean(rnorm(25, 100, 15))))
sd(replicate(1000, mean(rnorm(100, 100, 15))))
sd(replicate(1000, mean(rnorm(1000, 100, 15))))

```


--------

-   Mean of sampling distribution is not systematically influenced by sample size 

```{r}

mean(replicate(1000, mean(rnorm(5, 100, 15))))
mean(replicate(1000, mean(rnorm(25, 100, 15))))
mean(replicate(1000, mean(rnorm(100, 100, 15))))
mean(replicate(1000, mean(rnorm(1000, 100, 15))))

```


## Sampling distribution standard error

-   mean $SE_m = \frac{\hat{\sigma}}{\sqrt{N}}$

-   sd $SE_s = \frac{\hat{\sigma}}{\sqrt{2(n-1)}}$

-   r $$SE_r =\sqrt{\frac{1-r^2}{n-2}}$$

-   regression $$\text{SE}(\hat{\beta}_1) = \frac{\text{RSE}}{\sqrt{\sum(X_i - \bar{X})^2}}$$


## sampling distribution of the mean

```{r}
sd(replicate(1000, mean(rnorm(25, 100, 15))))
```

```{r}
sd(rnorm(25, 100, 15))/sqrt(25)
```

```{r}
set.seed(74)
sd(rnorm(25, 100, 15))/sqrt(25)
```

-   Sometimes we are unlucky. The good thing (?) is we will never know.

## sampling distribution of regression coefficient

- From our simulation: 
```{r}
sd(all_my_slopes)
```

- From our single experiment: 
```{r}
summary(sample_model)
```

## SE implies different regression lines

```{r}
#| code-fold: true
regression_lines_plot <- ggplot(current_sample, aes(x = x, y = y)) +
  geom_point(alpha = 0.6) +
  # Add the 50 regression lines from our first 50 simulations
  geom_abline(
    data = all_my_models[1:50, ], 
    aes(intercept = intercept, slope = slope), 
    color = "gray", alpha = 0.5
  ) +
  # Add the "true" population line in blue so we can see how our samples vary around it
  geom_abline(intercept = true_intercept, slope = true_slope, color = "blue", size = 1.5) +
  labs(
    title = "Regression Lines from 50 Different Samples",
    x = "Predictor (x)",
    y = "Outcome (y)"
  ) +
  theme_minimal()

print(regression_lines_plot)
```

## Regression Band

```{r}
#| code-fold: true
ggplot(sample_data, aes(x = x, y = y))+
geom_point() +
geom_smooth(method='lm', se = TRUE) +
  theme_minimal()
```


## Using Sampling distributions

- Sampling distributions are used to make probabilistic statements about statistics in the same way that the standard normal distribution is used to make probabilistic statements about scores.

- Can propose different values for the population parameter and ask how rare the sample statistic would be if that population parameter were true (NHST). 

- We can determine the range within which the population mean is likely to be with a particular level of confidence (ie CI).



## Inferences from data

1. Null hypothesis significance testing
2. Precision estimation of a particular parameter
3. Comparing alternative models to see which fits the data better

## NHST

Logic:

- We are interested in estimating some statistic, for which we do not know the true population parameter. 

- We are going to assume some true value (null hypothesis), to see if our statistic is likely (or not) based on this assumption. 

- We then either reject the null (and accept the alternative), or fail to reject (and then are back at square one. The theory lives to fight another day)


## Problem of induction

- Ideally we want deductive reasoning. If the premises are true, the conclusion must be true. e.g., all people are mortal...

- Learning about the world necessitates induction. Start with specific observations and move to a general conclusion. e.g., all swans are white

- We use our sample to try and make general claims about the population, but we can never observe everything. Thus we can't prove our theories are true through induction alone.

## Falsification

- Popper states science does not progress by proving theories true (verification), but by proving theories false (falsification). 

- Research Hypothesis (H₁): This is your scientific theory 
- Null Hypothesis (H₀): A testable, and falsifiable straw man
- Use deductive logic, grounded in probability theory, to challenge the null. 
- Then make a tentative inductive inference about (H₁)


## NHST

1. Assume some population value you want to reject (H₀) 
2. Collect a sample 
3. Use sample statistics to construct sampling distribution
4. Compare the sample "test" statistic to a predefined "critical" value on the sampling distribution. 
5. If test statistic > critical value, reject null (H₀)
6. If test statistic < critical value, fail to reject null (H₁)


## Null and alternative hypothesis

- (H₀) is often a specific parameter value. Typically it is zero, which is often referred to as the nil hypothesis. 
- It can be a range, but a specific (worse case) value is needed to create the sampling distribution

- (H₁) is ALL other values. Even if you have a specific Estimate in mind

## NHST in action

```{r}
#| code-fold: true
library(ggplot2)



# --- Step 1: Create a "True" Population ---


set.seed(123) # For reproducibility
population_size <- 100000

# World where the Null Hypothesis is TRUE (No difference between groups)
population_h0_control <- rnorm(population_size, mean = 100, sd = 15)
population_h0_treatment <- rnorm(population_size, mean = 100, sd = 15) # Identical to control


# --- Step 2: Build the Sampling Distribution of the COEFFICIENT ---
# We'll simulate thousands of experiments, and for each one, we will store the
# estimated regression coefficient (the slope).

n_simulations <- 5000
sample_size <- 30
# Create an empty vector to store the SLOPE from each simulated experiment
slope_dist_h0 <- numeric(n_simulations)

for (i in 1:n_simulations) {
  # Draw samples from the world where the NULL IS TRUE
  sample_control <- sample(population_h0_control, sample_size)
  sample_treatment <- sample(population_h0_treatment, sample_size)
  
  # Combine data and create a grouping variable for the linear model
  combined_data <- data.frame(
    value = c(sample_control, sample_treatment),
    group = factor(rep(c("control", "treatment"), each = sample_size), levels = c("control", "treatment"))
  )
  
  # Run a linear model
  lm_result <- lm(value ~ group, data = combined_data)
  
  # Extract and store the SLOPE for the group predictor
  # The slope is in the 2nd row, 1st column of the coefficients table
  slope_dist_h0[i] <- summary(lm_result)$coefficients[2, 1]
}


# --- Step 3: Analyze the Sampling Distribution of the Slope ---
# The standard deviation of this distribution is our "true" Standard Error.

mean_of_slopes <- mean(slope_dist_h0)
simulated_standard_error <- sd(slope_dist_h0)


# --- Step 4: Connecting to a Single Regression Summary ---
# Now, let's run just ONE experiment and see how its summary output compares to our simulation.

our_sample_control <- sample(population_h0_control, sample_size)
our_sample_treatment <- sample(population_h0_treatment, sample_size)

our_data <- data.frame(
  value = c(our_sample_control, our_sample_treatment),
  group = factor(rep(c("control", "treatment"), each = sample_size), levels = c("control", "treatment"))
)

our_lm_test <- lm(value ~ group, data = our_data)
summary_our_lm <- summary(our_lm_test)

# Extract the Standard Error from the summary output
estimated_se_from_summary <- summary_our_lm$coefficients[2, 2]

# --- Step 5: Visualize the Sampling Distribution ---
# Let's plot the distribution of the slopes we generated.

plot_data <- data.frame(slopes = slope_dist_h0)

se_plot <- ggplot(plot_data, aes(x = slopes)) +
  geom_histogram(aes(y = ..density..), bins = 40, fill = "skyblue", color = "black") +
  geom_density(color = "red", size = 1) +
  geom_vline(xintercept = 0, color = "blue", size = 1.5, linetype = "dashed") +
  labs(
    title = "Sampling Distribution of the Regression Coefficient (Slope)",
    subtitle = "The distribution of estimates from 5000 experiments. The blue line is the true slope (0).",
    x = "Estimated Slope (Difference between group means)",
    y = "Density"
  ) +
  annotate("text", x = 0.5, y = 0.05, label = "True Slope = 0", color = "blue", fontface = "bold", hjust = 0) +
  theme_minimal()

print(se_plot)

```



## Sampling distribution regression estimate

- Centered around the null. 

- "The standard deviation of the sampling distribution is the standard error of the statistic"

```{r}
round(simulated_standard_error, 4)
```

## Our sample statistics

```{r}
summary(our_lm_test)
```

## NHST in action

- I want to see if our statistic (-1.529) is unlikley or likely assuming the null. 


```{r}
#| code-fold: true
ggplot(plot_data, aes(x = slopes)) +
  geom_histogram(aes(y = ..density..), bins = 40, fill = "skyblue", color = "black") +
  geom_density(color = "red", size = 1) +
  geom_vline(xintercept = 0, color = "blue", size = 1.5, linetype = "dashed") +
  geom_segment(aes(x = -1.529, xend = -1.529, y = 0, yend = .1), size = 1.5, linetype = "dashed") + 
  labs(
    title = "Sampling Distribution of the Regression Coefficient (Slope)",
    subtitle = "The distribution of estimates from 5000 experiments. The blue line is the null slope (0).",
    x = "Estimated Slope (Difference between group means)",
    y = "Density"
  ) +
  annotate("text", x = -8, y = 0.09, label = "Observed slope", color = "black", fontface = "bold", hjust = 0) +
  theme_minimal()
```

## NHST in action

- Next we need to define critical values. 

- Can be anywhere. Convention is to define extreme values as happening less than .05 percent of time. 

- Since extreme can be both above and below the mean, we set area under curve at 2.5% on each tail. 

```{r}
alpha <- 0.05
qnorm(1 - alpha/2,  0, 3.883)
qnorm(alpha/2,  0, 3.883)
```

---------------

```{r}
#| code-fold: true
# ---  Define the Rejection Region on the ESTIMATE's Scale ---
alpha <- 0.05
degrees_freedom <- 100
standard_error <- 3.883
# First, find the critical t-values
critical_t_upper <- qt(1 - alpha/2, df = degrees_freedom)
critical_t_lower <- qt(alpha/2, df = degrees_freedom)

# Now, convert these t-values back to the scale of your original estimate
critical_value_upper <- critical_t_upper * standard_error
critical_value_lower <- critical_t_lower * standard_error


 ggplot(plot_data, aes(x = slopes)) +
  geom_histogram(aes(y = ..density..), bins = 40, fill = "skyblue", color = "black") +
  geom_density(color = "red", size = 1) +
  geom_vline(xintercept = 0, color = "blue", size = 1.5, linetype = "dashed") +
  geom_segment(aes(x = -1.529, xend = -1.529, y = 0, yend = .1), size = 1.5, linetype = "dashed") + 
  labs(
    title = "Sampling Distribution of the Regression Coefficient (Slope)",
    subtitle = "The distribution of estimates from 5000 experiments. The blue line is the null slope (0).",
    x = "Estimated Slope (Difference between group means)",
    y = "Density"
  ) +
  annotate("text", x = -7.9, y = 0.09, label = "Observed slope", color = "black", fontface = "bold", hjust = 0) +
  geom_rect(aes(xmin = critical_value_upper, xmax = Inf, ymin = 0, ymax = Inf), fill = "grey", alpha = 0.1) +
  geom_rect(aes(xmin = -Inf, xmax = critical_value_lower, ymin = 0, ymax = Inf), fill = "grey", alpha = 0.1) +
  annotate("text", x = critical_value_upper + 2, y = 0.05, label = "Rejection\nRegion", color = "red", fontface = "bold") +
  annotate("text", x = critical_value_lower - 2, y = 0.05, label = "Rejection\nRegion", color = "red", fontface = "bold") 


```

## Decision

- Given that our statistic was not in the rejection region, we fail to reject the null. 

- Our value was consistent with a process that occurred assuming no effect. 

- This doesn't meant the null is true, just that the statistic from our sample was indistinguishable from a sampling process from a null distribution. 

## NHST in action

- Rather than work with our sampling distribution in original units, we are going to translate this distribution into a standardized form. We do this by dividing by the standard error (which is a standard deviation). 

- The information is the same, this is just convention. Because we have to estimate the SE, we use the t distribution

$$\frac{(\hat{\beta} - H_0)} {\hat{\sigma}}$$

## Critical values 


```{r}
#| code-fold: true
critical_t_upper <- qt(1 - alpha/2, df = 100)
critical_t_lower <- qt(alpha/2, df = 100)

# Now, convert these t-values back to the scale of your original estimate
critical_o.unit_upper <- critical_t_upper * standard_error
critical_o.unit_lower <- critical_t_lower * standard_error

critical_t_lower 
critical_o.unit_lower 
```

## Sampling distribution of the test statistic 

```{r}
#| code-fold: true
library(ggplot2)

set.seed(123) # For reproducibility
population_size <- 100000

# World 1: The Null Hypothesis is TRUE (No difference between groups)
population_h0_control <- rnorm(population_size, mean = 100, sd = 15)
population_h0_treatment <- rnorm(population_size, mean = 100, sd = 15) # Identical to control

# World 2: The Alternative Hypothesis is TRUE (A real difference exists)
population_h1_control <- rnorm(population_size, mean = 100, sd = 15)
population_h1_treatment <- rnorm(population_size, mean = 105, sd = 15) # Mean is 5 points higher


# --- Step 2: Build the Sampling Distribution ASSUMING THE NULL IS TRUE ---

n_simulations <- 5000
sample_size <- 30
# Create an empty vector to store the t-statistic from each simulated experiment
t_dist_h0 <- numeric(n_simulations)

for (i in 1:n_simulations) {
  # Draw samples from the world where the NULL IS TRUE
  sample_control <- sample(population_h0_control, sample_size)
  sample_treatment <- sample(population_h0_treatment, sample_size)
  
  # Combine data and create a grouping variable for the linear model
  combined_data <- data.frame(
    value = c(sample_control, sample_treatment),
    group = factor(rep(c("control", "treatment"), each = sample_size))
  )
  
  # Run a linear model 
  lm_result <- lm(value ~ group, data = combined_data)
  
  # Extract and store the t-statistic for the group predictor
  # The t-statistic is in the 2nd row, 3rd column of the coefficients table
  t_dist_h0[i] <- summary(lm_result)$coefficients[2, 3]
}


# --- Step 3: Define Our Rejection Region ---
# We set our standard of proof, alpha (α). This defines our "zone of surprise."
# If a result falls in this zone, it's too unlikely to have happened by chance.

alpha <- 0.05
# Find the critical t-values that mark the boundaries of our rejection region
degrees_freedom <- (sample_size - 1) * 2
critical_value_upper <- qt(1 - alpha/2, df = degrees_freedom)
critical_value_lower <- qt(alpha/2, df = degrees_freedom)


# --- Step 4: Conduct "Our" Single, Real Experiment ---
our_sample_control <- sample(population_h1_control, sample_size)
our_sample_treatment <- sample(population_h1_treatment, sample_size)

# Combine our single experiment's data for the linear model
our_data <- data.frame(
  value = c(our_sample_control, our_sample_treatment),
  group = factor(rep(c("control", "treatment"), each = sample_size))
)


# Extract the t-statistic and p-value for the group predictor
our_observed_t <- summary_our_lm$coefficients[2, 3]
p_value <- summary_our_lm$coefficients[2, 4]

# --- Step 5: Visualize the Verdict ---
# Let's plot the sampling distribution (from Step 2) and see where our
# single result (from Step 4) lands.

# Create a dataframe for the plot
plot_data <- data.frame(t_values = t_dist_h0)

ggplot(plot_data, aes(x = t_values)) +
  # Create the histogram of the sampling distribution under the null
  geom_histogram(aes(y = ..density..), bins = 40, fill = "skyblue", color = "black") +
  stat_function(fun = dt, args = list(df = degrees_freedom), color = "blue", size = 1) 

```


------------

- Don't need to work with the simulated sampling distribution, as we just need sample statistics to estimate.

$$\text{SE}(\hat{\beta}_1) = \frac{\text{RSE}}{\sqrt{\sum(X_i - \bar{X})^2}}$$

```{r}
#| code-fold: true
ggplot(data.frame(x = c(-4, 4)), aes(x = x)) +
  stat_function(fun = dt, args = list(df = 50)) 
```


## Our sample statistics

- Different from above, like we ran a new experiment

```{r}
summary(our_lm_test)
```


## Test statistic

```{r}
#| code-fold: true
ggplot(plot_data, aes(x = t_values)) +
    stat_function(fun = dt, args = list(df = degrees_freedom))  + geom_segment(aes(x = our_observed_t, xend = our_observed_t, y = 0, yend = .4), size = 1.5, linetype = "dashed") +

  labs(title = "The null distribution",
    x = "t-statistic",
    y = "Density"
  ) +  annotate("text", x = our_observed_t, y = -0.035, label = paste("Our observed test t-statistic\n(t =", round(our_observed_t, 2), ")"), color = "black", fontface = "bold") +
  theme_minimal()


```

## Critical values and Rejection regions

- Convention is two look at more extreme regardless of direction, a two tailed test. 

- Convention is to set it at .05 AUC. This is your *alpha* level. 

- A "one tail" test does the same, such that you specify your direction prior to analzying. Your critical value is closer to the null because now 5% AUC is on one side. 

---------------------

```{r}
#| code-fold: true
ggplot(plot_data, aes(x = t_values)) +
    stat_function(fun = dt, args = list(df = degrees_freedom))  + geom_segment(aes(x = our_observed_t, xend = our_observed_t, y = 0, yend = .4), size = 1.5, linetype = "dashed") +
  annotate("rect", xmin=-Inf, xmax=critical_value_lower, ymin=0, ymax=Inf, alpha=0.2, fill="red") + annotate("rect", xmin=critical_value_upper, xmax=Inf, ymin=0, ymax=Inf, alpha=0.2, fill="red") 
  

```
## NHST in action

- Again, our test statistic is not more extreme than our critical value so we fail to reject the null. 

- The logic of rejecting or not rejecting should be it, as in nothing else is needed given it is a decision rule. 

- Unfortunately, current practice awkwardly combines two differnt views on NHST: Fisherian and Neymann-Pearson. 

- This results in people wanting to also interpret p-values. 

## p-values

- Fisher only cared about p-values and the null. There were no alternative hypothesess or decisions to be made. Provided a strength against the null. 

P-value: 

$$p(\text{data} | H_0)$$

"The probability of getting the sample value or more extreme, assuming the null hypothesis is true"

-----------------

```{r}
#| code-fold: true
ggplot(data.frame(x = c(-4, 4)), aes(x = x)) +
  
annotate("rect", xmin=-Inf, xmax=our_observed_t, ymin=0, ymax=Inf, alpha=0.2, fill="red") + annotate("rect", xmin=abs(our_observed_t), xmax=Inf, ymin=0, ymax=Inf, alpha=0.2, fill="red") +
  
  # Draw the t-distribution curve
  stat_function(fun = dt, args = list(df = degrees_freedom), color = "blue", size = 1) +
  
  # Add a vertical line to mark our specific t-statistic
  geom_vline(xintercept = our_observed_t, linetype = "dashed", size = 1)+
  annotate("text", x = -3, y = 0.15, label = "This area under the curve...", color = "black") +
  annotate("text", x = 3, y = 0.15, label = "...plus this area...", color = "black") +
  annotate("text", x = 0.2, y = 0.12, label = paste("...equals the p-value\n(p =", round(p_value, 2), ")"), color = "black", fontface = "bold", size = 5) +
  
  # Add labels and title
  labs(
    title = "Conceptualizing a P-Value",
    x = "t-statistic",
    y = "Density"
  ) +
  theme_minimal()
```

---------


```{r}
pt(-0.394, df = 100) + (1 - pt(0.394, df = degrees_freedom)) 
```

- What would 1 sided p value be if 2 sided is .7? 

```{r}
pt(-0.394, df = 100) 
```

## Problems with p-values

- Doesn't tell you what you want to know, which leads to misunderstanding

"What's wrong with [NHST]? Well, among many other things, it does not tell us what we want to know, and we so much want to know what we want to know that, out of desperation, we nevertheless believe that it does!" – Cohen (1994)

- p(data | H0) when we really want p(H1 | data)

## Problems with p-values

- Focus on p-values leads to weird incentives

- p < .05 as a condition for publication, publication as a condition for job/tenure/grant$, which leads to gaming the system given it all hinges on this one value. Makes this all you care about at the expense of other things. 

- Leads to p-hacking, file drawers, etc (more later)

## Problems with p-values

- Leads to dichotomous thinking. 

- "The world is flat p < .05";  "Surely god loves .051 as much as .049"

- Fail vs not fail when the world is much more grey. Both in terms of original hypotheses (nil isn't risky) as well as interpreting results. 

- Lacks a measure of interpretable effect size and an estimate of precision 


## Problems with p-values

- Leads to uncritical thinking

- "I wouldn't trust a study with larger sample size than is standard (in the field), as that suggests they just gamed the system". - WashU big wig 

- Not reporting necessary information as long as p-value is there, ritualistic p-value corrections without thought, power estimates without thought, asterisk  


## Problems with p-values

- Assuming no effect, we are still wrong at best 1/20 times. Likely much more (low power, p-hacking, etc) in published lit. 

- Assuming an effect exists, do our results tell us anything? Are they risky hypotheses at all? Instead low value, low precision (“differences” are enough), almost no non-nil null hypotheses. 

## p-values and NHST

... are intrinsically tied. 

But there are other ways! 

## Estimation

Logic: 

- Sampling distribution provides us with information about how much we can trust the parameter estimate. 

- We can quantify this precision via a confidence interval (CI), created from the sampling distribution.

- Rather than focus on null/nil, focus on effect size estimate. 

- Can also do NHST, but with a focus on gauging uncertainty

## Confidence intervals


## Effect sizes

- practical versus statistical significance

- "One weird trick to significance. PhDs hate it."



## Model comparison

- We should explicitly test theories against one another rather than weak null/nil hypotheses.  

- Build more than one regression model and pit two (or more) models against one another

- Compare which fit the best. The focus can be on the overall model fit or on specific coefficients. Each depend on sigma, which is used estimate SE. 

## Revisit ANOVA table




