---
title: "Power"
format: revealjs
slide-number: true
editor: visual
execute:
  echo: true
html:
  code-fold: true
  code-summary: Show the code
  scrollable: true
---


## This time

- Power! Or, why we should do studies so they succeed (if they are supposed to succeed)

## Errors we make

|  | Fail to reject null | reject null | 
|------------|------------|------------|
| $H_0$ is True    | Correct decision   |    Type 1 error (false positive) | 
| $H_0$ is False      | Type II error (False Negative) |   Correct decision  |


## NHST and errors

- We directly control Type I errors via alpha level. 

- Controlling Type II errors is more challenging because it depends on several factors. Some argue that the null hypothesis is usually false, so the only error we can make is a Type II error -- a failure to detect a signal that is present.


## Power

- The complement of a Type II error is statistical power. 

- Power is the probability of correctly rejecting a false null hypothesis. It is the ability to detect an effect, if an effect is truly there.

α -- The rate at which we make Type I errors.  
β -- The rate at which we make Type II errors.  
1 − β -- statistical power.  

-----------

- Controlling Type II errors is the goal of power analysis and must contend with four quantities that are interrelated:

Sample size. 

Effect size. 

Significance level (α). 

Power. 

- When any three are known, the remaining one can be determined. 

- We must specify a specific value for the alternative hypothesis (e.g., our expected effect size) to estimate and control Type II errors.

## Example

Suppose we have a measure of social sensitivity that we have administered to a random sample of 20 psychology students. This measure has a population mean (μ
) of 10 and a standard deviation (σ) of 2. We suspect that psychology students are more sensitive to others than is typical and want to know if their mean, which is 11, is sufficient evidence to reject the null hypothesis that they are no more sensitive than the rest of the population.

We would also like to know how likely it is that we could make a mistake by concluding that psychology students are not different when they really are: A Type II error.

## Example

- Where would our critical value be? 

```{r}
10 + (qt(df = 20,.975) * (2/sqrt(20)))
```

```{r}
#| code-fold: true
library(ggplot2)
library(tidyverse)
library(ggdist)
library(distributional)

df <- tribble( ~df, ~mu, ~sigma, 20, 10, .44)

 ggplot(df, aes(xdist = distributional::dist_student_t(df, mu, sigma))) +
  stat_slab(aes())+
   annotate("rect", xmin=10.93287, xmax=Inf, ymin=0, ymax=Inf, alpha=0.2, fill="red")  
```





## Alternative distribution

- Just as when we centered our sampling distribution around the null merely for doing NHST.... 

- We can center a sampling distribution around our hypothesized (or observed) effect size. 

- They will have the same width. 


-------

```{r}
#| code-fold: true

distributions_df <- tribble(
  ~distribution,      ~df, ~mu, ~sigma,
  "Null (H0)",        20,  10,   0.44,
  "Alternative (H1)", 20,  11,   0.44
)

# --- Step 2: Define the Critical Value ---
critical_value <- 10.93287

ggplot(distributions_df, aes(
    xdist = distributional::dist_student_t(df = df, mu = mu, sigma = sigma)
  )) +  stat_dist_slab(
    alpha = 0.4,
    xlim = c(10.93287, "inf")
  ) +

  stat_dist_slab(
    data = . %>% filter(distribution == "Null (H0)"),
    alpha = 0.6) +

  geom_vline(xintercept = critical_value, linetype = "dashed", color = "darkred", size = 1) +
  
  labs(
    title = "Visualizing Statistical Power",
    subtitle = "Comparing the Null (H0) and Alternative (H1) Distributions",
    x = "Sample Mean",
    y = "Density"
  )

```


-----------


```{r}
#| code-fold: true
distributions_df <- tribble(
  ~distribution,      ~df, ~mu, ~sigma,
  "Null (H0)",        20,  10,   0.44,
  "Alternative (H1)", 20,  14,   0.44
)

# --- Step 2: Define the Critical Value ---
critical_value <- 10.93287

ggplot(distributions_df, aes(
    xdist = distributional::dist_student_t(df = df, mu = mu, sigma = sigma)
  )) +  stat_dist_slab(
    alpha = 0.5,
    xlim = c(10.93287, "inf"),
    normalize = "all" # Ensures curves are scaled relative to each other
  ) +

  stat_dist_slab(
    data = . %>% filter(distribution == "Null (H0)"),
    alpha = 0.7) +

  geom_vline(xintercept = critical_value, linetype = "dashed", color = "darkred", size = 1) +
  
    geom_vline(xintercept = 14, linetype = "dashed", color = "black", size = 1) +
  
  labs(
    title = "Visualizing Statistical Power",
    subtitle = "Comparing the Null (H0) and Alternative (H1) Distributions",
    x = "Sample Mean",
    y = "Density"
  )

```



-----------


```{r}
#| code-fold: true

distributions_df <- tribble(
  ~distribution,      ~df, ~mu, ~sigma,
  "Null (H0)",        20,  10,   0.44,
  "Alternative (H1)", 20,  10.5,   0.44
)

# --- Step 2: Define the Critical Value ---
critical_value <- 10.93287

ggplot(distributions_df, aes(
    xdist = distributional::dist_student_t(df = df, mu = mu, sigma = sigma)
  )) +  stat_dist_slab(
    alpha = 0.5,
    xlim = c(10.93287, "inf"),
    normalize = "all" # Ensures curves are scaled relative to each other
  ) +

  stat_dist_slab(
    data = . %>% filter(distribution == "Null (H0)"),
    alpha = 0.7) +

  geom_vline(xintercept = critical_value, linetype = "dashed", color = "darkred", size = 1) +
  
    geom_vline(xintercept = 10.5, linetype = "dashed", color = "black", size = 1) +
  
  labs(
    title = "Visualizing Statistical Power",
    subtitle = "Comparing the Null (H0) and Alternative (H1) Distributions",
    x = "Sample Mean",
    y = "Density"
  )

```


## Beta

- The rate that we make type II errors. Area under curve to the left of our critical value. 


```{r}
#| code-fold: true

distributions_df <- tribble(
  ~distribution,      ~df, ~mu, ~sigma,
  "Null (H0)",        20,  10,   0.44,
  "Alternative (H1)", 20,  11,   0.44
)

# --- Step 2: Define the Critical Value ---
critical_value <- 10.93287

ggplot(distributions_df, aes(
    xdist = distributional::dist_student_t(df = df, mu = mu, sigma = sigma)
  )) +  stat_dist_slab(
    alpha = 0.4
  ) +

  stat_dist_slab(
    data = . %>% filter(distribution == "Alternative (H1)"),
    alpha = 0.6,
    fill = "green") +
  
  geom_vline(xintercept = critical_value, linetype = "dashed", color = "darkred", size = 1) +
  
  labs(
    title = "Beta (type 2 error) is AUC left of C.V.",
    subtitle = "Power is AUC right of C.V.",
    x = "Sample Mean",
    y = "Density"
  )

```

----------------------

- In the long run, if our samples have a mean of 11 (σ = 2 , N = 20), we will correctly reject the null with probability of .56 (power). We will incorrectly fail to reject the null with probability of .44 (β). 

```{r}
# c.v. 10.93287 from before
(10.93 - 11) / (2/sqrt(20)) # standardized CV from alternative

pt(df = 20, -0.1565248) # Beta / Type 2

1 - .4386 # power
```
- EVEN IF 11 IS TRUE, our experiment will be for naught almost 1/2 the time!!

## Increase sample size

- You should never collect just 20 subjects. What happens when we collect more? 
```{r}
#| code-fold: true

distributions_df <- tribble(
  ~distribution,      ~df, ~mu, ~sigma,
  "Null (H0)",        75,  10,   0.23,
  "Alternative (H1)", 75,  11,   0.23
)

# --- Step 2: Define the Critical Value ---
critical_value <- 10 + (qt(df = 75,.975) * (2/sqrt(75)))

ggplot(distributions_df, aes(
    xdist = distributional::dist_student_t(df = df, mu = mu, sigma = sigma)
  )) +  stat_dist_slab(
    alpha = 0.4
  ) +

  stat_dist_slab(
    data = . %>% filter(distribution == "Alternative (H1)"),
    alpha = 0.6,
    fill = "green") +
  
  geom_vline(xintercept = critical_value, linetype = "dashed", color = "darkred", size = 1) +
  
  labs(
    title = "Beta (type 2 error) is AUC left of C.V.",
    subtitle = "Power is AUC right of C.V.",
    x = "Sample Mean",
    y = "Density"
  )

```


----------

```{r}
10 + (qt(df = 75,.975) * (2/sqrt(75))) # c.v. under null
(10.46 - 11) / (2/sqrt(75)) # standardized c.v. under alternative

pt(df = 75, -2.338269) # Beta / Type 2

1 - .01 # power
```

-----------

https://rpsychologist.com/d3/nhst/


## How can power be increased?

- Increase effect size (move alternative father away)

- increase your alpha (.056), which moves your critival value closer to the null

- increase N (reduces standard error --> more precision)

- Reduce observed standard deviation/sigma via better measurement or a better manipulation

## How do you choose an effect size?

- Past research can often provide some guidance, especially if a meta-analysis is available.

- Afield might have standards regarding what counts as a meaningful effect (e.g., minimal clinically important difference).

- Abstract benchmarks or rules of thumb about what are “small,” “medium”, and “large” effects.

## Packages

- There are many helper packages to do power analyses

{pwr} {simr} {pwrss} {InteractionPoweR} among others

- You have to be careful that for some packages they focus on either the omnibus model versus specific coefficients

- For some models there are no analytic solutions and one must simulate

## pwr

- Can only do very simple models. Rather than specifying your regression model, it asks you to pick a type of model. 

- Cannot do standard regression! Booo

```{r}
#| code-fold: true
library(pwr)
pwr::pwr.t.test(n = NULL,
                sig.level = 0.05, 
                type = "two.sample", 
                alternative = "two.sided", 
                power = 0.80, 
                d = .4)
```


## pwrss

- much better than pwr

```{r}
library(pwrss)
power.t.regression(beta = 0.40,
                   sd.outcome = 1.2,
                   sd.predictor = .8,
                   k.total = 1,
                   r.squared = 0.09,
                   power = .80,
                   alpha = 0.05,
                   alternative = "two.sided")
```

---------

```{r}
power.t.regression(beta = 0.30,
                   k.total = 1,
                   r.squared = 0.09,
                   power = .80,
                   alpha = 0.05,
                   alternative = "two.sided")
```

----------


```{r}
power.z.onecor(rho = 0.30,
               null.rho = 0.0,
               power = 0.80,
               alpha = 0.05,
               alternative = "one.sided")
```

----------

```{r}
# --- Define the Simulation Parameters ---

n_simulations <- 10000
sample_size <- 103
alpha <- 0.05
true_intercept <- 10
true_slope <- 0.4
true_error_sd <- 1.145 #(SD of Y * square root (1 - assumed R^2))

# --- Run the Simulation ---

p_values_slope <- numeric(n_simulations)

for (i in 1:n_simulations) {
  # --- For each simulation, create one sample dataset ---
  # 1. Generate the predictor variable 'x'
  x <- rnorm(sample_size, mean = 3, sd = .8) 
  
  # 2. Generate the outcome 'y' based on our TRUE model
  y <- true_intercept + true_slope * x + rnorm(sample_size, mean = 0, sd = true_error_sd)
  
  simulated_data <- data.frame(x, y)
  
  # --- Fit the linear model for this one experiment ---
  model <- lm(y ~ x, data = simulated_data)
  
  # --- Extract the p-value for our predictor 'x' ---
  # The summary() output gives us the t-test for each coefficient.
  # The slope for 'x' is the second coefficient in the model.
  # The p-value is in the 4th column of the coefficients table.
  p_values_slope[i] <- summary(model)$coefficients[2, 4]
}


# --- Step 4: Calculate and Report the Power ---
# Power is the proportion of our simulated experiments where we found a
# "significant" result (p < alpha).

power <- mean(p_values_slope < alpha)

# Print the final result
cat("--- Power Simulation Results ---\n")
cat("Number of simulations:", n_simulations, "\n")
cat("Sample size:", sample_size, "\n")
cat("Assumed True Slope (Effect Size):", true_slope, "\n")
cat("Alpha level:", alpha, "\n")
cat("\nEstimated Power to detect the effect:", power, "\n")
cat("Interpretation: With this sample size and effect size, we would expect to find a statistically significant relationship", power * 100, "% of the time.\n")


# --- Step 5: Visualize the Distribution of p-values ---
# This plot shows the distribution of p-values from our 2000 experiments.
# Because a real effect exists, the p-values will be skewed toward zero.

p_value_plot <- ggplot(data.frame(p_values = p_values_slope), aes(x = p_values)) +
  geom_histogram(bins = 40, fill = "skyblue", color = "black", boundary = 0) +
  geom_vline(xintercept = alpha, color = "red", linetype = "dashed", size = 1) +
  labs(
    title = "Distribution of p-values from Simulated Experiments",
    subtitle = paste("Power is the proportion of p-values to the left of the red line (α = 0.05)"),
    x = "p-value",
    y = "Frequency"
  ) +
  annotate("text", x = alpha, y = 200, label = "Significance\nThreshold", color = "red", hjust = -0.1) +
  theme_minimal(base_size = 14)

print(p_value_plot)
```

